---
title: "GTEx_PamGene"
output: html_document
date: "2025-01-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##load required packages

```{r}
library(anndata)
library(sceasy)
library(Seurat)
library(BPCells)
library(dplyr)
library(patchwork)
library(cowplot)
library(RColorBrewer)
library(ggplot2)
library(gridExtra)
library(grid)
library(cowplot)
library(ggplot2)
library(patchwork)
library(ggtext)
library(magrittr)
library(stringr)
library(ggVennDiagram)
library(readxl)
library(gplots)
#library(EnhancedVolcano)
library(ComplexUpset)
library(tidyr)
```

## Load GTEx datasets

Use write_matrix when you first run the code, once the matrix dir is already created, do not use

```{r}
GTEX_file = "data/GTEX_single_nuclear_data/GTEx_8_tissues_snRNAseq_atlas_071421.public_obs.h5ad"
GTEX_metadata_path = "data/GTEX_single_nuclear_data/GTEx_8_metadata.csv"
GTEX_genes = "data/GTEX_single_nuclear_data/GTEx_8_genes.csv"
GTEX_metadata <- read.csv(GTEX_metadata_path)
GTEX_genes <- read.csv(GTEX_genes)
 

parse.data <- open_matrix_anndata_hdf5(GTEX_file)

#write_matrix_dir(mat = parse.data, dir = "data/GTEX_single_nuclear_data/GTEx_8_tissues_parsed")

GTEX_mat <- open_matrix_dir(dir = "data/GTEX_single_nuclear_data/GTEx_8_tissues_parsed")

```

## Load cultured fibroblast dataset

```{r}
fibroblast_sample1 <- read.csv("data/Cultered_fibroblast/GSM8157769_HFF-50-CTR1.csv")
fibroblast_sample2 <- read.csv("data/Cultered_fibroblast/GSM8157770_HFF-50-CTR2.csv")


combined_fibroblast <- fibroblast_sample1 %>%
  left_join(fibroblast_sample2, keep = FALSE) 


rownames(combined_fibroblast) = combined_fibroblast$X
combined_fibroblast %<>%
  dplyr::select(-c(X, X.1))
```

#Next we convert the fibroblast data to log(10k + 1), this is the same as GTEx

```{r}
library_sizes <- colSums(combined_fibroblast)
tp10k <- t(t(combined_fibroblast) / library_sizes) * 10000
tp10k_1 <- tp10k + 1

log10k_1 <- log10(tp10k_1)
```

## Create a seurat object, for both fibroblasts as GTEx

GTEx first

```{r}
GTEX_object <- CreateSeuratObject(GTEX_mat, meta.data = GTEX_metadata)
raw_counts <- GetAssayData(object = GTEX_object, layer = "counts")
matrix <- as.matrix(raw_counts)

cardiac_cells <- GTEX_object@meta.data[grepl("Heart", GTEX_object@meta.data$Tissue.Site.Detail, ignore.case = TRUE), ]
skin_uv_cells <- GTEX_object@meta.data[grepl("skin", GTEX_object@meta.data$Tissue.Site.Detail, ignore.case = TRUE), ] %>%
  filter(Broad.cell.type == "Fibroblast")

subset_df <- rbind(cardiac_cells, skin_uv_cells) %>%
  distinct(X) #remove duplicates

GTEX_Seurat_subset <- subset(GTEX_object, cells = rownames(subset_df))
GTEX_Seurat_subset@assays$RNA@layers[["counts"]] <- as(GTEX_Seurat_subset@assays$RNA@layers[["counts"]], "dgCMatrix")

GTEX_Seurat_subset@meta.data %<>%
  mutate(Broad.cell.type = if_else(
    str_starts(Tissue.Site.Detail, "Skin"),
    str_c(Broad.cell.type, "_Skin"),
    Broad.cell.type))
```

Clutured fibroblasts

```{r}
fibroblast_seurat <- CreateSeuratObject(log10k_1)
fibroblast_seurat <- NormalizeData(fibroblast_seurat, normalization.method = "LogNormalize", scale.factor = 10000)
fibroblast_seurat <- ScaleData(fibroblast_seurat)
```

## Merge Seuratobjects

```{r}
fibro_gtex_seurat <- merge(GTEX_Seurat_subset, fibroblast_seurat, merge.data = TRUE)

fibro_gtex_seurat@meta.data %<>%
  mutate(Broad.cell.type = replace_na(Broad.cell.type, "Fibroblasts_cultered"))

fibro_gtex_seurat <- JoinLayers(fibro_gtex_seurat, layers= "counts")
```

## Create expression heatmap

```{r}
finished_pam_table <- read.csv("data/Objective3_MultiOmics_overlaps_PPRE/Pam_chip/Pam_with_ensembl.csv")
Pam_kinase_data <- read.csv("data/Objective3_MultiOmics_overlaps_PPRE/Pam_chip/Pam_112_kinases_with_ensembl.csv")

Combined_pam_pept_kin <- finished_pam_table %>%
  left_join(Pam_kinase_data, by = c("hgnc_symbol" = "Kinase.Name")) %>%
  filter(!is.na(`Kinase.Uniprot.ID`)) #overlapping genes between peptides and kinases

bold_labels <- function(labels) {
  sapply(labels, function(label) {
    if (label %in% Combined_pam_pept_kin$hgnc_symbol) {
      paste0("**", label, "**") # Add bold markdown syntax
    } else {
      label
    }
  })
}

heatmap_GTEX_Fib_pam_peptides <- SCpubr::do_ExpressionHeatmap(
  fibro_gtex_seurat,
  slot = 'counts',
  features = finished_pam_table$hgnc_symbol,
  axis.text.x.angle = 90,
  group.by = c("Broad.cell.type"),
  legend.position = "none",
  groups.order = list( Broad.cell.type = sort(levels(factor(fibro_gtex_seurat@meta.data$Broad.cell.type)), decreasing = TRUE)),
  min.cutoff = 0,
  diverging.palette = "RdYlBu",
  use_viridis = FALSE)


heatmap_GTEX_Fib_pam_peptides <- heatmap_GTEX_Fib_pam_peptides + scale_fill_gradientn(
  colors = colorRampPalette(c("black", "lightblue", "darkblue"))(100),
  values = c(0, 0.001, 1),       # Adjust distribution
  limits = c(0, 3)
) + labs(title = "Expression of Pam-peptides in GTEx snRNAseq", y = "Cell type", x = "Pam gene petides") + scale_x_discrete(labels = bold_labels)  + theme(axis.title.x = element_text(size = 14, face = "bold", hjust = 0.5), axis.text.x = ggtext::element_markdown(size = 12), plot.title = element_text(size = 16, face = "bold", hjust = 0.5))


heatmap_GTEX_Fib_pam_kinase <- SCpubr::do_ExpressionHeatmap(
  fibro_gtex_seurat,
  slot = 'counts',
  features = Pam_kinase_data$`Kinase.Name`,
  axis.text.x.angle = 90,
  group.by = c("Broad.cell.type"),
  groups.order = list( Broad.cell.type = sort(levels(factor(fibro_gtex_seurat@meta.data$Broad.cell.type)), decreasing = TRUE)),
  min.cutoff = 0,
  diverging.palette = "RdYlBu",
  flip = FALSE,
  use_viridis = FALSE)


heatmap_GTEX_Fib_pam_kinase <- heatmap_GTEX_Fib_pam_kinase + scale_fill_gradientn(
  colors = colorRampPalette(c("black", "lightblue", "darkblue"))(100),
  values = c(0, 0.001, 1),       # Adjust distribution
  limits = c(0, 3)
) + labs(title = "Expression of in silico calculated Pam Kinases in Gtex snRNAseq n = 112", y = "Cell type", x = "Pam gene kinase")  + scale_x_discrete(labels = bold_labels) +theme(axis.title.x = element_text(size = 14, face = "bold", hjust = 0.5), axis.text.x = ggtext::element_markdown(size = 12), plot.title = element_text(size = 16, face = "bold", hjust = 0.5))

heatmap_GTEX_Fib_pam_kinase
```

```{r}
"Broad.cell.type" %in% colnames(fibro_gtex_seurat@meta.data)

# Check number of kinase genes in your table
length(Pam_kinase_data$Kinase.Name)

# Remove NA and duplicates
kinase_features <- Pam_kinase_data$Kinase.Name
kinase_features <- unique(na.omit(kinase_features))

# Check overlap with Seurat object
sum(kinase_features %in% rownames(fibro_gtex_seurat))
kinase_features[kinase_features %in% rownames(fibro_gtex_seurat)][1:20]
```




```{r}
library(pheatmap)
Idents(fibro_gtex_seurat) <- "Broad.cell.type"
fibro_gtex_matrix_average <- AverageExpression(fibro_gtex_seurat, features = finished_pam_table$hgnc_symbol,  slot = 'counts') 



# Define color gradient similar to the original
color_palette <- colorRampPalette(c("black", "lightblue", "darkblue"))(100)
# Generate heatmap with hierarchical clustering
Heatmap <- pheatmap(t(fibro_gtex_matrix_average$RNA), 
         color = color_palette, 
         breaks =  c(
  seq(0, 0.001, length.out = 50),  # Negative values to center
  seq(0.00101, 3, length.out = 51)   # Center to max
),
         cluster_rows = FALSE, 
         cluster_cols = TRUE, 
clustering_distance_cols = "manhattan" , 
clustering_method = "average",
cutree_cols = 4,
         legend = TRUE,
scale = "none",
         fontsize_row = 10,
         fontsize_col = 10,
         main = "Expression of Pam-peptides in GTEx snRNAseq")


install.packages("gplotify")
library(ggplotify)
```
```{r}

```

```{r}
Pam_peptide_final_file_path = "data/Objective3_MultiOmics_overlaps_PPRE/Pam_chip/gene_list_pamgene_complete.csv"

Pam_peptide_final_table <- read.csv(Pam_peptide_final_file_path) 

mart <- useEnsembl("ensembl", dataset="hsapiens_gene_ensembl", mirror = 'useast')
Ensembl_list <- getBM(mart = mart,
           attribute = c('hgnc_symbol', 'ensembl_gene_id'),
           filters = 'hgnc_symbol',
           values = Pam_peptide_final_table$Gene)

Pam_peptide_final_table %>%
  left_join(Ensembl_list, by = c("Gene"= "hgnc_symbol"), relationship = "many-to-many") -> finished_pam_table_full_list
```


```{r}
heatmap_GTEX_Fib_pam_final_peptides <- SCpubr::do_ExpressionHeatmap(
  fibro_gtex_seurat,
  slot = 'counts',
  features = finished_pam_table_full_list$Gene,
  axis.text.x.angle = 90,
  group.by = c("Broad.cell.type"),
  legend.position = "none",
  groups.order = list( Broad.cell.type = sort(levels(factor(fibro_gtex_seurat@meta.data$Broad.cell.type)), decreasing = TRUE)),
  min.cutoff = 0,
  diverging.palette = "RdYlBu",
  flip = FALSE,
  use_viridis = FALSE)


heatmap_GTEX_Fib_pam_final_peptides <- heatmap_GTEX_Fib_pam_final_peptides + scale_fill_gradientn(
  colors = colorRampPalette(c("black", "lightblue", "darkblue"))(100),
  values = c(0, 0.001, 1),       # Adjust distribution
  limits = c(0, 3)
) + labs(title = "Expression of Pam Peptides in GTEx snRNAseq n = 176", y = "Cell type", x = "Pam gene petides") + scale_x_discrete(labels = bold_labels)  + theme(axis.title.x = element_text(size = 14, face = "bold", hjust = 0.5), axis.text.x = ggtext::element_markdown(size = 12), plot.title = element_text(size = 16, face = "bold", hjust = 0.5))

heatmap_GTEX_Fib_pam_final_peptides
```
```{r}
Pam_peptide_2026_pth <- "data/PamGene_peptides_260226.xlsx"

Pam_peptide_2026 <- read_excel(Pam_peptide_2026_pth)
```
```{r}
heatmap_GTEX_Fib_pam_final_peptides_2026 <- SCpubr::do_ExpressionHeatmap(
  fibro_gtex_seurat,
  slot = 'counts',
  features = Pam_peptide_2026$`HGNC symbol`,
  axis.text.x.angle = 90,
  group.by = c("Broad.cell.type"),
  legend.position = "none",
  groups.order = list( Broad.cell.type = sort(levels(factor(fibro_gtex_seurat@meta.data$Broad.cell.type)), decreasing = TRUE)),
  min.cutoff = 0,
  diverging.palette = "RdYlBu",
  flip = FALSE,
  use_viridis = FALSE)


heatmap_GTEX_Fib_pam_final_peptides_2026 <- heatmap_GTEX_Fib_pam_final_peptides_2026 + scale_fill_gradientn(
  colors = colorRampPalette(c("black", "lightblue", "darkblue"))(100),
  values = c(0, 0.001, 1),       # Adjust distribution
  limits = c(0, 3)
) + labs(title = "Expression of Pam Peptides (2026) in GTEx snRNAseq n = 121", y = "Cell type", x = "Pam gene petides") + scale_x_discrete(labels = bold_labels)  + theme(axis.title.x = element_text(size = 14, face = "bold", hjust = 0.5), axis.text.x = ggtext::element_markdown(size = 12), plot.title = element_text(size = 16, face = "bold", hjust = 0.5))

heatmap_GTEX_Fib_pam_final_peptides_2026

#NA, H2BC3, H3C13, H3C15, H3C14
```

Save heatmap

```{r}
pdf("data/Objective3_MultiOmics_overlaps_PPRE/Pam_chip/Pam_heatmap_GTEX_Fib_combined_newscale1.pdf", height = 15, width = 30)
heatmap_arranged <- wrap_plots(patchworkGrob(heatmap_GTEX_Fib_pam_final_peptides) , patchworkGrob(heatmap_GTEX_Fib_pam_kinase), nrow=2)
plot(heatmap_arranged)
dev.off()

```
```{r}
pdf("data/Objective3_MultiOmics_overlaps_PPRE/Pam_chip/Pam_heatmap_2026_GTEX_Fib_combined.pdf", height = 15, width = 30)
heatmap_arranged <- wrap_plots(patchworkGrob(heatmap_GTEX_Fib_pam_final_peptides_2026) , patchworkGrob(heatmap_GTEX_Fib_pam_kinase), nrow=2)
plot(heatmap_arranged)
dev.off()
```

```{r}
pdf("data/Objective3_MultiOmics_overlaps_PPRE/Pam_chip/Pam_heatmap_2026_GTEX_Fib_kinases.pdf", height = 15, width = 30)
heatmap_arranged <- wrap_plots(patchworkGrob(heatmap_GTEX_Fib_pam_final_peptides_2026) , patchworkGrob(heatmap_GTEX_Fib_pam_final_peptides), nrow=2)
plot(heatmap_arranged)
dev.off()
```

```{r}
png("data/Pam_chip/Pam_heatmap_GTEX_Fib_combined_newscale.png", height = 100, width = 200)
heatmap_arranged <- wrap_plots(heatmap_GTEX_Fib_pam_final_peptides , heatmap_GTEX_Fib_pam_kinase, nrow=2)
plot(heatmap_arranged)
dev.off()
dev.list()
print(heatmap_GTEX_Fib_pam_final_peptides)
class(heatmap_GTEX_Fib_pam_kinase)
ggsave("data/Pam_chip/Pam_heatmap_GTEX_Fib_combined_newscale1.pdf", 
       plot = heatmap_arranged, 
       height = 15, width = 30, device = "pdf")
```
## PCAplt
to control of the dataset of the cultured fibroblast arent real outliers
```{r}
fibro_gtex_seurat %<>%
  NormalizeData() %>%
  FindVariableFeatures(selection.method = "vst", nfeatures = 2000) %>%
  ScaleData()


fibro_gtex_seurat <- RunPCA(fibro_gtex_seurat, features = VariableFeatures(object = fibro_gtex_seurat), dims = 1:10)

fibro_gtex_seurat %<>%
  FindNeighbors( dims = 1:10) %>%
  FindClusters( resolution = 0.5) %>%
  RunUMAP( dims = 1:10)

cells_to_highlight <- WhichCells(fibro_gtex_seurat, expression = Broad.cell.type == "Fibroblasts_cultered")

DimPlot(fibro_gtex_seurat, reduction = "pca", group.by = "Broad.cell.type")
DimPlot(fibro_gtex_seurat, reduction = "pca", group.by = "Broad.cell.type", cells.highlight  = cells_to_highlight)
DimPlot(fibro_gtex_seurat, reduction = "umap", group.by = "Broad.cell.type")
```



