---
title: "PPAR/RXR Motif Alignment and MEME Export"
author: "Eli Brouwer"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
---

This file loads the HOCOMOCO and JASPAR PPAR/RXR motifs, aligns them alongside a reference PPRE motif, and exports them in MEME format for use as FIMO input.

```{r setup, include = FALSE}
library(tools)
library(DiffLogo)
library(TFBSTools)
library(motifStack)
library(Cairo)
library(universalmotif)
library(gridExtra)
library(grid)
```

```{r set-directory}
hocomoco_motif_directory = "data/PPAR_RXR_motifs_Hocomoco_Jaspar/HOCOMOCO"

jaspar_motif_directory = "data/PPAR_RXR_motifs_Hocomoco_Jaspar/JASPAR"
```

Load in both JASPAR and HOCOMOCO motifs and convert to PFM file structure.

```{r load-jaspar}
jaspar_motifs_lst <- list.files(jaspar_motif_directory)
for (i in 1:length(jaspar_motifs_lst)){
  pwm_matrix = getPwmFromPfmOrJasparFile(paste0(jaspar_motif_directory, "/", jaspar_motifs_lst[i]))
  write.table(pwm_matrix[,-1], file = paste0(jaspar_motif_directory, "/", file_path_sans_ext(jaspar_motifs_lst[i]), ".pwm"), sep = "\t", quote = FALSE, col.names = FALSE)
}
```

```{r jaspar-to-pfm}
# making a pfm list from jaspar
jaspar_motifs_pfm_lst <-  grep("\\.jaspar$", list.files(jaspar_motif_directory), value = TRUE)
jaspar_pfm_list <- list()
for (i in 1:length(jaspar_motifs_pfm_lst)){
  jaspar_pfm_list[[jaspar_motifs_pfm_lst[i]]] <- pcm2pfm(importMatrix(paste0(jaspar_motif_directory, "/", jaspar_motifs_pfm_lst[i]), to = "pcm"))[[1]]
}
```

```{r meme-to-pfm}
hocomoco_motifs_pfm_lst <- grep("\\.meme$", list.files(hocomoco_motif_directory), value = TRUE) # import all hocomoco

# Restrict to the 11 HOCOMOCO half-site motifs used in this alignment (matching
# `desired_order` below). RXRB.H13CORE.2.PS.A and the RXRG.H13CORE half-sites
# are intentionally excluded; the RXRG half-site is instead represented by the
# JASPAR motifs MA0856.1 and MA1556.1. Selecting by name (rather than by
# truncating to length(jaspar_motifs_pfm_lst), as before) keeps this list
# stable regardless of what else gets added to either motif directory.
hocomoco_motifs_pfm_lst <- hocomoco_motifs_pfm_lst[hocomoco_motifs_pfm_lst %in% c(
  "PPARA.H13CORE.0.P.B_meme_format.meme",
  "PPARA.H13CORE.1.P.B_meme_format.meme",
  "PPARD.H13CORE.0.PSM.A_meme_format.meme",
  "PPARG.H13CORE.0.P.B_meme_format.meme",
  "PPARG.H13CORE.1.P.B_meme_format.meme",
  "RXRA.H13CORE.0.PS.A_meme_format.meme",
  "RXRA.H13CORE.1.S.C_meme_format.meme",
  "RXRA.H13CORE.2.SM.B_meme_format.meme",
  "RXRA.H13CORE.3.P.B_meme_format.meme",
  "RXRB.H13CORE.0.P.C_meme_format.meme",
  "RXRB.H13CORE.1.SM.B_meme_format.meme"
)]

hocomoco_pfm_list <- list()
for (i in 1:length(hocomoco_motifs_pfm_lst)){
  hocomoco_pfm_list[[hocomoco_motifs_pfm_lst[i]]] <- importMatrix(paste0(hocomoco_motif_directory, "/", hocomoco_motifs_pfm_lst[i]))[[1]]
}
```

# Reference PPRE Motif

A canonical PPRE (DR1) motif is defined manually as a reference for alignment and comparison against the HOCOMOCO and JASPAR motifs.

```{r ppre-reference-motif}
PPRE_mat <- matrix(c(
  1,1,1,0,0,1,0,0,0,0,1,1,1,0,0,0,0,1,  # A
  0,0,0,0,1,0,0,0,0,1,0,0,0,0,0,0,1,0,  # C
  0,0,0,0,0,0,1,1,0,0,0,0,0,1,1,0,0,0,  # G
  0,0,0,1,0,0,0,0,1,0,0,0,0,0,0,1,0,0   # T
), nrow=4, byrow=TRUE)

rownames(PPRE_mat) <- c("A", "C", "G", "T")

ppre_motif <- new("pfm", mat=PPRE_mat, name="PPRE motif")
plotMotifLogoStack(list(ppre_motif), fontsize = 2, yaxis = FALSE, ylab = "", font = 'Arial')
```

```{r convert-pfm-to-ppm}
pfm_list <- c(hocomoco_pfm_list, jaspar_pfm_list)
pfm_list <- c(pfm_list, ppre_motif)
meme_list <- list()
annotation_file <- list()
for (i in 1:length(pfm_list)){
  meme_list[[i]] <- convert_type(motifs = pfm_list[[i]], type = 'PPM')
  annotation_file[[i]] <- meme_list[[i]]$name
}
```

# Motif Alignment

All motifs are aligned together and reordered into the desired display order (grouped by transcription factor, HOCOMOCO half-sites followed by their JASPAR counterparts).

```{r motif-alignment}
pdf("data/PPAR_RXR_motifs_Hocomoco_Jaspar/Objective1_results/aligned_motifs_threshold0.5.pdf", height = 15)
aligned_motifs <- DNAmotifAlignment(pfm_list, threshold = 0.5, minimalConsensus = 6)
motif_names <- sapply(aligned_motifs, function(m) m@name)
desired_order <- c("PPRE motif",
                   "MA1148.1\tPPARA::RXRA",
                   "MA1148.2\tPPARA::RXRA",
                   "MA0065.1\tPPARG::RXRA",
                   "PPARA.H13CORE.0.P.B",
                   "PPARA.H13CORE.1.P.B",
                   "PPARD.H13CORE.0.PSM.A",
                   "MA1550.1\tPPARD",
                   "MA1550.2\tPPARD",
                   "PPARG.H13CORE.0.P.B",
                   "PPARG.H13CORE.1.P.B",
                   "MA0066.1\tPPARG",
                   "MA0066.2\tPPARG",
                   "RXRA.H13CORE.0.PS.A",
                   "RXRA.H13CORE.1.S.C",
                   "RXRA.H13CORE.2.SM.B",
                   "RXRA.H13CORE.3.P.B",
                   "RXRB.H13CORE.0.P.C",
                   "RXRB.H13CORE.1.SM.B",
                   "MA0855.1\tRXRB",
                   "MA1555.1\tRXRB(RC)",
                   "MA0856.1\tRXRG",
                   "MA1556.1\tRXRG")
indices <- match(desired_order, motif_names)
aligned_motifs_order <- rev(aligned_motifs[indices])
plotMotifLogoStack(aligned_motifs_order, fontsize = 2, yaxis = FALSE, ylab = "", font = 'Arial')
dev.off()
```

```{r plot-alignment, fig.width=20, fig.height=30}
plotMotifLogoStack(aligned_motifs_order, fontsize = 4, yaxis = FALSE, ylab = "", font = 'Arial')
```

# PPRE Architecture Validation Figure

This figure checks whether the individual PPAR and RXR half-site motifs, plus the three full PPAR::RXR heterodimer motifs, recover the expected PPRE (DR1) architecture: a `AGGTCA`-like PPAR half-site, a 1 bp spacer, and a second `AGGTCA`-like RXR half-site. The synthetic `PPRE motif` reference itself is kept separate as the template.

Two consensus logos are built from the same 22 motifs and compared:

* **Manual alignment** — each motif is placed on a shared canvas at a hand-curated start column (with reverse-complementing where needed), based on where its `AGGTCA`-like core is known to sit.
* **Unsupervised alignment** — the same 22 motifs, but taken from the existing algorithmic `DNAmotifAlignment()` result (`aligned_motifs`) computed above, with no manual placement.

```{r ppre-manual-alignment-table}
# Hand-curated start column (on a shared canvas) and orientation for every
# individual motif, based on where its own AGGTCA-like core is known to sit.
# MA1555.1's source file already stores it reverse-complemented relative to
# the other RXR motifs, so it is flipped back here to align consistently.
manual_alignment <- data.frame(
  Motif_name = c(
    "MA1148.1\tPPARA::RXRA", "MA1148.2\tPPARA::RXRA", "MA0065.1\tPPARG::RXRA",

    "PPARA.H13CORE.0.P.B", "PPARA.H13CORE.1.P.B", "PPARD.H13CORE.0.PSM.A",
    "MA1550.1\tPPARD", "MA1550.2\tPPARD", "PPARG.H13CORE.0.P.B", "PPARG.H13CORE.1.P.B",
    "MA0066.1\tPPARG", "MA0066.2\tPPARG",

    "RXRA.H13CORE.0.PS.A", "RXRA.H13CORE.1.S.C", "RXRA.H13CORE.2.SM.B", "RXRA.H13CORE.3.P.B",
    "RXRB.H13CORE.0.P.C", "RXRB.H13CORE.1.SM.B",
    "MA0855.1\tRXRB", "MA1555.1\tRXRB", "MA0856.1\tRXRG", "MA1556.1\tRXRG"
  ),
  Start = c(
    2, 2, 2,
    2, 3, 4, 4, 5, 2, 1, 4, 5,
    10, 10, 5, 1, 6, 4, 5, 12, 5, 12
  ),
  Reverse_complement = c(
    FALSE, FALSE, FALSE,
    FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, FALSE,
    FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, TRUE, FALSE, FALSE
  ),
  # Which guide box(es) this motif's core should land under: the three full
  # heterodimers span both half-sites, everything else spans just one
  Family = c(
    "BOTH", "BOTH", "BOTH",
    "PPAR", "PPAR", "PPAR", "PPAR", "PPAR", "PPAR", "PPAR", "PPAR", "PPAR",
    "RXR", "RXR", "RXR", "RXR", "RXR", "RXR", "RXR", "RXR", "RXR", "RXR"
  ),
  stringsAsFactors = FALSE
)

normalize_matrix <- function(mat) {
  mat <- mat[c("A", "C", "G", "T"), , drop = FALSE]
  mat[is.na(mat)] <- 0
  totals <- colSums(mat)
  informative <- totals > 0
  if (any(informative)) {
    mat[, informative] <- sweep(mat[, informative, drop = FALSE], MARGIN = 2, STATS = totals[informative], FUN = "/")
  }
  mat
}

# Reverse-complement a PFM: reverse column order and swap A<->T, C<->G rows
reverse_complement_matrix <- function(mat) {
  mat <- mat[c("T", "G", "C", "A"), ncol(mat):1, drop = FALSE]
  rownames(mat) <- c("A", "C", "G", "T")
  mat
}

place_on_canvas <- function(mat, start, canvas_width) {
  end <- start + ncol(mat) - 1
  if (start < 1 || end > canvas_width) {
    stop("Motif does not fit on canvas. Start = ", start, ", end = ", end, ", canvas width = ", canvas_width)
  }
  canvas <- matrix(0, nrow = 4, ncol = canvas_width, dimnames = list(c("A", "C", "G", "T"), seq_len(canvas_width)))
  canvas[, start:end] <- mat
  canvas
}

motifs_by_name <- setNames(pfm_list, vapply(pfm_list, function(motif) motif@name, character(1)))
motifs_by_name <- motifs_by_name[names(motifs_by_name) != "PPRE motif"]

missing_names <- setdiff(manual_alignment$Motif_name, names(motifs_by_name))
if (length(missing_names) > 0) {
  stop("These motifs were not found in pfm_list: ", paste(missing_names, collapse = ", "))
}

canvas_width <- max(vapply(seq_len(nrow(manual_alignment)), function(i) {
  manual_alignment$Start[i] + ncol(motifs_by_name[[manual_alignment$Motif_name[i]]]@mat) - 1
}, numeric(1)))

manual_aligned_matrices <- lapply(seq_len(nrow(manual_alignment)), function(i) {
  motif_name <- manual_alignment$Motif_name[i]
  mat <- normalize_matrix(motifs_by_name[[motif_name]]@mat)
  if (manual_alignment$Reverse_complement[i]) mat <- reverse_complement_matrix(mat)
  place_on_canvas(mat = mat, start = manual_alignment$Start[i], canvas_width = canvas_width)
})
names(manual_aligned_matrices) <- manual_alignment$Motif_name
```

```{r ppre-manual-consensus}
# Coverage-aware consensus: average only the motifs that actually cover each
# canvas position, then keep only positions covered by at least half of the
# 22 motifs. Without this trim, the few columns only 1-2 motifs happen to
# reach (the wide flanks of the longer heterodimer motifs) would dilute the
# consensus with near-background noise instead of being dropped.
n_motifs <- length(manual_aligned_matrices)
consensus_matrix <- matrix(0, nrow = 4, ncol = canvas_width, dimnames = list(c("A","C","G","T"), seq_len(canvas_width)))
position_coverage <- numeric(canvas_width)
for (position in seq_len(canvas_width)) {
  contributing <- vapply(manual_aligned_matrices, function(mat) sum(mat[, position]) > 0, logical(1))
  position_coverage[position] <- sum(contributing)
  if (any(contributing)) {
    position_values <- vapply(manual_aligned_matrices[contributing], function(mat) mat[, position], numeric(4))
    consensus_matrix[, position] <- rowMeans(position_values)
  }
}

minimum_coverage <- ceiling(n_motifs / 2)
positions_to_keep <- position_coverage >= minimum_coverage

manual_consensus <- consensus_matrix[, positions_to_keep, drop = FALSE]
manual_consensus <- sweep(manual_consensus, 2, colSums(manual_consensus), "/")
colnames(manual_consensus) <- seq_len(ncol(manual_consensus))

manual_pfm <- new("pfm", mat = manual_consensus, name = "Consensus from manual alignment")
```

```{r ppre-manual-stack, fig.width=14, fig.height=22}
# Every individually-placed motif plotted on its own, with the PPRE template
# on top for reference
stack_pfms <- c(
  list(`PPRE motif` = new("pfm", mat = normalize_matrix(PPRE_mat), name = "PPRE motif")),
  lapply(seq_len(nrow(manual_alignment)), function(i) {
    motif_name <- manual_alignment$Motif_name[i]
    mat <- normalize_matrix(motifs_by_name[[motif_name]]@mat)
    if (manual_alignment$Reverse_complement[i]) mat <- reverse_complement_matrix(mat)
    new("pfm", mat = mat, name = motif_name)
  })
)
plotMotifLogoStack(rev(stack_pfms), fontsize = 3, yaxis = FALSE, ylab = "", font = "Arial")
```

```{r ppre-unsupervised-consensus}
# Same 22 motifs, but pulled from the existing DNAmotifAlignment() result
# instead of being manually placed. DNAmotifAlignment() appends "(RC)" to the
# @name of motifs it decides to flip during alignment, so match on the name
# with any "(RC)" suffix stripped.
strip_rc <- function(x) sub("\\(RC\\)$", "", x)
aligned_names <- sapply(aligned_motifs, function(m) m@name)
unsupervised_idx <- match(strip_rc(manual_alignment$Motif_name), strip_rc(aligned_names))
stopifnot(!any(is.na(unsupervised_idx)))

unsupervised_mats <- lapply(aligned_motifs[unsupervised_idx], function(m) m@mat)
common_width <- unique(vapply(unsupervised_mats, ncol, integer(1)))
stopifnot(length(common_width) == 1)  # DNAmotifAlignment() pads all motifs to equal width

unsupervised_stack <- array(NA_real_, dim = c(4, common_width, length(unsupervised_mats)))
for (i in seq_along(unsupervised_mats)) unsupervised_stack[,,i] <- unsupervised_mats[[i]]

unsupervised_consensus <- apply(unsupervised_stack, c(1,2), mean, na.rm = TRUE)
unsupervised_consensus[is.nan(unsupervised_consensus)] <- 0.25
unsupervised_consensus <- sweep(unsupervised_consensus, 2, colSums(unsupervised_consensus), "/")
rownames(unsupervised_consensus) <- c("A","C","G","T")

unsupervised_pfm <- new("pfm", mat = unsupervised_consensus, name = "Consensus from unsupervised alignment")
```

```{r ppre-reference-diagram}
# Rendered with motifStack's own plotMotifLogo (same Cairo engine, same
# "Information content" y-axis as panels 2/3 below), not a hand-drawn
# reconstruction. PPRE_mat is deterministic (one base per column at
# probability 1), so every position reaches the maximum 2 bits and the
# letters fill the full plot height.
ppre_pfm <- new("pfm", mat = normalize_matrix(PPRE_mat), name = "PPRE motif")

# margins match panels 2/3's y-axis gutter (left = 4.1 lines); extra top
# margin leaves room for the "PPAR motif" / "RXR motif" labels above the
# boxes, below the "PPRE motif" title
ppre_reference_grob <- grid::grid.grabExpr({
  plotMotifLogo(pfm = ppre_pfm, motifName = "PPRE motif", p = ppre_pfm@background,
                colset = ppre_pfm@color, yaxis = TRUE, ylab = "Information content", xlab = NA,
                margins = c(1.5, 4.1, 6.5, 1.0), newpage = FALSE)

  # data region occupies roughly npc [0.10, 0.98] once the y-axis gutter is
  # accounted for; calibrated against the rendered axis tick positions
  left_frac <- 0.10
  right_frac <- 0.98
  col_width <- (right_frac - left_frac) / 18
  ppar_box <- left_frac + c(5, 11) * col_width
  rxr_box  <- left_frac + c(12, 18) * col_width

  box_bottom <- 0
  box_top <- 0.88
  for (box in list(ppar_box, rxr_box)) {
    grid.rect(x = unit(mean(box), "npc"), y = unit((box_bottom + box_top) / 2, "npc"),
              width = unit(diff(box), "npc"), height = unit(box_top - box_bottom, "npc"),
              gp = gpar(col = "red", fill = NA, lwd = 2.5))
  }
  grid.text("PPAR motif", x = unit(mean(ppar_box), "npc"), y = unit(box_top + 0.03, "npc"),
            just = c("center", "bottom"), gp = gpar(fontface = "bold", fontsize = 13))
  grid.text("RXR motif", x = unit(mean(rxr_box), "npc"), y = unit(box_top + 0.03, "npc"),
            just = c("center", "bottom"), gp = gpar(fontface = "bold", fontsize = 13))
})

grid.newpage()
grid.draw(ppre_reference_grob)
```

```{r ppre-comparison-figure, fig.width=10, fig.height=12}
panel2_grob <- grid::grid.grabExpr(
  plotMotifLogo(manual_pfm, motifName = "Consensus from manual alignment", ylab = "Information content", margins = c(4.1, 4.1, 2.1, 0.1))
)
panel3_grob <- grid::grid.grabExpr(
  plotMotifLogo(unsupervised_pfm, motifName = "Consensus from unsupervised alignment", ylab = "Information content", margins = c(4.1, 4.1, 2.1, 0.1))
)

pdf("data/PPAR_RXR_motifs_Hocomoco_Jaspar/Objective1_results/PPRE_architecture_validation.pdf", width = 10, height = 12)
grid.arrange(ppre_reference_grob, panel2_grob, panel3_grob, ncol = 1, heights = c(1.3, 1, 1))
dev.off()

grid.arrange(ppre_reference_grob, panel2_grob, panel3_grob, ncol = 1, heights = c(1.3, 1, 1))
```

# Convert to MEME Format for FIMO Input

The aligned motifs (excluding the synthetic PPRE reference) are exported as MEME files in batches of four, matching the FIMO run's input file groupings.

```{r pfm-to-meme-fimo}
# Motif order going into the MEME export, for reference against the batch files below
motif_names

write_meme(meme_list[1:4], "data/PPAR_RXR_motifs_Hocomoco_Jaspar/RXR_PPAR_combined_files_1_4.meme" ,overwrite = TRUE)
write_meme(meme_list[5:8], "data/PPAR_RXR_motifs_Hocomoco_Jaspar/RXR_PPAR_combined_files_5_8.meme" ,overwrite = TRUE)
write_meme(meme_list[9:12], "data/PPAR_RXR_motifs_Hocomoco_Jaspar/RXR_PPAR_combined_files_9_12.meme" ,overwrite = TRUE)
write_meme(meme_list[13:16], "data/PPAR_RXR_motifs_Hocomoco_Jaspar/RXR_PPAR_combined_files_13_16.meme" ,overwrite = TRUE )
write_meme(meme_list[17:20], "data/PPAR_RXR_motifs_Hocomoco_Jaspar/RXR_PPAR_combined_files_17_20.meme" , overwrite = TRUE)
write_meme(meme_list[21:22], "data/PPAR_RXR_motifs_Hocomoco_Jaspar/RXR_PPAR_combined_files_21_22.meme" , overwrite = TRUE)
```
