---
title: "UCSC Custom Tracks, FAM53A Multi-Track Figure, and TSS-Centered Motif Density"
author: "Eli Brouwer"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
---

This extends the FIMO/ChIPseeker pipeline, run on the **unmasked** genome
(`Objective1_chipseeker_fimo_analysis_no_max.Rmd`; see `Masking_decision_methods_note.md` for
why repeat-masking was checked and not adopted -- it doesn't meaningfully change the final gene
count) with three deliverables:

1. **Per-motif BED files** of every genome-wide FIMO hit, formatted as UCSC custom
   tracks (one file per motif, ready to upload at
   [genome.ucsc.edu/cgi-bin/hgCustom](https://genome.ucsc.edu/cgi-bin/hgCustom)).
2. **A synthetic multi-track figure** for the FAM53A promoter (the representative
   promoter where essentially all 30 motifs hit), showing every motif as its own row
   -- the same idea as a HOCOMOCO/MoLoTool-style multi-motif screenshot.
3. **A TSS-centered binding density plot and heatmap**, using the `distanceToTSS`
   ChIPseeker already computed for every promoter hit, to see whether motifs cluster
   right at the TSS or spread out toward the 5' UTR / further upstream.

**On "32 motifs":** the alignment script (`Objective1_motif_alignment_CISBP_additions.Rmd`)
exports 32 MEME motifs in total -- the 30 individual motifs (22 original + 8 CIS-BP), plus 2
synthetic motifs (`PPRE_canonical_motif.meme`, `PPRE_consensus_all30_motif.meme`). All 32 were
run against the same **unmasked** genome, order-0 background, uncapped `--max-stored-scores` --
a clean apples-to-apples comparison, unlike an earlier version of this file where the 2
synthetic motifs had accidentally been scanned on a different genome/background than the other
30. Their promoter-annotated hits (`PPRE_canonical.txt`, `PPRE_consensus_all30.txt`) were copied
alongside the other 30's ChIPseeker output so all three sections below cover all 32 motifs
together.

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(pacman)
p_load(data.table, GenomicRanges, GenomeInfoDb, Gviz, ComplexHeatmap, circlize, ggplot2, tools,
       ChIPseeker, TxDb.Hsapiens.UCSC.hg38.knownGene, org.Hs.eg.db)
```

```{r set-directory}
# Unmasked genome, uncapped max-stored-scores -- the pipeline's final setting (see
# Masking_decision_methods_note.md). All 32 motifs (30 individual + 2 synthetic) were run
# against this same genome/background, so this single directory covers everything below.
project_dir <- "data/PPAR_RXR_motifs_Hocomoco_Jaspar"
no_max_fimo_dir <- file.path(project_dir, "FIMO_results_no_max")
fimo_batches <- c("1_4", "5_8", "9_12", "13_16", "17_20", "21_22", "23_26", "27_30")
fimo_file_paths <- file.path(no_max_fimo_dir, fimo_batches, "fimo.tsv")

# The 2 synthetic motifs (canonical DR1, all-30 consensus), same genome/background as the
# other 30. Picked up automatically if their fimo.tsv exists at these paths.
extra_batches <- c("PPRE_canonical", "PPRE_consensus_all30")
extra_fimo_paths <- file.path(no_max_fimo_dir, extra_batches, "fimo.tsv")
extra_available <- file.exists(extra_fimo_paths)
if (any(extra_available)) {
  fimo_file_paths <- c(fimo_file_paths, extra_fimo_paths[extra_available])
}
cat(length(fimo_file_paths), "FIMO batch file(s) found (", sum(extra_available), "/ 2 extra motifs present ).\n")

# Existing, already-annotated per-motif promoter output from the unmasked/no_max run -- used
# by sections 2 and 3 below (fast: promoter-filtered, ~1-9 MB per motif, vs. re-parsing the
# full ~1.2 GB of genome-wide fimo.tsv). The 2 synthetic motifs' promoter output was copied in
# alongside the other 30's (see intro note above), so all 32 are present here.
results_root <- file.path(no_max_fimo_dir, "Objective1_analysis")
chip_file_directory <- file.path(results_root, "ChIPseeker_output")
results_dir <- file.path(results_root, "Objective1_results")

bed_out_dir <- file.path(results_dir, "Bedfiles_output", "per_motif_UCSC_tracks")
fam53a_out_dir <- file.path(results_dir, "FAM53A_multitrack")
tss_out_dir <- file.path(results_dir, "TSS_metaplot")
dir.create(bed_out_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(fam53a_out_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(tss_out_dir, showWarnings = FALSE, recursive = TRUE)
```

# Per-Motif UCSC Custom Track BED Files

**Promoter-restricted, not genome-wide.** Each motif's BED file is built from its
already-annotated `chip_file_directory/<motif>.txt` (ChIPseeker's `annotatePeak(tssRegion =
c(-3000, 1000))` "Promoter" hits -- the same file every other section in this document reads),
rather than the raw genome-wide `fimo.tsv`. That guarantees the BED tracks use the *exact*
same -3000/+1000 TSS window as the FAM53A/CPT1B/HADHA/HADHB figures and the TSS-density
analysis, instead of shipping ~36.8M unfiltered genome-wide hits (the earlier version of this
section did that; it's a browsable resource but not a meaningful thing to load as a "PPRE
binding sites" track, and it wasn't using any promoter definition at all). Each motif's file
still gets its own `track` header line, ready for UCSC's
["Add Custom Tracks"](https://genome.ucsc.edu/cgi-bin/hgCustom) page (assembly: **hg38**).
Scores are clipped to BED's required 0-1000 range (`useScore=1` shades each feature by it).

```{r bedfile-per-motif}
export_motif_promoter_bed <- function(motif_id, out_dir) {
  path <- file.path(chip_file_directory, paste0(motif_id, ".txt"))
  df <- read.delim(path, sep = " ", header = TRUE, check.names = FALSE)
  dt <- as.data.table(df)

  bed <- data.frame(
    chrom      = dt$seqnames,
    chromStart = dt$start - 1L,
    chromEnd   = dt$end,
    name       = motif_id,
    score      = pmin(pmax(round(dt$score), 0), 1000),
    strand     = dt$strand
  )
  bed <- bed[order(bed$chrom, bed$chromStart), ]
  out_file <- file.path(out_dir, paste0(motif_id, ".bed"))
  track_line <- sprintf(
    'track name="%s" description="%s FIMO hits (hg38, promoter-restricted: -3000/+1000 of TSS)" visibility=2 useScore=1',
    motif_id, motif_id)
  writeLines(track_line, out_file)
  fwrite(bed, out_file, sep = "\t", col.names = FALSE, append = TRUE, quote = FALSE)
  nrow(bed)
}

bed_motif_ids <- file_path_sans_ext(basename(list.files(chip_file_directory, pattern = "\\.txt$")))
bed_hit_counts <- sapply(bed_motif_ids, export_motif_promoter_bed, out_dir = bed_out_dir)
cat("Wrote", length(bed_motif_ids), "per-motif BED files to", bed_out_dir,
    "(", sum(bed_hit_counts), "total promoter-restricted hits )\n")

# Concatenating all per-motif files (each already carries its own `track` line) into one
# file lets UCSC's custom-track uploader register every motif as a separate track from a
# single upload, instead of doing 30 (or 32) separate uploads.
combined_bed <- file.path(bed_out_dir, paste0("all_", length(bed_motif_ids), "_motifs_UCSC_multitrack.bed"))
file.create(combined_bed)
for (mid in bed_motif_ids) {
  file.append(combined_bed, file.path(bed_out_dir, paste0(mid, ".bed")))
}
cat("Combined multi-track file:", combined_bed, "\n")
```

# FAM53A Promoter Multi-Track Figure

FAM53A (chr4, minus strand, TSS at the annotated 3' end of `ENST00000461064.5`,
`geneEnd = 1,668,905`) is the representative promoter: with the current unmasked,
no-max-truncation FIMO run, most of the 30 motifs have at least
one hit inside its `[-3000, +1000]` promoter window (exact count printed below). One row per motif is drawn, colored by
TF family, over the same promoter window ChIPseeker used
(`annotatePeak(..., tssRegion = c(-3000, 1000))`), giving a HOCOMOCO/MoLoTool-style
stacked view of where each motif's hits fall relative to the TSS.

```{r fam53a-window}
# FAM53A has several overlapping transcripts in TxDb.Hsapiens.UCSC.hg38.knownGene (short
# isoforms, a readthrough transcript, etc.), each with its own geneStart/geneEnd/TSS, and
# ChIPseeker assigns each individual hit to whichever transcript it lands nearest -- so
# picking an arbitrary FAM53A row (e.g. the first one found) can silently pick a minor
# isoform's TSS instead of the gene's main one. Instead, load every motif's FAM53A rows and
# use the modal (most-hit-supported) transcript as the reference -- ENST00000461064.5,
# geneStart=1,640,495 / geneEnd=1,668,905, in ~56% of the 386 promoter hits below, vs. the
# next-most-common transcript at ~18%.
motif_files_probe <- list.files(chip_file_directory, pattern = "\\.txt$", full.names = TRUE)
fam53a_all_rows <- rbindlist(lapply(motif_files_probe, function(f) {
  df <- read.delim(f, sep = " ", header = TRUE, check.names = FALSE)
  dt <- as.data.table(df)
  dt[SYMBOL == "FAM53A", .(transcriptId, geneStart, geneEnd, geneStrand)]
}), use.names = TRUE)
transcript_counts <- fam53a_all_rows[, .N, by = .(transcriptId, geneStart, geneEnd, geneStrand)][order(-N)]
cat("FAM53A transcripts seen across all motifs' promoter hits:\n")
print(transcript_counts)
fam53a_gene_row <- transcript_counts[1]
gene_start <- fam53a_gene_row$geneStart
gene_end <- fam53a_gene_row$geneEnd
gene_strand <- if (fam53a_gene_row$geneStrand == 2) "-" else "+"

tss_region <- c(-3000, 1000)  # matches the annotatePeak() call in the no_max (unmasked) script
tss <- if (gene_strand == "-") gene_end else gene_start
window_start <- if (gene_strand == "-") tss - tss_region[2] else tss + tss_region[1]
window_end   <- if (gene_strand == "-") tss - tss_region[1] else tss + tss_region[2]
cat("FAM53A promoter window: chr4:", window_start, "-", window_end, " (TSS =", tss, ", strand", gene_strand, ")\n")
```

```{r fam53a-load-hits}
motif_files <- list.files(chip_file_directory, pattern = "\\.txt$", full.names = TRUE)
motif_ids <- file_path_sans_ext(basename(motif_files))

# read.delim() (not fread()) because these files were written by write.table() with the
# default row.names=TRUE, giving the header one fewer field than each data row -- fread()
# does not auto-handle that offset the way read.table()/read.delim() do.
read_promoter_file <- function(path, mid) {
  df <- read.delim(path, sep = " ", header = TRUE, check.names = FALSE)
  dt <- as.data.table(df)[, c("start", "end", "strand", "distanceToTSS", "SYMBOL")]
  dt[, motif_id := mid]
  dt
}
all_promoter_hits <- rbindlist(Map(read_promoter_file, motif_files, motif_ids), use.names = TRUE)
fam53a_hits <- all_promoter_hits[SYMBOL == "FAM53A"]
cat("FAM53A promoter hits:", nrow(fam53a_hits), "across", uniqueN(fam53a_hits$motif_id), "/", length(motif_ids), "motifs\n")
```

```{r fam53a-multitrack, fig.width=10, fig.height=16}
# TF-family lookup by exact motif_id, matching the `Family` groupings already used in
# Objective1_motif_alignment_CISBP_additions.Rmd. A regex like `^PPARG` works for the
# HOCOMOCO/CIS-BP motif_ids (e.g. "PPARG.H13CORE.0.P.B") but NOT for JASPAR's accession-style
# ids (e.g. "MA0066.1" is PPARG, "MA1550.1" is PPARD, "MA0855.1"/"MA1555.1" are RXRB,
# "MA0856.1"/"MA1556.1" are RXRG) -- those need to be listed explicitly, so all 30 are here.
family_lookup <- c(
  "MA1148.1" = "PPAR::RXR", "MA1148.2" = "PPAR::RXR", "MA0065.1" = "PPAR::RXR",
  "PPARA.H13CORE.0.P.B" = "PPARA", "PPARA.H13CORE.1.P.B" = "PPARA", "PPARA_M12541_3.10" = "PPARA",
  "PPARD.H13CORE.0.PSM.A" = "PPARD", "MA1550.1" = "PPARD", "MA1550.2" = "PPARD",
  "PPARD_M05794_3.10" = "PPARD", "PPARD_M05796_3.10" = "PPARD",
  "PPARG.H13CORE.0.P.B" = "PPARG", "PPARG.H13CORE.1.P.B" = "PPARG", "MA0066.1" = "PPARG",
  "MA0066.2" = "PPARG", "PPARG_M12496_3.10" = "PPARG", "PPARG_M12497_3.10" = "PPARG",
  "PPARG_M12498_3.10" = "PPARG",
  "RXRA.H13CORE.0.PS.A" = "RXRA", "RXRA.H13CORE.1.S.C" = "RXRA", "RXRA.H13CORE.2.SM.B" = "RXRA",
  "RXRA.H13CORE.3.P.B" = "RXRA", "RXRA_M03621_3.10" = "RXRA", "RXRA_M03623_3.10" = "RXRA",
  "RXRB.H13CORE.0.P.C" = "RXRB", "RXRB.H13CORE.1.SM.B" = "RXRB", "MA0855.1" = "RXRB", "MA1555.1" = "RXRB",
  "MA0856.1" = "RXRG", "MA1556.1" = "RXRG",
  # Each of the 2 synthetic motifs gets its own family/color rather than being lumped into
  # "PPAR::RXR" with the 3 real heterodimer motifs (MA1148.1/.2, MA0065.1) -- they aren't
  # measured TF binding motifs, they're hand-built/derived references (a synthetic DR1
  # consensus and an averaged PWM), so keeping them visually distinct avoids implying they're
  # on equal footing with the other 30, even though all 32 now share the same genome/background.
  "PPRE_canonical" = "Canonical DR1", "PPRE_consensus_all30" = "Consensus (all 30)"
)
family_of <- function(mid) unname(ifelse(mid %in% names(family_lookup), family_lookup[mid], "other"))
family_colors <- c(PPARA = "#e41a1c", PPARD = "#377eb8", PPARG = "#4daf4a",
                    RXRA = "#ff7f00", RXRB = "#984ea3", RXRG = "#a65628",
                    `PPAR::RXR` = "#000000", `Canonical DR1` = "#17becf",
                    `Consensus (all 30)` = "#bcbd22", other = "grey50")
family_order <- c("PPAR::RXR", "PPARA", "PPARD", "PPARG", "RXRA", "RXRB", "RXRG",
                   "Canonical DR1", "Consensus (all 30)")

motif_meta <- data.table(motif_id = motif_ids, family = sapply(motif_ids, family_of))
motif_meta[, family := factor(family, levels = family_order)]
setorder(motif_meta, family, motif_id)

tracks <- list(GenomeAxisTrack())
for (mid in motif_meta$motif_id) {
  sub <- fam53a_hits[motif_id == mid]
  col <- unname(family_colors[family_of(mid)])
  gr <- GRanges(seqnames = rep("chr4", nrow(sub)),
                ranges = IRanges(start = sub$start, end = sub$end),
                strand = if (nrow(sub) == 0) character(0) else sub$strand)
  seqlevels(gr) <- "chr4"
  tracks[[length(tracks) + 1]] <- AnnotationTrack(
    range = gr, chromosome = "chr4", genome = "hg38",
    name = mid, fill = col, col = col, stacking = "dense"
  )
}

# plotTracks() has no built-in legend for arbitrary per-track fill colors like this (that
# feature only applies to a single track's grouped features). Drawn manually instead: Gviz
# already renders through grid, and plotTracks() returns with the viewport stack back at the
# page root, so a legend can be added directly with grid.rect()/grid.text() in "npc" (0-1)
# page coordinates -- no extra dependency (e.g. gridBase) needed. Only families actually
# present among the plotted motifs get an entry.
library(grid)
draw_family_legend <- function(colors, order, present_families) {
  fams <- order[order %in% present_families]
  n <- length(fams)
  box_w <- 0.16; box_h <- 0.016 * n + 0.014
  x0 <- 0.99 - box_w; y0 <- 0.995
  grid.rect(x = unit(x0 + box_w / 2, "npc"), y = unit(y0 - box_h / 2, "npc"),
            width = unit(box_w, "npc"), height = unit(box_h, "npc"),
            gp = gpar(fill = "white", col = "black", lwd = 0.5))
  grid.text("Motif family", x = unit(x0 + 0.01, "npc"), y = unit(y0 - 0.009, "npc"),
            just = "left", gp = gpar(fontsize = 8, fontface = "bold"))
  for (i in seq_along(fams)) {
    y <- y0 - 0.024 - (i - 1) * 0.016
    grid.rect(x = unit(x0 + 0.014, "npc"), y = unit(y, "npc"),
              width = unit(0.012, "npc"), height = unit(0.012, "npc"),
              gp = gpar(fill = unname(colors[fams[i]]), col = NA))
    grid.text(fams[i], x = unit(x0 + 0.026, "npc"), y = unit(y, "npc"),
              just = "left", gp = gpar(fontsize = 7.5))
  }
}
present_families <- unique(as.character(motif_meta$family))

pdf(file.path(fam53a_out_dir, "FAM53A_promoter_multitrack.pdf"), width = 10, height = 16)
plotTracks(tracks, from = window_start, to = window_end, chromosome = "chr4",
           background.title = "white", col.title = "black", col.axis = "black",
           fontsize = 9, cex.title = 0.7, title.width = 2.2)
draw_family_legend(family_colors, family_order, present_families)
dev.off()

plotTracks(tracks, from = window_start, to = window_end, chromosome = "chr4",
           background.title = "white", col.title = "black", col.axis = "black",
           fontsize = 9, cex.title = 0.7, title.width = 2.2)
draw_family_legend(family_colors, family_order, present_families)
```

# TSS-Centered Motif Binding Density

Using `distanceToTSS` from every motif's promoter-filtered ChIPseeker output (i.e. every
FIMO hit that already fell inside some gene's `[-3000, +1000]` promoter window, genome-wide,
not just FAM53A), binned into 100 bp bins to see whether hits cluster right at the TSS,
spread into the 5' UTR (positive distances), or sit further upstream.

All 32 motifs (including `PPRE_canonical` and `PPRE_consensus_all30`) are pooled below --
now that all 32 share the same unmasked genome/background, there is no longer a reason to
exclude the 2 synthetic motifs from the shared pooled curve.

```{r tss-density-real-only}
tss_density_hits <- all_promoter_hits
tss_density_motif_ids <- motif_ids
cat("TSS-density analysis uses all", length(tss_density_motif_ids), "motifs\n")
```

```{r tss-density-overview, fig.width=8, fig.height=5}
p_density <- ggplot(as.data.frame(tss_density_hits), aes(x = distanceToTSS)) +
  geom_density(fill = "steelblue", alpha = 0.4) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(x = "Distance to TSS (bp)", y = "Density",
       title = paste0("PPAR/RXR motif hit density around annotated TSS (",
                       length(tss_density_motif_ids), " motifs pooled, unmasked genome)")) +
  theme_minimal()
ggsave(file.path(tss_out_dir, "TSS_density_overview.pdf"), p_density, width = 8, height = 5)
p_density
```

```{r tss-density-heatmap, fig.width=9, fig.height=8}
# Unlike the density plot above, this keeps all 32 motifs (incl. PPRE_canonical /
# PPRE_consensus_all30) -- the row-normalized heatmap already shows each motif's own shape
# on its own row, so a background-model mismatch between motifs shows up as a visibly
# different row rather than distorting a shared pooled curve the way it would in a density
# plot with everything summed together.
bin_width <- 100
breaks <- seq(tss_region[1], tss_region[2], by = bin_width)
bin_mids <- breaks[-length(breaks)] + bin_width / 2

hits_binned <- copy(all_promoter_hits)
hits_binned[, bin := cut(distanceToTSS, breaks = breaks, include.lowest = TRUE, labels = FALSE)]
count_mat <- dcast(hits_binned[!is.na(bin)], motif_id ~ bin, fun.aggregate = length, value.var = "bin", fill = 0)
mat_motif_ids <- count_mat$motif_id
count_mat[, motif_id := NULL]
m <- as.matrix(count_mat)
rownames(m) <- mat_motif_ids

full_cols <- as.character(seq_len(length(breaks) - 1))
missing_cols <- setdiff(full_cols, colnames(m))
if (length(missing_cols) > 0) {
  m <- cbind(m, matrix(0, nrow(m), length(missing_cols), dimnames = list(rownames(m), missing_cols)))
}
m <- m[, full_cols, drop = FALSE]
colnames(m) <- bin_mids

# Row-normalized: the *shape* of each motif's distribution around the TSS, independent of
# how many total promoter hits that motif has (which varies ~10x across the 30 motifs).
row_norm <- m / rowSums(m)

pdf(file.path(tss_out_dir, "TSS_density_heatmap.pdf"), width = 9, height = 8)
ht <- Heatmap(row_norm, name = "fraction of\nmotif's hits",
              cluster_columns = FALSE, cluster_rows = TRUE,
              col = colorRamp2(c(0, max(row_norm)), c("white", "firebrick")),
              column_title = "Distance to TSS (bp)",
              row_names_gp = grid::gpar(fontsize = 7),
              heatmap_legend_param = list(title_gp = grid::gpar(fontsize = 8)))
draw(ht)
dev.off()

draw(ht)
```

The heatmap and density plot both show a sharp, shared peak within ~100 bp of the TSS
across nearly all `r length(motif_ids)` motifs, essentially all landing in the *same* few
bins despite very different PWM lengths, half-site spacing and TF families. That similarity
-- not just the peak's existence -- is the reason this is checked against a control below
before reading it as PPAR/RXR-specific biology: motifs that differ that much converging on
the exact same position is at least as consistent with something shared by all 30 (local
promoter sequence composition, or annotation density near well-annotated TSSs) as with 30
independent, sequence-specific enrichments.

## Control: Column-Shuffled Motif

To test that directly, the `PPRE_consensus_all30_motif.meme` PWM (the coverage-aware
consensus of all 30 motifs, exported in the alignment script) was column-shuffled 3 times --
same per-position base composition and overall information content, but the specific
`AGGTCA`-spacer-`AGGTCA` pattern destroyed -- and scanned genome-wide with the exact same
FIMO command (threshold, `--max-stored-scores`, unmasked genome, order-0 background) and the
same ChIPseeker `annotatePeak(tssRegion = c(-3000, 1000))` call used for the real 30 motifs.
If the TSS-centered spike is specific to real PPAR/RXR sequence, the shuffled control should
not reproduce it.

```{r control-shuffled-motif, eval=FALSE}
# Not run as part of the normal knit -- this was a one-off control, run manually via WSL FIMO
# (the project's existing pipeline is Windows-only for the R/ChIPseeker side). Kept here for
# reproducibility rather than folded into the main loop above. (control_shuffled_consensus.meme
# itself is unchanged from the original masked-genome run of this control -- only the genome
# it's scanned against changed, to match the pipeline's unmasked decision.)
library(universalmotif)
cons <- read_meme(file.path(project_dir, "PPRE_consensus_all30_motif.meme"))
if (is.list(cons)) cons <- cons[[1]]
set.seed(42)
make_shuffled <- function(template, name) {
  m <- template
  m@motif <- template@motif[, sample(ncol(template@motif))]
  colnames(m@motif) <- NULL
  m@name <- name
  m
}
shuf_motifs <- list(make_shuffled(cons, "SHUFFLED_CONSENSUS_1"),
                     make_shuffled(cons, "SHUFFLED_CONSENSUS_2"),
                     make_shuffled(cons, "SHUFFLED_CONSENSUS_3"))
write_meme(shuf_motifs, file.path(project_dir, "control_shuffled_consensus.meme"), overwrite = TRUE)

# Then, from a WSL shell in the same working directory as the other 32 motifs:
#   fimo --thresh 1e-4 --max-stored-scores 20000000 --bgfile genome_bg_order0.txt \
#     --oc FIMO_results_no_max/FIMO_results_control_shuffled control_shuffled_consensus.meme \
#     Homo_sapiens.GRCh38.dna.primary_assembly.fa
```

```{r control-annotate-compare, fig.width=8, fig.height=5}
control_fimo_path <- file.path(no_max_fimo_dir, "FIMO_results_control_shuffled", "fimo.tsv")

if (!file.exists(control_fimo_path)) {
  cat("Control fimo.tsv not found at", control_fimo_path,
      "-- run the control-shuffled-motif chunk above (in WSL) first.\n")
} else {
  fimo_ctrl <- fread(control_fimo_path, sep = "\t", header = TRUE, fill = TRUE, colClasses = "character",
                      select = c("motif_id", "sequence_name", "start", "stop", "strand", "score"))
  fimo_ctrl <- fimo_ctrl[!is.na(start) & start != "" & !startsWith(motif_id, "#")]
  fimo_ctrl[, `:=`(start = as.integer(start), stop = as.integer(stop), score = as.numeric(score))]

  gr_ctrl <- GRanges(seqnames = fimo_ctrl$sequence_name,
                      ranges = IRanges(start = fimo_ctrl$start, end = fimo_ctrl$stop),
                      strand = fimo_ctrl$strand, score = fimo_ctrl$score, motif_id = fimo_ctrl$motif_id)
  seqlevelsStyle(gr_ctrl) <- "UCSC"

  peaks_ctrl <- annotatePeak(gr_ctrl, tssRegion = tss_region, TxDb = TxDb.Hsapiens.UCSC.hg38.knownGene,
                              annoDb = "org.Hs.eg.db", sameStrand = TRUE, verbose = FALSE)
  peaks_ctrl_df <- as.data.frame(peaks_ctrl)
  control_promoter_hits <- peaks_ctrl_df[grepl("Promoter", peaks_ctrl_df$annotation), ]

  cat("Control (shuffled) raw genome-wide hits:", nrow(fimo_ctrl), "\n")
  cat("Control promoter-filtered hits:", nrow(control_promoter_hits),
      sprintf(" (%.1f%% of raw)\n", 100 * nrow(control_promoter_hits) / nrow(fimo_ctrl)))

  frac_near_tss <- function(x) mean(abs(x) <= 200)
  cat("Fraction of promoter hits within +/-200bp of TSS -- real:",
      round(frac_near_tss(all_promoter_hits$distanceToTSS), 4),
      " | shuffled control:", round(frac_near_tss(control_promoter_hits$distanceToTSS), 4), "\n")

  compare_df <- rbind(
    data.frame(distanceToTSS = all_promoter_hits$distanceToTSS,
               group = paste0("Real PPAR/RXR motifs (n=", length(motif_ids), ")")),
    data.frame(distanceToTSS = control_promoter_hits$distanceToTSS, group = "Shuffled control (n=3)")
  )
  p_control <- ggplot(compare_df, aes(x = distanceToTSS, fill = group, colour = group)) +
    geom_density(alpha = 0.35, linewidth = 0.8) +
    geom_vline(xintercept = 0, linetype = "dashed") +
    labs(x = "Distance to TSS (bp)", y = "Density",
         title = "Real PPAR/RXR motifs vs. column-shuffled control",
         subtitle = "Same genome, same FIMO threshold/background, same ChIPseeker promoter window") +
    theme_minimal() + theme(legend.position = "bottom", legend.title = element_blank())
  ggsave(file.path(tss_out_dir, "TSS_density_real_vs_shuffled_control.pdf"), p_control, width = 8, height = 5)
  p_control
}
```

**Result: the control reproduces the same TSS-centered peak, if anything slightly more
sharply than the real motifs** (in the run behind this file: 111,968/846,418 = 13.3% of the
shuffled control's raw hits fell in a promoter, vs. 1,695,076/15,945,409 = 10.6% for the real
30 motifs; 30.1% of the shuffled control's promoter hits sat within +/-200bp of the TSS vs.
24.0% for the real motifs). The two density curves are close to indistinguishable.

**That means the TSS-centered spike above should *not* be read as evidence that these
PPAR/RXR PPRE motifs are specifically enriched right at the TSS.** A column-shuffled motif
of matched length and per-position base composition, carrying none of the real
`AGGTCA`-spacer-`AGGTCA` pattern, produces the same pileup under the same FIMO
threshold/background and the same ChIPseeker annotation. The much more likely explanation is
shared by *any* motif scanned this way: local sequence composition right around annotated
TSSs (CpG-island promoters have different base composition than the order-0, genome-wide
background FIMO scores against) and/or the sheer density of overlapping transcript
annotations right at well-annotated TSSs (as FAM53A's 6 overlapping transcripts illustrated
above) inflating how many hits get labeled "close to a TSS" independent of any real
regulatory preference. This also can't speak to whether real PPAR/RXR binding is
promoter-proximal or distal at all -- both this plot and the control are restricted, by
construction, to the subset of hits ChIPseeker already called "Promoter"; genuine PPAR/RXR
ChIP-seq literature generally emphasizes distal-enhancer binding, which this design can't see
either way. A composition- or dinucleotide-matched genomic background (rather than order-0)
and/or a permutation test (shuffling hit *positions* within promoters, not the motif itself)
would be needed before drawing any positional-preference conclusion from this pipeline.
