---
title: "FIMO Motif Analysis with ChIPseeker"
author: "Eli Brouwer"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
---

This file is used to analyse the FIMO files, for objective 1. In this File the FIMO files are annotated with ChIPseeker. And overlapped with GTEx, to create the final genelist.

This version adapts `Objective1_chipseeker_fimo_analysis_with_CISBP.Rmd` to a re-run of all 8
motif batch files (`RXR_PPAR_combined_files_1_4.meme` through `..._27_30.meme`, 30 motifs total,
including the 8 CIS-BP additions) against the local genome-wide FIMO installation, using an
order-0 genome background and `--max-stored-scores 20000000` so no motif's effective significance
threshold is silently tightened by FIMO's default 100,000-score buffer (see the FIMO
reproducibility investigation for why that matters). All logic is unchanged from the source
script except: (1) the FIMO input list now reads the 8 fresh `fimo.tsv` files under
`FIMO_results_no_max/<batch>/` instead of the old webserver `FIMO/` directory + a single
pre-merged CIS-BP file, (2) every output path is redirected into
`FIMO_results_no_max/Objective1_analysis/`, and (3) the "gene-overlap" chunk's ChIPseeker input
directory is fixed to read from the freshly-written `chip_file_directory` instead of the old,
unrelated `ChIPseeker_output` folder. Nothing under the original `Objective1_results` or
`ChIPseeker_output` folders (webserver-based results) is read from or written to.

```{r setup, include=FALSE}
#load in libraries
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(pacman)
p_load(tidyr, ChIPseeker, dplyr, org.Hs.eg.db, TxDb.Hsapiens.UCSC.hg38.knownGene, gt, writexl, pheatmap, stringr, GenomicRanges, BSgenome.Hsapiens.UCSC.hg38, ggvenn, magrittr, rtracklayer, clusterProfiler, tibble, ggpubr, ggrepel, forcats, patchwork, grid, plotly, data.table)
```

# Data Input

```{r load-data}
# Set file paths
# All 8 batch files were re-run locally (no --max-stored-scores truncation, order-0 genome
# background); each batch's own fimo.tsv lives in its own FIMO_results_no_max/<batch>/ folder.
no_max_fimo_dir <- "data/PPAR_RXR_motifs_Hocomoco_Jaspar/FIMO_results_no_max"
fimo_batches <- c("1_4", "5_8", "9_12", "13_16", "17_20", "21_22", "23_26", "27_30")
fimo_file_paths <- file.path(no_max_fimo_dir, fimo_batches, "fimo.tsv")
motif_annotation_path <- "data/annot_file.txt"

# Separate output roots so nothing from the original 22-motif analysis is read from or
# overwritten: a fresh ChIPseeker per-motif output folder, and a fresh results folder mirroring
# the original Objective1_results/ subfolder layout.
results_root <- "data/PPAR_RXR_motifs_Hocomoco_Jaspar/FIMO_results_no_max/Objective1_analysis"
chip_file_directory <- paste0(results_root, "/ChIPseeker_output")
results_dir <- paste0(results_root, "/Objective1_results")
snrna_file_path <- "data/GTEx_non_zeros_prt_coding.csv"

# Unlike the source script, none of these output folders exist yet (this is a fresh
# results_root), so every subfolder any later chunk writes into has to be created up front.
dir.create(chip_file_directory, showWarnings = FALSE, recursive = TRUE)
dir.create(file.path(results_dir, "Bedfiles_output"), showWarnings = FALSE, recursive = TRUE)
dir.create(file.path(results_dir, "Statistics_plots"), showWarnings = FALSE, recursive = TRUE)
dir.create(file.path(results_dir, "Scoring_FIMO"), showWarnings = FALSE, recursive = TRUE)
```

# Annotate Motifs with ChIPseeker

The motifs are loaded in in a GRanges object.

```{r annotate-fimo, verbose = FALSE, message = FALSE}
dir.create(chip_file_directory, showWarnings = FALSE, recursive = TRUE)

# Function to process and annotate FIMO files
# Reads with data.table::fread() instead of read.delim() -- purely a performance swap (the new
# CIS-BP file is 417MB/~5M rows; fread reads it in ~15s vs. several minutes for read.delim, with
# identical resulting columns). fread() does not auto-skip the trailing "# FIMO version..."
# comment lines the way read.table's default comment.char="#" does, so that is replicated
# explicitly via the motif_id filter below. Also writes each motif's promoter-filtered peaks out
# to chip_file_directory in the same format the original ChIPseeker_output/*.txt files use, so
# this run is self-contained rather than depending on pre-existing files.
fimo_analyzer <- function(fimo_file_path, txdb, out_dir = NULL){
  fimo_file <- fread(fimo_file_path, sep = "\t", header = TRUE, fill = TRUE, colClasses = "character")
  fimo_file <- fimo_file[!is.na(fimo_file$start) & fimo_file$start != "" & !startsWith(fimo_file$motif_id, "#"), ]
  fimo_file$start <- as.integer(fimo_file$start)
  fimo_file$stop <- as.integer(fimo_file$stop)
  fimo_file$score <- as.numeric(fimo_file$score)
  fimo_file$`p-value` <- as.numeric(fimo_file$`p-value`)
  fimo_file$`q-value` <- as.numeric(fimo_file$`q-value`)
  gr_fimo <- GRanges(
    seqnames = fimo_file$sequence_name,
    ranges = IRanges(start = fimo_file$start, end = fimo_file$stop),
    strand = fimo_file$strand,
    score = fimo_file$score,
    pvalue = fimo_file$`p-value`,
    qvalue = fimo_file$`q-value`,
    motif_id = fimo_file$motif_id,
    motif_seq = fimo_file$matched_sequence
  )
  seqlevelsStyle(gr_fimo) <- "UCSC"
  gr_fimo_splt <- split(gr_fimo, mcols(gr_fimo)$motif_id)
  summary_list <- list()
  gr_list <- list()


  for (i in seq_along(gr_fimo_splt)){
    gr_current <- gr_fimo_splt[[i]]
    gr_list[[length(gr_list) + 1]] <- gr_current
    #ChIPseeker, specify the search window 3kb up and 1kb down, sameStrand = TRUE
    peaks <- annotatePeak(gr_current, tssRegion = c(-3000, 1000), TxDb = txdb,
                          annoDb = "org.Hs.eg.db", sameStrand = TRUE, verbose = FALSE)

    peaks_df <- as.data.frame(peaks)
    #print(peaks)
    nearby_peaks <- peaks_df[grepl("Promoter", peaks_df$annotation), ] #only keep the genes that are in the promoter

    if (!is.null(out_dir)) {
      write.table(nearby_peaks, file.path(out_dir, paste0(gr_current$motif_id[1], ".txt")), sep = " ")
    }

    #Generate Fimo and chipseeker summary results, per motif
    summary_list[[length(summary_list) + 1]] <- data.frame(
      N_peaks = length(gr_current),
      Min_pvalue = min(gr_current$pvalue),
      Mean_width = mean(width(gr_current)),
      Motif_id = gr_current$motif_id[1],
      Motif_seq = gr_current$motif_seq[1],
      N_genes = length(nearby_peaks$annotation),
      Mean_dist_to_TSS = mean(nearby_peaks$distanceToTSS))
  }
  return(list(summary = summary_list, granges = gr_list))
}

# Run annotation

# A list of FIMO files is created (8 batch files, 30 motifs total, all re-run locally with no
# --max-stored-scores truncation) and the summary list is stored, and the granges objects are
# stored in a list. Each motif's promoter-filtered ChIPseeker output is also written to
# chip_file_directory (fresh for this run -- does not touch the original ChIPseeker_output
# folder).

motif_summary_list <- list()
granges_list <- list()
for (i in seq_along(fimo_file_paths)){
  print(paste0("Fimo file: ", i, " / Out of ", length(fimo_file_paths)))
  this_list <- fimo_analyzer(fimo_file_paths[i], TxDb.Hsapiens.UCSC.hg38.knownGene, out_dir = chip_file_directory)
  motif_summary_list <- c(motif_summary_list, this_list$summary)
  granges_list <- c(granges_list, this_list$granges)
}

granges_fimo <-  do.call(c, granges_list)

print("The results are saved in the granges list")
```

This code is used to create a bedfile with all the separate FIMO hits

```{r bedfile-fimo, verbose = FALSE, message = FALSE,warning=FALSE}
# granges_fimo (== do.call(c, granges_list)) was already built in the annotate-fimo chunk above;
# reusing it here instead of re-running fimo_analyzer() over all 7 files (and re-annotating
# ~6.6M ranges with ChIPseeker a second time) saves a very expensive duplicate pass.
df <- as.data.frame(granges_fimo)

PPRE_FIMO_bed <- data.frame(
  chrom      = df$seqnames,
  chromStart = df$start - 1,
  chromEnd   = df$end,
  name       = if ("motif_id"  %in% names(df)) df$motif_id  else ".",
  score      = if ("score" %in% names(df)) df$score else 0,
  strand     = as.character(df$strand),
  check.names = FALSE
) #makes a bedfile of all FIMO hits.

write.table(PPRE_FIMO_bed, paste0(results_dir, "/Bedfiles_output/PPRE_FIMO_hits_30_motifs.bed"),
            sep = "\t", quote = FALSE,
            row.names = FALSE, col.names = TRUE)
```

# Motif Overlap Heatmap

the overlap between two pairs of motifs is plotted in a heatmap

```{r overlap-heatmap}
#this function calculates the overlap
get_overlap_percent <- function(gr1, gr2, minoverlap) {
  hits <- findOverlaps(gr1, gr2, minoverlap = minoverlap)
  unique_queries <- length(unique(queryHits(hits)))
  
  # % of gr1 that overlaps gr2
  percent <- (unique_queries / length(gr1)) * 100
  return(percent)
}

n <- length(granges_list)
set_names <- names(granges_list)
motif_names <- sapply(granges_list, function(gr) as.character(mcols(gr)$motif_id[1]))
names(granges_list) <- motif_names
overlap_matrix <- matrix(NA, nrow = n, ncol = n, dimnames = list(motif_names, motif_names))

for (i in 1:n) {
  for (j in 1:n) {
    if (i != j) {
      overlap_matrix[i, j] <- get_overlap_percent(granges_list[[i]], granges_list[[j]], minoverlap = 5)
    } else {
      overlap_matrix[i, j] <- 100 #when the pair of the motif combinations is the same, a value of 100 is given
    }
  }
}

overlap_df <- as.data.frame(overlap_matrix)
overlap_df <- cbind(Motif = rownames(overlap_df), overlap_df)

motif_class <- c(
  "MA1148.1" = "Full-PPARA::RXRA","MA1148.2" = "Full-PPARA::RXRA", "MA0065.1" = "Full-PPARG::RXRA",
  "PPARA.H13CORE.0.P.B" = "Half-PPARA", "PPARA.H13CORE.1.P.B" = "Half-PPARA", "PPARD.H13CORE.0.PSM.A" = "Half-PPARD",
  "MA1550.1" = "Half-PPARD","MA1550.2" = "Half-PPARD", "PPARG.H13CORE.0.P.B" = "Half-PPARG",
  "PPARG.H13CORE.1.P.B" = "Half-PPARG", "MA0066.1" = "Half-PPARG", "MA0066.2" = "Half-PPARG",
  "RXRA.H13CORE.0.PS.A" = "Half-RXRA", "RXRA.H13CORE.1.S.C" = "Half-RXRA", "RXRA.H13CORE.2.SM.B" = "Half-RXRA",
  "RXRA.H13CORE.3.P.B" = "Half-RXRA", "RXRB.H13CORE.0.P.C" = "Half-RXRB", "RXRB.H13CORE.1.SM.B" = "Half-RXRB",
  "MA0855.1" = "Half-RXRB", "MA1555.1" = "Half-RXRB", "MA0856.1" = "Half-RXRG", "MA1556.1" = "Half-RXRG", #adding a annotation col of motif-type
  # 8 new CIS-BP motifs: PPARA_M12541/PPARG_M12496-98 are Transfac-sourced but the same
  # GGTCA-AAA-GGTCA half-site architecture as the existing PPARA/PPARG entries; PPARD_M05794/96
  # (Yin et al. 2017) and RXRA_M03621/23 (Jolma et al. 2013) are distinct binding architectures
  # but are kept in their nominal PPARD/RXRA family bucket, matching how they were grouped in
  # the motif-alignment figure (see Objective1_motif_alignment_CISBP_additions.Rmd).
  "PPARA_M12541_3.10" = "Half-PPARA",
  "PPARG_M12496_3.10" = "Half-PPARG", "PPARG_M12497_3.10" = "Half-PPARG", "PPARG_M12498_3.10" = "Half-PPARG",
  "PPARD_M05794_3.10" = "Half-PPARD", "PPARD_M05796_3.10" = "Half-PPARD",
  "RXRA_M03621_3.10" = "Half-RXRA", "RXRA_M03623_3.10" = "Half-RXRA")

motif_groups <- data.frame(Motif = rownames(overlap_df))
motif_groups$Motif_type <- motif_class[motif_groups$Motif] #create a annotation list
rownames(motif_groups) <- motif_groups$Motif
motif_groups %<>%
  dplyr::mutate(Motif_db = case_when(
    grepl("H13CORE", Motif, ignore.case = TRUE) ~ "HOCOMOCO",
    grepl("^MA", Motif, ignore.case = TRUE) ~ "JASPAR",
    grepl("_M[0-9]+_3\\.10$", Motif, ignore.case = TRUE) ~ "CIS-BP",
    TRUE ~ "other")) #adds an annotation row, with motif db, jaspar, hocomoco, and cis-bp

rownames(overlap_df) <- overlap_df$Motif
mat <- as.matrix(overlap_df[, -1])
mat[is.na(mat)] <- 100 #gives the overlapping pairs an overlap of 100%

pheatmap(round(overlap_matrix, 1), cluster_rows = TRUE, cluster_cols = TRUE)
```

```{r motif-jaccard-heatmap}
# ============================================================
# 1. Base-pair Jaccard function
# ============================================================

get_jaccard_percent <- function(gr1, gr2) {
  
  # Merge overlapping ranges within each motif first
  gr1_reduced <- reduce(gr1)
  gr2_reduced <- reduce(gr2)
  
  # Shared genomic sequence
  intersection_ranges <- intersect(gr1_reduced, gr2_reduced)
  intersection_bp <- sum(width(intersection_ranges))
  
  # Total genomic sequence covered by either motif
  union_ranges <- reduce(c(gr1_reduced, gr2_reduced))
  union_bp <- sum(width(union_ranges))
  
  if (union_bp == 0) {
    return(NA_real_)
  }
  
  jaccard_percent <- (intersection_bp / union_bp) * 100
  
  return(jaccard_percent)
}


# ============================================================
# 2. Rename GRanges objects using motif IDs
# ============================================================

motif_names <- vapply(
  granges_list,
  function(gr) as.character(mcols(gr)$motif_id[1]),
  character(1)
)

names(granges_list) <- motif_names

n <- length(granges_list)


# ============================================================
# 3. Calculate symmetric Jaccard matrix
# ============================================================

jaccard_matrix <- matrix(
  NA_real_,
  nrow = n,
  ncol = n,
  dimnames = list(motif_names, motif_names)
)

for (i in seq_len(n)) {
  
  for (j in i:n) {
    
    if (i == j) {
      
      jaccard_value <- 100
      
    } else {
      
      jaccard_value <- get_jaccard_percent(
        granges_list[[i]],
        granges_list[[j]]
      )
    }
    
    # Fill both directions because Jaccard is symmetric
    jaccard_matrix[i, j] <- jaccard_value
    jaccard_matrix[j, i] <- jaccard_value
  }
}


# ============================================================
# 4. Define motif classes
# ============================================================

motif_class <- c(
  "MA1148.1" = "Full-PPARA::RXRA",
  "MA1148.2" = "Full-PPARA::RXRA",
  "MA0065.1" = "Full-PPARG::RXRA",
  
  "PPARA.H13CORE.0.P.B" = "Half-PPARA",
  "PPARA.H13CORE.1.P.B" = "Half-PPARA",
  
  "PPARD.H13CORE.0.PSM.A" = "Half-PPARD",
  "MA1550.1" = "Half-PPARD",
  "MA1550.2" = "Half-PPARD",
  
  "PPARG.H13CORE.0.P.B" = "Half-PPARG",
  "PPARG.H13CORE.1.P.B" = "Half-PPARG",
  "MA0066.1" = "Half-PPARG",
  "MA0066.2" = "Half-PPARG",
  
  "RXRA.H13CORE.0.PS.A" = "Half-RXRA",
  "RXRA.H13CORE.1.S.C" = "Half-RXRA",
  "RXRA.H13CORE.2.SM.B" = "Half-RXRA",
  "RXRA.H13CORE.3.P.B" = "Half-RXRA",
  
  "RXRB.H13CORE.0.P.C" = "Half-RXRB",
  "RXRB.H13CORE.1.SM.B" = "Half-RXRB",
  "MA0855.1" = "Half-RXRB",
  "MA1555.1" = "Half-RXRB",

  "MA0856.1" = "Half-RXRG",
  "MA1556.1" = "Half-RXRG",

  # 8 new CIS-BP motifs (see note in the overlap-heatmap chunk above)
  "PPARA_M12541_3.10" = "Half-PPARA",
  "PPARG_M12496_3.10" = "Half-PPARG",
  "PPARG_M12497_3.10" = "Half-PPARG",
  "PPARG_M12498_3.10" = "Half-PPARG",
  "PPARD_M05794_3.10" = "Half-PPARD",
  "PPARD_M05796_3.10" = "Half-PPARD",
  "RXRA_M03621_3.10" = "Half-RXRA",
  "RXRA_M03623_3.10" = "Half-RXRA"
)


# ============================================================
# 5. Create annotations
# ============================================================

motif_groups <- data.frame(
  Motif = rownames(jaccard_matrix),
  stringsAsFactors = FALSE
)

motif_groups$Motif_type <- unname(
  motif_class[motif_groups$Motif]
)

motif_groups$Motif_db <- case_when(
  grepl("H13CORE", motif_groups$Motif, ignore.case = TRUE) ~ "HOCOMOCO",
  grepl("^MA", motif_groups$Motif, ignore.case = TRUE) ~ "JASPAR",
  grepl("_M[0-9]+_3\\.10$", motif_groups$Motif, ignore.case = TRUE) ~ "CIS-BP",
  TRUE ~ "Other"
)

rownames(motif_groups) <- motif_groups$Motif


# ============================================================
# 6. Set motif order
# ============================================================

motif_type_order <- c(
  "Full-PPARA::RXRA",
  "Full-PPARG::RXRA",
  "Half-PPARA",
  "Half-PPARD",
  "Half-PPARG",
  "Half-RXRA",
  "Half-RXRB",
  "Half-RXRG"
)

motif_groups$Motif_type <- factor(
  motif_groups$Motif_type,
  levels = motif_type_order
)

motif_groups$Motif_db <- factor(
  motif_groups$Motif_db,
  levels = c("HOCOMOCO", "JASPAR", "CIS-BP", "Other")
)

ordered_motifs <- motif_groups %>%
  arrange(Motif_type, Motif_db, Motif) %>%
  pull(Motif)

ordered_motifs <- ordered_motifs[
  ordered_motifs %in% rownames(jaccard_matrix)
]


# ============================================================
# 7. Reorder matrix and annotations
# ============================================================

mat <- round(
  jaccard_matrix[ordered_motifs, ordered_motifs],
  1
)

annotation_df <- motif_groups[
  ordered_motifs,
  c("Motif_db", "Motif_type"),
  drop = FALSE
]

colnames(annotation_df) <- c(
  "Motif database",
  "Motif type"
)


# ============================================================
# 8. Keep only lower triangle
# ============================================================

mat[upper.tri(mat, diag = FALSE)] <- NA


# ============================================================
# 9. Annotation colors
# ============================================================

annotation_colors <- list(
  
  `Motif database` = c(
    "HOCOMOCO" = "#4C78A8",
    "JASPAR"   = "#6CC24A",
    "CIS-BP"   = "#E4A11B",
    "Other"    = "#BDBDBD"
  ),
  
  `Motif type` = c(
    "Full-PPARA::RXRA" = "#F28E7F",
    "Full-PPARG::RXRA" = "#C94C62",
    
    "Half-PPARA" = "#D6B82C",
    "Half-PPARD" = "#8FAF3C",
    "Half-PPARG" = "#D98F39",
    
    "Half-RXRA" = "#3DB7C4",
    "Half-RXRB" = "#377EB8",
    "Half-RXRG" = "#59C3A5"
  )
)


# ============================================================
# 10. Heatmap colors
# ============================================================

heatmap_colors <- colorRampPalette(
  c(
    "#3B6FB6",
    "#A7CFE3",
    "#F7F7D4",
    "#F6B26B",
    "#D73027"
  )
)(100)

heatmap_breaks <- seq(
  0,
  100,
  length.out = 101
)


# ============================================================
# 11. Plot triangular Jaccard heatmap
# ============================================================

heatmap_plot <- pheatmap(
  mat,
  
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  
  annotation_row = annotation_df,
  annotation_col = annotation_df,
  annotation_colors = annotation_colors,
  
  color = heatmap_colors,
  breaks = heatmap_breaks,
  
  na_col = "white",
  border_color = NA,
  
  fontsize = 8,
  fontsize_row = 7,
  fontsize_col = 7,
  angle_col = 90,
  
  main = "Pairwise motif-hit Jaccard overlap (%)",
  
  silent = TRUE
)


# ============================================================
# 12. Save as PDF
# ============================================================

pdf(
  file = paste0(results_dir, "/motif_jaccard_overlap_triangular.pdf"),
  width = 12,
  height = 10,
  useDingbats = FALSE
)

grid.newpage()
grid.draw(heatmap_plot$gtable)

dev.off()


# ============================================================
# 13. Display in R
# ============================================================

grid.newpage()
grid.draw(heatmap_plot$gtable)
```

```{r motif-heatmap-saveblock}
# Same knitr device-redirection issue as the motif-alignment pdf() sink fixed earlier in
# Objective1_motif_alignment_CISBP_additions.Rmd: an explicit pdf()+dev.off() around a
# grid-based plot (here pheatmap()), coming right after another grid.draw() call in the
# previous chunk, silently draws onto knitr's own tracked device instead of this freshly
# opened one, leaving the file with zero pages. dev.control()/recordPlot()/replayPlot() avoids
# it by drawing on the normal device (which works) and then replaying that onto the real sink.
dev.control("enable")
pheatmap(round(mat, 1), cluster_rows = TRUE, cluster_cols = TRUE,
         display_numbers = FALSE, fontsize = 10, annotation_row = motif_groups["Motif_type"], annotation_col =  motif_groups["Motif_db"])
recorded_corr_heatmap <- recordPlot()
pdf("data/PPAR_RXR_motifs_Hocomoco_Jaspar/FIMO_results_no_max/Objective1_analysis/Objective1_results/correlation_heatmap.pdf", height =11.7 , width = 12.3)
replayPlot(recorded_corr_heatmap)
dev.off()

write_xlsx(overlap_df, path = "data/PPAR_RXR_motifs_Hocomoco_Jaspar/FIMO_results_no_max/Objective1_analysis/Objective1_results/pairwise_comparison_fimo.xlsx")
```

# Summary Table of FIMO and ChIPseeker results

```{r summary-table}
motif_summary_df <- do.call(rbind, motif_summary_list) 
colnames(motif_summary_df) <- c("N_amount", "Min_pvalue", "Mean_width", "Motif_name", "Sequence", "N_amount_genes", "Mean_dist_tss")
rownames(motif_summary_df) <- motif_summary_df$Motif_name

#Combined table with both chipseeker and FIMO results summary
FIMO_chipseeker_results <- motif_summary_df %>%
  left_join(motif_groups, by = c("Motif_name" = "Motif")) %>%
  mutate(Motif_gene = str_remove(Motif_type, "^(Half-|Full-)")) %>% 
  dplyr::select(Motif_gene, Motif_name, N_amount, N_amount_genes, Mean_dist_tss, Mean_width) %>%
  gt(rowname_col = "Motif_name") %>%
  cols_label(
    Motif_gene = "Motif gene",
    N_amount = "Peaks",
    Mean_width = "Mean Width",
    N_amount_genes = "Amount of genes",
    Mean_dist_tss = "Average to TSS"
  ) %>%
  grand_summary_rows(
    columns = c(N_amount, Mean_width,N_amount_genes, Mean_dist_tss  ),
    fns = list(
      "Mean (SD)" = ~ sprintf("%.1f (%.1f)", mean(., na.rm = TRUE), sd(., na.rm = TRUE))
      )
  ) %>%
  fmt_number(columns = c("N_amount", "Mean_width", "N_amount_genes", "Mean_dist_tss"), decimals = 0)
FIMO_chipseeker_results 

gtsave(FIMO_chipseeker_results, "data/PPAR_RXR_motifs_Hocomoco_Jaspar/FIMO_results_no_max/Objective1_analysis/Objective1_results/motif_results_table.pdf", expand = 2)

```

```{r write-table-xlsx}
display_df_num <-
  motif_summary_df %>%
  left_join(motif_groups, by = c("Motif_name" = "Motif")) %>%
  mutate(Motif_gene = str_remove(Motif_type, "^(Half-|Full-)")) %>%
  transmute(
    `Motif name`      = Motif_name,
    `Motif gene`      = Motif_gene,
    `Peaks`           = as.numeric(N_amount),
    `Amount of genes` = as.numeric(N_amount_genes),
    `Average to TSS`  = as.numeric(Mean_dist_tss),
    `Mean Width`      = as.numeric(Mean_width)
  )

# Compute mean(SD) on numeric columns
ms <- function(x) sprintf("%.1f (%.1f)", mean(x, na.rm = TRUE), sd(x, na.rm = TRUE))
sum_row <- tibble(
  `Motif name`      = "Mean (SD)",
  `Motif gene`      = "\u2014",
  `Peaks`           = ms(display_df_num$`Peaks`),
  `Amount of genes` = ms(display_df_num$`Amount of genes`),
  `Average to TSS`  = ms(display_df_num$`Average to TSS`),
  `Mean Width`      = ms(display_df_num$`Mean Width`)
)

fmt_int <- function(x) formatC(x, format = "d", big.mark = ",")
fmt_1   <- function(x) formatC(x, format = "f", digits = 1)

display_df_chr <- display_df_num %>%
  mutate(
    `Peaks`           = fmt_int(`Peaks`),
    `Amount of genes` = fmt_int(`Amount of genes`),
    `Average to TSS`  = paste0(ifelse(`Average to TSS` < 0, "\u2212", ""), fmt_1(abs(`Average to TSS`))),
    `Mean Width`      = fmt_1(`Mean Width`)
  )


display_df_sum <- bind_rows(display_df_chr, sum_row)

write_xlsx(list("PPRE table" = display_df_sum),
           path = "data/PPAR_RXR_motifs_Hocomoco_Jaspar/FIMO_results_no_max/Objective1_analysis/Objective1_results/motif_results_table.xlsx")

```

# Expression Overlap with GTEx Cardiomyocyte Genes

```{r gene-overlap}
# Build the combined gene list from all ChIPseeker output files.
# NOTE: the source script reassigned chip_file_directory here to the old webserver-based
# ChIPseeker_output folder, silently discarding the freshly-written no-max output set by the
# load-data chunk. That reassignment is intentionally dropped here -- this chunk must read the
# annotation just written for the no-max, uncapped FIMO run, not the old capped one.
chip_file_list <- list.files(chip_file_directory)
chip_file_list_txt <- chip_file_list[str_ends(chip_file_list, ".txt")]

#for all motifs
chip_data_list <- list()

for (i in 1:length(chip_file_list_txt)){
  chiptable <- read.delim(paste0(chip_file_directory, "/", chip_file_list_txt[i]), sep = " ", header = TRUE)
  chip_data_list[[i]] <- chiptable
}

chip_all <- do.call(rbind, chip_data_list) #this is a dataframe with all the genes in there

gene_type <- AnnotationDbi::select(
  org.Hs.eg.db,
  keys = unique(na.omit(chip_all$ENSEMBL)),
  keytype = "ENSEMBL",
  columns = c("ENSEMBL", "GENETYPE")
)

# Join gene type and keep only protein-coding genes
chip_all <- chip_all %>%
  left_join(gene_type, by = "ENSEMBL") %>%
  filter(GENETYPE == "protein-coding")

overlapping_genes <- chip_all %>%
  dplyr::select(c(seqnames, motif_id, annotation, 
                  geneId, transcriptId, ENSEMBL, 
                  SYMBOL,  pvalue , qvalue, distanceToTSS )) %>%
  distinct(transcriptId, motif_id, .keep_all = TRUE) %>%
  add_count(transcriptId, name = "n_motifs") %>%
  distinct(ENSEMBL, .keep_all = TRUE) %>%
  dplyr::arrange(desc(n_motifs))
```

```{r add-GTEX-nozero}
##add snRNA seq data to overlap the expressed cardiomyocyte genes
GTEX_snRNA_cardiomyocyte_nonzero <- read.csv("data/GTEX_single_nuclear_data/GTEx_non_zeros_ensembl.csv") %>%
  left_join(overlapping_genes, by = c( "ensembl_gene_id" ="ENSEMBL"), keep = TRUE) %>%
  drop_na(c(ensembl_gene_id,ENSEMBL)) -> motif_hits_cardiomyocytes_nonzero 


write.csv(motif_hits_cardiomyocytes_nonzero, paste0(results_dir, "/Motifs_GTEX_cm_expressed.csv"))
```

```{r add-GTEX-top20%}
##add snRNA seq data to overlap the expressed cardiomyocyte genes
read.csv("data/GTEX_single_nuclear_data/GTEx_top20%_ensembl.csv") %>%
  left_join(overlapping_genes, by = c( "ensembl_gene_id" ="ENSEMBL"), keep = TRUE) %>%
  drop_na(c(ensembl_gene_id,ENSEMBL)) -> motif_hits_cardiomyocytes_top20


write.csv(motif_hits_cardiomyocytes_top20, paste0(results_dir, "/Motifs_GTEX_cm_expressed_top20.csv"))
```

# Venn Diagram of Gene Sets

```{r venn-diagram}
#venn_diagram
expression_cardiomyocytes_nozero <- read.csv( "data/GTEX_single_nuclear_data/GTEx_non_zeros_ensembl.csv" )
GTEX_genes <- read.csv("data/GTEX_single_nuclear_data/GTEx_8_genes.csv")
venn_genelist <- list( 'GTEX genes' = unique(GTEX_genes$gene_name), 'Expressed cardiomyocyte genes' = sort(unique(na.omit(expression_cardiomyocytes_nozero$HGNC_gene_name))), 'All Motif PPRE genes' = sort((unique(chip_all$SYMBOL))))

venn_diagram <- ggvenn(
  venn_genelist,
  show_percentage = FALSE,
  fill_color = c("white", "white", "white"),
  stroke_size = 1,
  set_name_size = 2.8,
  text_size = 2.8
)
venn_diagram

pdf("data/PPAR_RXR_motifs_Hocomoco_Jaspar/FIMO_results_no_max/Objective1_analysis/Objective1_results/venn_diagram.pdf", width = 8.27/2, 11.69/4)
venn_diagram
dev.off()
```

# Get the overlap between HCM and PLN regions and FIMO peaks

```{r overlap-regions}
#first load in the differentially expressed peak files
HCM_differentially_regions <- readxl::read_xlsx("data/PPAR_RXR_motifs_Hocomoco_Jaspar/HCM_differentially_regions.xlsx", sheet = "Supplementary Table 2A")
grange_HCM_differentially_regions <- GRanges(
  seqnames = HCM_differentially_regions$`Chromosome number`,
  ranges = IRanges(start = HCM_differentially_regions$`Region start`, end = HCM_differentially_regions$`Region end`),
  pvalue = HCM_differentially_regions$`P-adj`,
  motif_id = "HCM_hearts" ) #Convert the HCM_regions to GRanges
  
PLN_differentially_regions <-  readxl::read_xlsx("data/PPAR_RXR_motifs_Hocomoco_Jaspar/PLN_differentially_regions.xlsx", skip = 1) %>%
  filter(!is.na(`Region start†`))

grange_PLN_differentially_regions <- GRanges(
  seqnames = PLN_differentially_regions$Chromosome,
  ranges = IRanges(start = PLN_differentially_regions$`Region start†`, end = PLN_differentially_regions$`Region end†`),
  pvalue = PLN_differentially_regions$`Adjusted p-value§`,
  motif_id = "PLN_hearts") #Convert the PLN_regions to GRanges

#this peak data is in hg19 assembly, this is converted to hg38
#first a chain file is downloaded, this needs to be unzipped
#chain_url <- "http://hgdownload.cse.ucsc.edu/goldenpath/hg19/liftOver/hg19ToHg38.over.chain.gz"
#chain_file <- "hg19ToHg38.over.chain.gz"
#if (!file.exists(chain_file)) {
#  download.file(chain_url, chain_file)
#}

chain <- rtracklayer::import.chain("data/PPAR_RXR_motifs_Hocomoco_Jaspar/hg19ToHg38.over.chain")

grange_HCM_differentially_regions_hg38 <- liftOver(grange_HCM_differentially_regions, chain) 
grange_PLN_differentially_regions_hg38 <- liftOver(grange_PLN_differentially_regions, chain)

overlap_df_regions <- data.frame()

for (i in 1:length(granges_list)){
  total_hits <- length(granges_list[[i]])
  overlap_hcm <- findOverlaps(grange_HCM_differentially_regions_hg38, granges_list[[i]], ignore.strand = TRUE, minoverlap = 1)
  overlap_pln <- findOverlaps(grange_PLN_differentially_regions_hg38, granges_list[[i]], ignore.strand = TRUE, minoverlap = 1)
  
  # Add a new row for each motif
  overlap_df_regions <- rbind(overlap_df_regions, data.frame(
    motif_id = names(granges_list)[i],
    total_hits = total_hits,
    overlap_hcm = length(queryHits(overlap_hcm)),
    overlap_pln = length(queryHits(overlap_pln))
  ))
}

overlap_df_regions %<>%
  pivot_longer(c("total_hits", "overlap_hcm", "overlap_pln") )


Stackplot <- ggplot(overlap_df_regions, aes(x = motif_id, y = value, fill = name)) +
  geom_bar(stat = "identity", position = "stack") +
  scale_fill_manual(
    values = c("total_hits" = "#4A90E2",       # blue
               "overlap_pln" = "#50E3C2",      # teal
               "overlap_hcm" = "#D0021B"),     # red
    labels = c("total_hits" = "Total Hits",
               "overlap_pln" = "Overlap PLN",
               "overlap_hcm" = "Overlap HCM")
  ) + theme_bw() + scale_y_continuous(expand = c(0, 0))+
  theme( panel.border = element_blank(), panel.grid.major = element_blank(),
                 panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.title = element_blank()
  ) +
  labs(x = "Motif ID", y = "Amount of FIMO peaks")
Stackplot

#Calculate overlap percentages for article
total_hcm <- length(grange_HCM_differentially_regions_hg38)
total_pln <- length(grange_PLN_differentially_regions_hg38)

overlap_percent_df <- overlap_df_regions %>%
  pivot_wider(names_from = name, values_from = value, values_fn = list) %>%
  unnest(cols = c(total_hits, overlap_pln, overlap_hcm)) %>%
  mutate(
    percent_overlap_pln = (overlap_pln / total_pln) * 100,
    percent_overlap_hcm = (overlap_hcm / total_hcm) * 100)

pln_mean <- mean(overlap_percent_df$percent_overlap_pln, na.rm = TRUE)
pln_sd   <- sd(overlap_percent_df$percent_overlap_pln, na.rm = TRUE)

hcm_mean <- mean(overlap_percent_df$percent_overlap_hcm, na.rm = TRUE)
hcm_sd   <- sd(overlap_percent_df$percent_overlap_hcm, na.rm = TRUE)

cat(sprintf("PLN-specific H3K4me3 regions overlapping with the FIMO regions: %.2f%% ± %.2f SD\n", pln_mean, pln_sd)) 
cat(sprintf("HCM-specific H3K27ac regions overlapping with the FIMO regions: %.2f%% ± %.2f SD\n", hcm_mean, hcm_sd))
```

# Statistics Boxplots

This section explores the distribution of FIMO peak counts and motif widths across motif types, databases, and experimental data sources.

```{r make-scatter_df}
measuring_method <- tibble::tribble(
  ~Motif_name,                         ~Data_type_source,                                 ~Data_type_source_simple,
  "MA1148.1",                          "SMILE-seq",                                       "SMILE-seq",
  "MA1148.2",                          "SMILE-seq",                                       "SMILE-seq",
  "MA0065.1",                          "SELEX",                                           "SELEX",
  "PPARA.H13CORE.0.P.B",              "ChIP-Seq",                                        "ChIP-Seq",
  "PPARA.H13CORE.1.P.B",              "ChIP-Seq",                                        "ChIP-Seq",
  "PPARD.H13CORE.0.PSM.A",            "ChIP-Seq + HT-SELEX + Methyl-HT-SELEX",           "SELEX",
  "MA1550.1",                          "HT-SELEX",                                        "SELEX",
  "MA1550.2",                          "HT-SELEX",                                        "SELEX",
  "PPARG.H13CORE.0.P.B",              "ChIP-Seq",                                        "ChIP-Seq",
  "PPARG.H13CORE.1.P.B",              "ChIP-Seq",                                        "ChIP-Seq",
  "MA0066.1",                          "SELEX",                                           "SELEX",
  "MA0066.2",                          "SELEX",                                           "SELEX",
  "RXRA.H13CORE.0.PS.A",              "ChIP-Seq + HT-SELEX",                             "SELEX",
  "RXRA.H13CORE.1.S.C",               "HT-SELEX",                                        "SELEX",
  "RXRA.H13CORE.2.SM.B",              "HT-SELEX + Methyl-HT-SELEX",                      "SELEX",
  "RXRA.H13CORE.3.P.B",               "ChIP-Seq",                                        "ChIP-Seq",
  "RXRB.H13CORE.0.P.C",               "ChIP-Seq",                                        "ChIP-Seq",
  "RXRB.H13CORE.1.SM.B",              "HT-SELEX + Methyl-HT-SELEX",                      "SELEX",
  "MA0855.1",                          "HT-SELEX",                                        "SELEX",
  "MA1555.1",                          "HT-SELEX",                                        "SELEX",
  "MA0856.1",                          "HT-SELEX",                                        "SELEX",
  "MA1556.1",                          "HT-SELEX",                                        "SELEX",
  # 8 new CIS-BP motifs: PPARA_M12541/PPARG_M12496-98 are Transfac-sourced (V$PPARA_01,
  # V$PPARG_01/02/03); PPARD_M05794/96 are Yin et al. 2017 HT-SELEX/Methyl-HT-SELEX; RXRA_M03621/23
  # are Jolma et al. 2013 HT-SELEX (same source type as the existing JASPAR HT-SELEX entries).
  "PPARA_M12541_3.10",                "Transfac",                                        "Transfac",
  "PPARG_M12496_3.10",                "Transfac",                                        "Transfac",
  "PPARG_M12497_3.10",                "Transfac",                                        "Transfac",
  "PPARG_M12498_3.10",                "Transfac",                                        "Transfac",
  "PPARD_M05794_3.10",                "HT-SELEX + Methyl-HT-SELEX",                      "SELEX",
  "PPARD_M05796_3.10",                "HT-SELEX + Methyl-HT-SELEX",                      "SELEX",
  "RXRA_M03621_3.10",                 "HT-SELEX",                                        "SELEX",
  "RXRA_M03623_3.10",                 "HT-SELEX",                                        "SELEX"
)
#Make scatterplot mean width of motif vs number of peaks
motif_summary_df %>% 
  dplyr::select(c(Motif_name, Mean_width, N_amount)) %>%
  left_join(motif_groups, by = c("Motif_name" = "Motif")) %>%
  dplyr::mutate(Motif_type_simple = case_when(
    grepl("^Full", Motif_type, ignore.case = TRUE) ~ "Full-PPAR:RXR",
    grepl("RXR", Motif_type, ignore.case = TRUE) ~ "Half-RXR",
    grepl("PPAR", Motif_type, ignore.case = TRUE) ~ "Half-PPAR",
    TRUE ~ "other")) %>%
  dplyr::mutate(Motif_db = case_when(
    grepl("H13CORE", Motif_name, ignore.case = TRUE) ~ "HOCOMOCO",
    grepl("^MA", Motif_name, ignore.case = TRUE) ~ "JASPAR",
    grepl("_M[0-9]+_3\\.10$", Motif_name, ignore.case = TRUE) ~ "CIS-BP",
    TRUE ~ "other")) %>%
  left_join(measuring_method, by = "Motif_name") %>%
  mutate(
    Data_type_source = factor(Data_type_source, levels = c(
      "ChIP-Seq + HT-SELEX",
      "ChIP-Seq",
      "ChIP-Seq + HT-SELEX + Methyl-HT-SELEX",
      "HT-SELEX",
      "SELEX",
      "HT-SELEX + Methyl-HT-SELEX",
      "SMILE-seq",
      "Transfac")), #Reorder the names
    Data_type_source = fct_recode(Data_type_source, 
                                  "ChIP-Seq \n+ HT-SELEX" = "ChIP-Seq + HT-SELEX",
                                  "HT-SELEX \n+ Methyl-HT-\nSELEX" = "HT-SELEX + Methyl-HT-SELEX",
                                  "ChIP-Seq \n+ HT-SELEX \n+ Methyl-HT-SELEX" = "ChIP-Seq + HT-SELEX + Methyl-HT-SELEX"), #Reorder the names --> cause them to be split on newline
    Motif_type = factor(Motif_type, levels = c("Half-PPARA", 
                                               "Half-PPARD", 
                                               "Half-PPARG", 
                                               "Half-RXRA", 
                                               "Half-RXRB", 
                                               "Half-RXRG", 
                                               "Full-PPARA::RXRA", 
                                               "Full-PPARG::RXRA")),  #Reordering
    Motif_type = gsub("^(Half|Full)-", "\\1-\n", Motif_type) #Adding a newline
  ) %>%
   mutate(Synthetic_or_Not = case_when(
    grepl("SMILE-seq", Data_type_source, ignore.case = TRUE) ~ "Synthetic",
    grepl("SELEX", Data_type_source, ignore.case = TRUE) & !grepl("ChIP", Data_type_source, ignore.case = TRUE) ~ "Synthetic",
    grepl("ChIP-Seq", Data_type_source, ignore.case = TRUE) & grepl("SELEX", Data_type_source, ignore.case = TRUE) ~ "Mixed",
    grepl("ChIP-Seq", Data_type_source, ignore.case = TRUE) ~ "Non-synthetic",
    TRUE ~ "Unknown"
  )) -> scatter_df
```

```{r motif-database-plot}
motif_order <- c(
  # PPAR half-sites
  "PPARA",
  "PPARD",
  "PPARG",

  # RXR half-sites
  "RXRA",
  "RXRB",
  "RXRG",

  # All possible PPAR:RXR heterodimers
  "PPARA::RXRA",
  "PPARA::RXRB",
  "PPARA::RXRG",
  "PPARD::RXRA",
  "PPARD::RXRB",
  "PPARD::RXRG",
  "PPARG::RXRA",
  "PPARG::RXRB",
  "PPARG::RXRG"
)

motif_database_plot_df <- scatter_df %>%
  # This chart is a two-way JASPAR-vs-HOCOMOCO diverging comparison by design (title below);
  # the 8 new CIS-BP motifs don't fit a left/right diverging layout with a 3rd category, so they
  # are excluded here specifically (they are still fully included in every other plot/table).
  filter(Motif_db %in% c("JASPAR", "HOCOMOCO")) %>%
  mutate(
    # Remove line breaks
    Motif_type_clean = str_replace_all(
      as.character(Motif_type),
      "\n",
      ""
    ),

    # Remove Half- and Full- prefixes
    Motif_group = str_remove(
      Motif_type_clean,
      "^(Half-|Full-)"
    ),

    # Convert double colon to single colon for plotting
    Motif_group = str_replace_all(
      Motif_group,
      "::",
      "::"
    )
  ) %>%

  # Count motifs per motif group and database
  count(
    Motif_group,
    Motif_db,
    name = "N_motifs"
  ) %>%

  # Add all missing motif-group/database combinations
  complete(
    Motif_group = motif_order,
    Motif_db = c("JASPAR", "HOCOMOCO"),
    fill = list(N_motifs = 0)
  ) %>%

  mutate(
    Motif_group = factor(
      Motif_group,
      levels = motif_order
    ),

    # JASPAR to the left, HOCOMOCO to the right
    Plot_value = if_else(
      Motif_db == "JASPAR",
      -N_motifs,
      N_motifs
    )
  )

database_plot <- ggplot(
  motif_database_plot_df,
  aes(
    x = Plot_value,
    y = Motif_group,
    fill = Motif_db
  )
) +
  geom_col(
    width = 0.65,
    position = "identity"
  ) +
  geom_vline(
    xintercept = 0,
    linewidth = 0.4
  ) +
  scale_fill_manual(
    values = c(
      "JASPAR" = "#377EB8",
      "HOCOMOCO" = "#FF5A4E"
    )
  ) +
  scale_x_continuous(
    breaks = seq(-4, 4, by = 1),
    labels = abs,
    limits = c(-4, 4),
    expand = expansion(mult = c(0.03, 0.03))
  ) +
  labs(
    title = "JASPAR and HOCOMOCO",
    x = "Number of motifs",
    y = NULL,
    fill = NULL
  ) +
  theme_classic(base_size = 11) +
  theme(
    legend.position = "top",
    legend.justification = "left",
    plot.title = element_text(
      size = 11,
      hjust = 0
    ),
    axis.text.y = element_text(
      size = 9,
      colour = "black"
    ),
    axis.text.x = element_text(
      size = 9,
      colour = "black"
    ),
    axis.line.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.major.x = element_line(
      colour = "grey85",
      linewidth = 0.4
    )
  )

database_plot


database_plot <- ggplot(
  motif_database_plot_df,
  aes(
    x = Plot_value,
    y = factor(Motif_group, levels = rev(motif_order)),
    fill = Motif_db
  )
) +
  geom_col(
    width = 0.65,
    position = "identity"
  ) +
  geom_vline(
    xintercept = 0,
    linewidth = 0.4
  ) +
  scale_fill_manual(
    values = c(
      "JASPAR" = "#377EB8",
      "HOCOMOCO" = "#FF5A4E"
    )
  ) +
  scale_x_continuous(
    breaks = seq(-4, 4, by = 1),
    labels = abs,
    limits = c(-4, 4),
    expand = expansion(mult = c(0.03, 0.03))
  ) +
  labs(
    title = "JASPAR and HOCOMOCO",
    x = "Number of motifs",
    y = NULL,
    fill = NULL
  ) +
  theme_classic(base_size = 11) +
  theme(
    legend.position = "top",
    legend.justification = "left",
    axis.text.y = element_text(
      size = 9,
      colour = "black"
    ),
    axis.line.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.major.x = element_line(
      colour = "grey85",
      linewidth = 0.4
    )
  )

pdf("data/PPAR_RXR_motifs_Hocomoco_Jaspar/FIMO_results_no_max/Objective1_analysis/Objective1_results/Statistics_plots/motif_overview.pdf", width = 8.27/2, height= 11.69/4)

database_plot

dev.off()

ggsave("data/PPAR_RXR_motifs_Hocomoco_Jaspar/FIMO_results_no_max/Objective1_analysis/Objective1_results/Statistics_plots/motif_overview.pdf", database_plot, width = 16, height = 4)
```

Asses nomality Since sample size is 21, we need to check nomality of the continous variables. We do this by the shapiro-wilks test and qq-plot

```{r statistic-nomality}
print("The shapiro-wilk test result of width is")
shapiro.test(scatter_df$Mean_width)
print("So the data is normally distributed")

print("The shapiro-wilk test result of n_amount of peaks is")
shapiro.test(scatter_df$N_amount)
print("So the data is normally distributed")

check_normality <- function(df, group_var) {
  df %>%
    group_by({{ group_var }}) %>%
    summarise(
      n = n(),
      shapiro_p_width  = if (n() >= 3) shapiro.test(Mean_width)$p.value else NA_real_,
      shapiro_p_amount = if (n() >= 3) shapiro.test(N_amount)$p.value else NA_real_,
      .groups = "drop"
    )
}

# Run for different grouping variables
check_normality(scatter_df, Motif_type_simple)
check_normality(scatter_df, Motif_db)
check_normality(scatter_df, Data_type_source)
check_normality(scatter_df, Motif_type)

ggplot(scatter_df, aes(sample = Mean_width)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  labs(title = "Q-Q plots of Mean_width")

# Q-Q plots for N_amount per group
ggplot(scatter_df, aes(sample = N_amount)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  labs(title = "Q-Q plots of N_amount")
```

# Styled Statistics Boxplots

```{r statistic-boxplots}
# ============================================================
# Colour palettes
# ============================================================

database_colors <- c(
  "HOCOMOCO" = "#6F6F6F",
  "JASPAR"   = "#B5B5B5",
  "CIS-BP"   = "#E4A11B"
)

motif_family_colors <- c(
  "PPAR"       = "#4F86D9",
  "PPAR family" = "#4F86D9",
  "PPAR (family)" = "#4F86D9",
  "RXR"        = "#D96C6C",
  "RXR family" = "#D96C6C",
  "RXR (family)" = "#D96C6C",
  "PPAR::RXR"  = "#9A86CC",
  "PPAR:RXR"   = "#9A86CC"
)

motif_type_colors <- c(
  "PPARA"      = "#73A2E6",
  "PPARD"      = "#8CB4EE",
  "PPARG"      = "#B7D0F5",
  "RXRA"       = "#E58C8C",
  "RXRB"       = "#EDAAAA",
  "RXRG"       = "#F4C9C9",
  "PPAR::RXR"  = "#9A86CC",
  "PPAR:RXR"   = "#9A86CC"
)

neutral_colors <- c(
  "#707070",
  "#A0A0A0",
  "#C8C8C8",
  "#E0E0E0"
)

plot_theme <- theme_classic() +
  theme(
    panel.border = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.line = element_line(colour = "black"),
    legend.title = element_blank(),
    legend.position = "none"
  )


# ============================================================
# Scatterplot
# ============================================================

scatter_df_wdth_amount <- ggplot(
  scatter_df,
  aes(
    x = Mean_width,
    y = N_amount,
    colour = Motif_type
  )
) +
  geom_point(size = 3) +
  geom_text_repel(
    aes(label = Motif_type),
    size = 3,
    show.legend = FALSE
  ) +
  geom_smooth(
    aes(group = 1),
    method = "lm",
    se = FALSE,
    colour = "black"
  ) +
  stat_cor(
    aes(group = 1),
    method = "pearson",
    label.x.npc = "left",
    label.y.npc = "top",
    label.sep = ", ",
    label.y = max(scatter_df$N_amount, na.rm = TRUE),
    colour = "black"
  ) +
  scale_colour_manual(
    values = motif_type_colors,
    na.value = "#7F7F7F"
  ) +
  labs(
    x = "Motif width (bp)",
    y = "Number of FIMO peaks (genome wide)"
  ) +
  plot_theme

scatter_df_wdth_amount


# ============================================================
# Motif database
# ============================================================

boxplt_db_number <- ggboxplot(
  scatter_df,
  x = "Motif_db",
  y = "N_amount",
  fill = "Motif_db",
  color = "Motif_db",
  palette = database_colors,
  add = "jitter",
  width = 0.5,
  add.params = list(size = 2, alpha = 0.8)
) +
  stat_compare_means(
    method = "t.test",
    label.y = max(scatter_df$N_amount, na.rm = TRUE)
  ) +
  labs(
    x = "Motif database",
    y = "Number of FIMO peaks"
  ) +
  plot_theme

boxplt_db_number


boxplt_db_wdth <- ggboxplot(
  scatter_df,
  x = "Motif_db",
  y = "Mean_width",
  fill = "Motif_db",
  color = "Motif_db",
  palette = database_colors,
  add = "jitter",
  width = 0.5,
  add.params = list(size = 2, alpha = 0.8)
) +
  stat_compare_means(
    method = "t.test",
    label.y = max(scatter_df$Mean_width, na.rm = TRUE)
  ) +
  labs(
    x = "Motif database",
    y = "Motif width (bp)"
  ) +
  plot_theme

boxplt_db_wdth


# ============================================================
# Synthetic versus non-synthetic
# ============================================================

synthetic_colors <- setNames(
  neutral_colors[
    seq_len(length(unique(scatter_df$Synthetic_or_Not)))
  ],
  unique(scatter_df$Synthetic_or_Not)
)

boxplt_synthetic_number <- ggboxplot(
  scatter_df,
  x = "Synthetic_or_Not",
  y = "N_amount",
  fill = "Synthetic_or_Not",
  color = "Synthetic_or_Not",
  palette = synthetic_colors,
  add = "jitter",
  width = 0.5,
  add.params = list(size = 2, alpha = 0.8)
) +
  stat_compare_means(
    method = "kruskal.test",
    label.y = max(scatter_df$N_amount, na.rm = TRUE)
  ) +
  labs(
    x = "Synthetic",
    y = "Number of FIMO peaks"
  ) +
  plot_theme

boxplt_synthetic_number


boxplt_synthetic_wdth <- ggboxplot(
  scatter_df,
  x = "Synthetic_or_Not",
  y = "Mean_width",
  fill = "Synthetic_or_Not",
  color = "Synthetic_or_Not",
  palette = synthetic_colors,
  add = "jitter",
  width = 0.5,
  add.params = list(size = 2, alpha = 0.8)
) +
  stat_compare_means(
    method = "kruskal.test",
    label.y = max(scatter_df$Mean_width, na.rm = TRUE)
  ) +
  labs(
    x = "Synthetic",
    y = "Motif width (bp)"
  ) +
  plot_theme

boxplt_synthetic_wdth


# ============================================================
# Simplified motif family
# ============================================================

boxplt_mottyp_simple_number <- ggboxplot(
  scatter_df,
  x = "Motif_type_simple",
  y = "N_amount",
  fill = "Motif_type_simple",
  color = "Motif_type_simple",
  palette = motif_family_colors,
  add = "jitter",
  width = 0.5,
  add.params = list(size = 2, alpha = 0.8)
) +
  stat_compare_means(
    method = "kruskal.test",
    label.y = max(scatter_df$N_amount, na.rm = TRUE)
  ) +
  labs(
    x = "Motif family",
    y = "Number of FIMO peaks"
  ) +
  plot_theme

boxplt_mottyp_simple_number


boxplt_mottyp_simple_wdth <- ggboxplot(
  scatter_df,
  x = "Motif_type_simple",
  y = "Mean_width",
  fill = "Motif_type_simple",
  color = "Motif_type_simple",
  palette = motif_family_colors,
  add = "jitter",
  width = 0.5,
  add.params = list(size = 2, alpha = 0.8)
) +
  stat_compare_means(
    method = "kruskal.test",
    label.y = max(scatter_df$Mean_width, na.rm = TRUE)
  ) +
  labs(
    x = "Motif family",
    y = "Motif width (bp)"
  ) +
  plot_theme

boxplt_mottyp_simple_wdth


# ============================================================
# Experimental data source
# ============================================================

data_source_colors <- setNames(
  grDevices::gray.colors(
    n = length(unique(scatter_df$Data_type_source)),
    start = 0.35,
    end = 0.80
  ),
  unique(scatter_df$Data_type_source)
)

boxplt_dattyp_number <- ggboxplot(
  scatter_df,
  x = "Data_type_source",
  y = "N_amount",
  fill = "Data_type_source",
  color = "Data_type_source",
  palette = data_source_colors,
  add = "jitter",
  width = 0.5,
  add.params = list(size = 2, alpha = 0.8)
) +
  stat_compare_means(
    method = "kruskal.test",
    label.y = max(scatter_df$N_amount, na.rm = TRUE)
  ) +
  labs(
    x = "Data type",
    y = "Number of FIMO peaks"
  ) +
  plot_theme +
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    )
  )

boxplt_dattyp_number


boxplt_dattyp_wdth <- ggboxplot(
  scatter_df,
  x = "Data_type_source",
  y = "Mean_width",
  fill = "Data_type_source",
  color = "Data_type_source",
  palette = data_source_colors,
  add = "jitter",
  width = 0.5,
  add.params = list(size = 2, alpha = 0.8)
) +
  stat_compare_means(
    method = "kruskal.test",
    label.y = max(scatter_df$Mean_width, na.rm = TRUE)
  ) +
  labs(
    x = "Data type",
    y = "Motif width (bp)"
  ) +
  plot_theme +
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    )
  )

boxplt_dattyp_wdth


# ============================================================
# Detailed motif type
# ============================================================

boxplt_mottyp_number <- ggboxplot(
  scatter_df,
  x = "Motif_type",
  y = "N_amount",
  fill = "Motif_type",
  color = "Motif_type",
  palette = motif_type_colors,
  add = "jitter",
  width = 0.5,
  add.params = list(size = 2, alpha = 0.8)
) +
  stat_compare_means(
    method = "kruskal.test",
    label.y = max(scatter_df$N_amount, na.rm = TRUE)
  ) +
  labs(
    x = "Motif type",
    y = "Number of FIMO peaks"
  ) +
  plot_theme +
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    )
  )

boxplt_mottyp_number


boxplt_mottyp_wdth <- ggboxplot(
  scatter_df,
  x = "Motif_type",
  y = "Mean_width",
  fill = "Motif_type",
  color = "Motif_type",
  palette = motif_type_colors,
  add = "jitter",
  width = 0.5,
  add.params = list(size = 2, alpha = 0.8)
) +
  stat_compare_means(
    method = "kruskal.test",
    label.y = max(scatter_df$Mean_width, na.rm = TRUE)
  ) +
  labs(
    x = "Motif type",
    y = "Motif width (bp)"
  ) +
  plot_theme +
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    )
  )

boxplt_mottyp_wdth
```


# Saveblock boxplots

```{r boxplot-save, include=FALSE}
combined_plot_r1 <- 
  boxplt_db_number + boxplt_db_wdth + 
  boxplt_mottyp_simple_number + boxplt_mottyp_simple_wdth + 
  plot_layout(ncol = 4)

ggsave("data/PPAR_RXR_motifs_Hocomoco_Jaspar/FIMO_results_no_max/Objective1_analysis/Objective1_results/Statistics_plots/even_boxplot_panel.pdf", combined_plot_r1, width = 16, height = 4)


combined_plot_r2 <- 
  boxplt_dattyp_number + boxplt_dattyp_wdth +
  plot_layout(ncol = 2)

ggsave("data/PPAR_RXR_motifs_Hocomoco_Jaspar/FIMO_results_no_max/Objective1_analysis/Objective1_results/Statistics_plots/even_boxplot_panel_2.pdf", combined_plot_r2, width = 16, height = 4)


combined_plot_r3 <- 
  boxplt_mottyp_number + boxplt_mottyp_wdth +
  plot_layout(ncol = 2)

ggsave("data/PPAR_RXR_motifs_Hocomoco_Jaspar/FIMO_results_no_max/Objective1_analysis/Objective1_results/Statistics_plots/even_boxplot_panel_3.pdf", combined_plot_r3, width = 16, height = 4)



pdf("data/PPAR_RXR_motifs_Hocomoco_Jaspar/FIMO_results_no_max/Objective1_analysis/Objective1_results/Statistics_plots/FIMO_scat_plts.pdf", width = 8, height = 3, useDingbats = FALSE)  
print(scatter_df_wdth_amount)
dev.off()


pdf("data/PPAR_RXR_motifs_Hocomoco_Jaspar/FIMO_results_no_max/Objective1_analysis/Objective1_results/Statistics_plots/FIMO_db_plts.pdf", width = 1.5 * length(unique(scatter_df$Motif_db)), height = 3)
print(boxplt_db_number)
print(boxplt_db_wdth)
dev.off()

pdf("data/PPAR_RXR_motifs_Hocomoco_Jaspar/FIMO_results_no_max/Objective1_analysis/Objective1_results/Statistics_plots/FIMO_mottyp_smp_plts.pdf", width = 1.5* length(unique(scatter_df$Motif_type_simple)), height = 3)
print(boxplt_mottyp_simple_number)
print(boxplt_mottyp_simple_wdth)
dev.off()

pdf("data/PPAR_RXR_motifs_Hocomoco_Jaspar/FIMO_results_no_max/Objective1_analysis/Objective1_results/Statistics_plots/FIMO_dattyp_plts.pdf", width = 1 * length(unique(scatter_df$Data_type_source)), height = 3)
print(boxplt_dattyp_number)
print(boxplt_dattyp_wdth)
dev.off()

pdf("data/PPAR_RXR_motifs_Hocomoco_Jaspar/FIMO_results_no_max/Objective1_analysis/Objective1_results/Statistics_plots/FIMO_mottyp_plts.pdf", width = 1 * length(unique(scatter_df$Motif_type)), height = 3)
print(boxplt_mottyp_number)
print(boxplt_mottyp_wdth)
dev.off()
```

# Genome-Wide PPRE Score and BED File Export

A distance-weighted PPRE score is calculated for every ChIPseeker-annotated gene, and the results are written out as BED files.

```{r final-genelist-nonzero}
chip_scored <- chip_all %>%
  dplyr::mutate(ppar_score = score * exp(-(abs(distanceToTSS)) / 3000)) # decay constant of 3000 bp reflects PPAR's promoter-proximal binding preference; also used in CistromeGO


GTEX_snRNA_cardiomyocyte_nonzero <- read.csv("data/GTEX_single_nuclear_data/GTEx_non_zeros_ensembl.csv")

GTEX_snRNA_cardiomyocyte_nonzero %>%
  left_join(chip_scored, by = c( "ensembl_gene_id" ="ENSEMBL"), keep = TRUE) %>%
  drop_na(c(ensembl_gene_id,ENSEMBL)) -> motif_hits_cardiomyocytes_nonzero_scored

motif_presence <- motif_hits_cardiomyocytes_nonzero_scored %>%
  distinct(ENSEMBL, motif_id) %>%          # one row per (gene, motif) combination present
  mutate(hit = 1L) %>%                     # presence of the row indicates a hit
  pivot_wider(
    names_from  = motif_id,
    values_from = hit,
    values_fill = list(hit = 0L),
    names_prefix = "motif_"
  )

ranked_scores_nonzero <- motif_hits_cardiomyocytes_nonzero_scored %>%
  group_by(ENSEMBL) %>%
  dplyr::summarise(PPRE_score_sum = sum(ppar_score, na.rm = TRUE),
            PPRE_score_max = max(ppar_score, na.rm = TRUE),
            PPRE_score_median = median(ppar_score, na.rm = TRUE),
            PPRE_score_average = mean(ppar_score, na.rm = TRUE),
            chromStart = dplyr::first(as.integer(start)),
            chromEnd   = dplyr::first(end),
            Chr       = dplyr::first(seqnames),
            SYMBOL    = dplyr::first(SYMBOL),
            ensembl   = dplyr::first(ENSEMBL),
            score     = dplyr::first(PPRE_score_median),
            strand    = dplyr::first(strand),
            geneStart = dplyr::first(geneStart),
            geneEnd   = dplyr::first(geneEnd), 
            name      = dplyr::first(motif_id),
            distanceToTss = dplyr::first(distanceToTSS),
            expression= dplyr::first(expression_value)) %>%
  dplyr::mutate(across(.cols = starts_with("PPRE_score_"),
    .fns = ~ rank(-., ties.method = "first"),  # minus sign for descending ranking
    .names = "rank_{.col}")) #the name for the col is rank_ + colname 

ranked_scores_nonzero_with_motifs <- ranked_scores_nonzero %>%
  left_join(motif_presence, by = "ENSEMBL")
```

```{r bedfile-all-chip-genes}
chip_scored_single_gens <- chip_scored %>%
  group_by(ENSEMBL) %>%
  dplyr::summarise(PPRE_score_sum = sum(ppar_score, na.rm = TRUE),
            PPRE_score_max = max(ppar_score, na.rm = TRUE),
            PPRE_score_median = median(ppar_score, na.rm = TRUE),
            PPRE_score_average = mean(ppar_score, na.rm = TRUE),
            chromStart = dplyr::first(as.integer(start)),
            chromEnd   = dplyr::first(as.integer(end)),
            Chr       = dplyr::first(seqnames),
            SYMBOL    = dplyr::first(SYMBOL),
            ensembl   = dplyr::first(ENSEMBL),
            score     = dplyr::first(PPRE_score_median),
            strand    = dplyr::first(strand),
            geneStart = dplyr::first(geneStart),
            geneEnd   = dplyr::first(geneEnd), 
            name      = dplyr::first(motif_id),
            distanceToTSS = dplyr::first(distanceToTSS)) %>%
  dplyr::mutate(across(.cols = starts_with("PPRE_score_"),
    .fns = ~ rank(-., ties.method = "first"),  # minus sign for descending ranking
    .names = "rank_{.col}")) #the name for the col is rank_ + colname 


chip_scored_single_gens_bed <- data.frame(
  chrom      = chip_scored_single_gens$Chr,
  chromStart = chip_scored_single_gens$chromStart - 1,   # BED is 0-based
  chromEnd   = chip_scored_single_gens$chromEnd,
  motifName       = if ("name"  %in% names(chip_scored_single_gens)) chip_scored_single_gens$name  else ".",
  score      = if ("score" %in% names(chip_scored_single_gens)) chip_scored_single_gens$PPRE_score_median else NA_character_,
  strand     = as.character(chip_scored_single_gens$strand),
  ensembl    = if ("ensembl" %in% names(chip_scored_single_gens))
                 chip_scored_single_gens$ensembl else NA_character_,
  symbol     = if ("SYMBOL" %in% names(chip_scored_single_gens)) {
                 chip_scored_single_gens$SYMBOL
               } else NA_character_,
  geneStart  = if ("geneStart" %in% names(chip_scored_single_gens))
                 chip_scored_single_gens$geneStart else NA_integer_,
  geneEnd    = if ("geneEnd" %in% names(chip_scored_single_gens))
                 chip_scored_single_gens$geneEnd else NA_integer_,
  distanceToTSS = chip_scored_single_gens$distanceToTSS,
  PPRE_score_sum = chip_scored_single_gens$PPRE_score_sum,
  PPRE_score_max = chip_scored_single_gens$PPRE_score_max,
  PPRE_score_average = chip_scored_single_gens$PPRE_score_average,
  ensembl_genes = chip_scored_single_gens$ensembl,
  check.names = FALSE
)


write.table(chip_scored_single_gens_bed, "data/PPAR_RXR_motifs_Hocomoco_Jaspar/FIMO_results_no_max/Objective1_analysis/Objective1_results/Bedfiles_output/PPRE_CHIP_all_genes_test.bed",
            sep = "\t", quote = FALSE,
            row.names = FALSE, col.names = TRUE)
```
```{r create-BEDfile-PPRE_cardiomyocyte_expressed}
ranked_scores_nonzero <- ranked_scores_nonzero_with_motifs
ranked_scores_nonzero_bed <- data.frame(
  chrom      = ranked_scores_nonzero$Chr,
  chromStart = ranked_scores_nonzero$chromStart - 1,   # BED is 0-based
  chromEnd   = ranked_scores_nonzero$chromEnd,
  motifName       = if ("name"  %in% names(ranked_scores_nonzero)) ranked_scores_nonzero$name  else ".",
  score      = if ("score" %in% names(ranked_scores_nonzero)) ranked_scores_nonzero$PPRE_score_median else 0,
  strand     = as.character(ranked_scores_nonzero$strand),
  ensembl    = if ("ensembl" %in% names(ranked_scores_nonzero))
                 ranked_scores_nonzero$ensembl else NA_character_,
  symbol     = if ("SYMBOL" %in% names(ranked_scores_nonzero)) {
                 ranked_scores_nonzero$SYMBOL
               } else NA_character_,
  geneStart  = if ("geneStart" %in% names(ranked_scores_nonzero))
                 ranked_scores_nonzero$geneStart else NA_integer_,
  geneEnd    = if ("geneEnd" %in% names(ranked_scores_nonzero))
                 ranked_scores_nonzero$geneEnd else NA_integer_,
  distancetoTSS = ranked_scores_nonzero$distanceToTss,#
  PPRE_score_sum = ranked_scores_nonzero$PPRE_score_sum,
  PPRE_score_max = ranked_scores_nonzero$PPRE_score_max,
  PPRE_score_average = ranked_scores_nonzero$PPRE_score_average,
  GTEx_expression = ranked_scores_nonzero$expression,
  ensembl_genes = ranked_scores_nonzero$ensembl,
  check.names = FALSE
)

motif_cols <- grep("^motif_", names(ranked_scores_nonzero_with_motifs), value = TRUE)

ranked_scores_nonzero_bed <- cbind(
  ranked_scores_nonzero_bed,
  ranked_scores_nonzero[, motif_cols, drop = FALSE]
)

write.table(ranked_scores_nonzero_bed, "data/PPAR_RXR_motifs_Hocomoco_Jaspar/FIMO_results_no_max/Objective1_analysis/Objective1_results/Bedfiles_output/PPRE_gtex_ranked_scores_abs_3k_test.bed",
            sep = "\t", quote = FALSE,
            row.names = FALSE, col.names = TRUE)
```

```{r final-genelist-top20}
GTEX_snRNA_cardiomyocyte_top20 <- read.csv("data/GTEX_single_nuclear_data/GTEx_top20%_ensembl.csv" )

GTEX_snRNA_cardiomyocyte_top20 %>%
  left_join(chip_scored, by = c( "ensembl_gene_id" ="ENSEMBL"), keep = TRUE) %>%
  drop_na(c(ensembl_gene_id,ENSEMBL)) -> motif_hits_cardiomyocytes_top20_scored

ranked_scores_top20 <- motif_hits_cardiomyocytes_top20_scored %>%
  group_by(ENSEMBL) %>%
  dplyr::summarise(PPRE_score_sum = sum(ppar_score, na.rm = TRUE),
            PPRE_score_max = max(ppar_score, na.rm = TRUE),
            PPRE_score_median = median(ppar_score, na.rm = TRUE),
            PPRE_score_average = mean(ppar_score, na.rm = TRUE), .groups = "drop") %>%
  dplyr::mutate(across(.cols = starts_with("PPRE_score_"),
    .fns = ~ rank(-., ties.method = "first"),  # minus sign for descending ranking
    .names = "rank_{.col}")) %>% #the name for the col is rank_ + colname
  left_join(GTEX_snRNA_cardiomyocyte_nonzero, by = c("ENSEMBL" = "ensembl_gene_id"))
# Note: the source script's trailing `dplyr::select(!X)` here is dropped -- neither
# GTEx_top20%_ensembl.csv nor GTEx_non_zeros_ensembl.csv has ever had a stray "X" column
# (checked both directly), so that select() always errors with "Column `X` doesn't exist"
# whenever this chunk is actually executed. Pre-existing issue in the source
# Objective1_chipseeker_fimo_analysis_cleaned.Rmd, unrelated to the CIS-BP additions.
```



# Distributions of topX genes 

with different aggregation methods per gene (Sum, Max, Median, Average)

```{r Distribution-of-counts-SUM_MAX_MEDIAN_AVERAGE}
scoring_list <- list(
  nonzero = ranked_scores_nonzero
)

genes_to_annotate <- c("CPT1A","HADHA","HADHB","ACADVL","PPARA", "PPARB", "PPARG", "RXRA", "RXRB", "RXRG" )
plot_list <- list()

for (i in seq_along(scoring_list)) {
  df <- scoring_list[[i]] 
  ds_name <- names(scoring_list)[i]

  gene_candidates <- c("SYMBOL", "HGNC_gene_name", "hgnc_symbol", "gene_symbol")
  gene_col <- intersect(gene_candidates, names(df))[1]
  if (is.na(gene_col))  
    next
  df <- dplyr::rename(df, SYMBOL = dplyr::all_of(gene_col))
  for (col_name in names(df)) {
    
    if (!startsWith(tolower(col_name), "ppre"))
      next

    rank_col <- paste0("rank_", col_name)
    if (!rank_col %in% names(df))
      next

    if (!is.numeric(df[[col_name]])) 
      next

    # values for annotated genes
    gene_values <- df %>%
      dplyr::filter(SYMBOL %in% genes_to_annotate) %>%
      dplyr::select(SYMBOL, value = dplyr::all_of(col_name))

    p <- ggplot2::ggplot(df, ggplot2::aes(x = .data[[col_name]])) +
      ggplot2::geom_histogram(ggplot2::aes(fill = .data[[rank_col]]), bins = 300) +
      ggplot2::scale_fill_manual(values = c("TRUE" = "black", "FALSE" = "red")) +
      ggplot2::labs(title = gsub("_", " ", col_name),
                    x = gsub("_", " ", col_name),
                    y = "Frequency") +
      ggplot2::scale_x_continuous(limits = c(0, NA), expand = c(0, 0)) +
      ggplot2::scale_y_continuous(limits = c(0, NA), expand = c(0, 0)) +
      ggplot2::coord_cartesian(clip = "off") +
      ggplot2::guides(fill = "none") +
      ggplot2::theme_bw() +
      ggplot2::theme(
        panel.border = ggplot2::element_blank(),
        panel.grid.major = ggplot2::element_blank(),
        panel.grid.minor = ggplot2::element_blank(),
        axis.line.x = ggplot2::element_line(color = "black"),
        axis.ticks.x = ggplot2::element_line(color = "black"),
        axis.line.y = ggplot2::element_line(color = "black"),
        legend.title = ggplot2::element_blank()
      )

    # add vertical lines & labels for annotated genes
    if (nrow(gene_values) > 0) {
      p <- p +
        ggplot2::geom_vline(data = gene_values,
                            ggplot2::aes(xintercept = value),
                            color = "blue", linetype = "dashed") +
        ggplot2::geom_text(data = gene_values,
                           ggplot2::aes(x = value, y = Inf, label = SYMBOL),
                           inherit.aes = FALSE, angle = 90, vjust = 1.2, size = 3, color = "blue")
    }

    # store plot with dataset + column name
    plot_list[[paste(ds_name, col_name, sep = "_")]] <- p
  }
}

print(plot_list[["nonzero_PPRE_score_sum"]])
print(plot_list[["nonzero_PPRE_score_max"]])
print(plot_list[["nonzero_PPRE_score_median"]])
print(plot_list[["nonzero_PPRE_score_average"]])
```

```{r save-ppre-histograms}
pdf("data/PPAR_RXR_motifs_Hocomoco_Jaspar/FIMO_results_no_max/Objective1_analysis/Objective1_results/Scoring_FIMO/PPRE_score_histograms.pdf", width = 11, height = 6)  # adjust size if you want
for (nm in names(plot_list)) {
  print(plot_list[[nm]])
}
dev.off()
```
# Over Representation Analysis

Dotplots per different aggregation method

```{r run-ORA-dtplt}
run_ora_dotplots <- function(df, pdf_path, width, height){
  # open a multi-page PDF device
  dir.create(dirname(pdf_path), showWarnings = FALSE, recursive = TRUE)
  grDevices::pdf(pdf_path, onefile = TRUE, width = width, height = height, useDingbats = FALSE) #8.9 for article
  on.exit(try(grDevices::dev.off(), silent = TRUE), add = TRUE)

  cols <- names(df)[startsWith(tolower(names(df)), "ppre")]
  cat("Processing", length(cols), "columns:", paste(cols, collapse = ", "), "\n")
  printed <- 0

  for (col_name in cols) {
    cat("\n--", col_name, "--\n")
    if (!("SYMBOL" %in% names(df))) { cat("No SYMBOL column.\n"); next }

    top_syms <- df %>%
      arrange(desc(.data[[col_name]])) %>%
      slice_head(n = 5000) %>%
      pull(SYMBOL) %>%
      unique() %>% na.omit()

    cat("Top symbols:", length(top_syms), "\n")
    if (length(top_syms) < 10) { cat("Too few symbols.\n"); next }

    gene_map <- bitr(top_syms, fromType="SYMBOL", toType="ENTREZID", OrgDb=org.Hs.eg.db)
    cat("Mapped to ENTREZ:", nrow(gene_map), "\n")
    if (nrow(gene_map) < 10) { cat("Too few mapped; skipping.\n"); next }

    ego <- suppressMessages(enrichGO(
      gene=gene_map$ENTREZID, OrgDb=org.Hs.eg.db, ont="BP",
      pAdjustMethod="BH", pvalueCutoff=0.05, qvalueCutoff=0.05,
      minGSSize=10, maxGSSize=5000, readable=TRUE))

    n_terms <- nrow(as.data.frame(ego))
    cat("Significant terms:", n_terms, "\n")
    if (n_terms == 0) { cat("No significant enrichment under thresholds.\n"); next }

    p <- dotplot(ego, showCategory=20, x="GeneRatio", orderBy="x",
                 title=gsub("_", " ", col_name))
    print(p)  # each print adds a new page to the PDF
    printed <- printed + 1
  }

  if (printed == 0) {
    grid::grid.newpage()
    grid::grid.text("No ORA plots produced.\nCheck mapping/thresholds/top_n.", x=.5, y=.6)
  }
  cat("\nDone. Printed", printed, "plots to", pdf_path, "\n")
}

outdir <- "data/PPAR_RXR_motifs_Hocomoco_Jaspar/FIMO_results_no_max/Objective1_analysis/Objective1_results"
run_ora_dotplots(ranked_scores_top20,  file.path(outdir, "ORA_top20.pdf"), 8, 9)
run_ora_dotplots(ranked_scores_nonzero,file.path(outdir, "ORA_nonzero.pdf"), 8,9)
run_ora_dotplots(ranked_scores_nonzero,file.path(outdir, "ORA_nonzero_graphical abstract.pdf"), 5,5)
```
