---
title: "FIMO Motif Analysis with ChIPseeker"
author: "Eli Brouwer"
date: "`r Sys.Date()`"
output: 
  html_document
---

This file is used to analyse the FIMO files, for objective 1. In this File the FIMO files are annotated with ChipSeekr. And overlapped with GTEx, to create the final genelist.

```{r setup, include=FALSE}
#load in libraries that are used for this file. 
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(pacman)
p_load(
  tidyr,
  ChIPseeker,
  dplyr,
  org.Hs.eg.db,
  TxDb.Hsapiens.UCSC.hg38.knownGene,
  gt,
  writexl,
  pheatmap,
  stringr,
  GenomicRanges,
  BSgenome.Hsapiens.UCSC.hg38,
  ggvenn,
  magrittr,
  rtracklayer,
  clusterProfiler,
  tibble,
  ggpubr,
  ggrepel,
  forcats,
  patchwork,
  readxl,
  ggplot2,
  IRanges
)
```

# Data Input

```{r load-data}
# Set file paths
fimo_file_directory = "data/PPAR_RXR_motifs_Hocomoco_Jaspar/FIMO"
fimo_file_lst <- list.files(fimo_file_directory)
motif_annotation_path <- "data/annot_file.txt"
chip_file_directory <- "data/ChIPseeker_output"
snrna_file_path <- "data/GTEx_non_zeros_prt_coding.csv"
```

# Annotate Motifs with ChIPseeker

The motifs are loaded in in a GRanges object.

```{r annotate-fimo, verbose = FALSE, message = FALSE}
# Function to process and annotate FIMO files
fimo_analyzer <- function(fimo_file_path, txdb){
  fimo_file <- read.delim(fimo_file_path) %>% filter(!is.na(start))
  gr_fimo <- GRanges(
    seqnames = fimo_file$sequence_name,
    ranges = IRanges(start = fimo_file$start, end = fimo_file$stop),
    strand = fimo_file$strand,
    score = fimo_file$score,
    pvalue = fimo_file$p.value,
    qvalue = fimo_file$q.value,
    motif_id = fimo_file$motif_id, 
    motif_seq = fimo_file$matched_sequence
  )
  seqlevelsStyle(gr_fimo) <- "UCSC"
  gr_fimo_splt <- split(gr_fimo, mcols(gr_fimo)$motif_id)
  summary_list <- list()
  gr_list <- list()


  for (i in seq_along(gr_fimo_splt)){
    gr_current <- gr_fimo_splt[[i]]
    gr_list[[length(gr_list) + 1]] <- gr_current
    #ChIPseeker, specify the search window 3kb up and 1kb down, sameStrand = TRUE
    peaks <- annotatePeak(gr_current, tssRegion = c(-3000, 1000), TxDb = txdb, 
                          annoDb = "org.Hs.eg.db", sameStrand = TRUE, verbose = FALSE)
    peaks_df <- as.data.frame(peaks)
    nearby_peaks <- peaks_df[grepl("Promoter", peaks_df$annotation), ] #only keep the genes that are in the promoter
    
    #Generate Fimo and chipseeker summary results, per motif
    summary_list[[length(summary_list) + 1]] <- data.frame(
      N_peaks = length(gr_current),
      Min_pvalue = min(gr_current$pvalue),
      Mean_width = mean(width(gr_current)),
      Motif_id = gr_current$motif_id[1],
      Motif_seq = gr_current$motif_seq[1],
      N_genes = length(nearby_peaks$annotation),
      Mean_dist_to_TSS = mean(nearby_peaks$distanceToTSS))
  }
  return(list(summary = summary_list, granges = gr_list))
}

# Run annotation

#A list of FIMO files is created and the summary list is stored (which is used for Table S2 and the granges objects are stored in a list

motif_summary_list <- list()
granges_list <- list()
for (i in 1:length(fimo_file_lst)){
  print(paste0("Fimo file: ", i, " / Out of 6"))
  this_list <- fimo_analyzer(paste0(fimo_file_directory, "/", fimo_file_lst[i]), TxDb.Hsapiens.UCSC.hg38.knownGene)
  motif_summary_list <- c(motif_summary_list, this_list$summary)
  granges_list <- c(granges_list, this_list$granges)
}

granges_fimo <-  do.call(c, granges_list)

print("The results are safed in the granges list")
```

This code is used to create a bedfile with all the seperate FIMO hits

```{r bedfile-fimo, verbose = FALSE, message = FALSE,warning=FALSE}
granges_all <- GRanges()  # empty GRanges
for (i in seq_along(fimo_file_lst)) {
  this_list <- fimo_analyzer(paste0(fimo_file_directory, "/", fimo_file_lst[i]),
                             TxDb.Hsapiens.UCSC.hg38.knownGene)
  granges_all <- c(granges_all, do.call(c, this_list$granges))
}

df <- as.data.frame(granges_all)

PPRE_FIMO_bed <- data.frame(
  chrom      = df$seqnames,
  chromStart = df$start - 1,   
  chromEnd   = df$end,
  name       = if ("motif_id"  %in% names(df)) df$motif_id  else ".",
  score      = if ("score" %in% names(df)) df$score else 0,
  strand     = as.character(df$strand),
  check.names = FALSE
) #makes a bedfile of all FIMO hits. 

write.table(PPRE_FIMO_bed, "data/PPAR_RXR_motifs_Hocomoco_Jaspar/Objective1_results/Bedfiles_output/PPRE_FIMO_hits_21_motifs.bed",
            sep = "\t", quote = FALSE,
            row.names = FALSE, col.names = TRUE)
```

# Motif Overlap Heatmap

the overlap between two pairs of motifs is plotted in a heatmap

```{r overlap-heatmap}
#this function calculates the overlap
get_overlap_percent <- function(gr1, gr2, minoverlap) {
  hits <- findOverlaps(gr1, gr2, minoverlap = minoverlap)
  unique_queries <- length(unique(queryHits(hits)))
  
  # % of gr1 that overlaps gr2
  percent <- (unique_queries / length(gr1)) * 100
  return(percent)
}

n <- length(granges_list)

set_names <- names(granges_list)

motif_names <- sapply(granges_list, function(gr) as.character(mcols(gr)$motif_id[1]))

names(granges_list) <- motif_names

overlap_matrix <- matrix(NA, nrow = n, ncol = n, dimnames = list(motif_names, motif_names))
for (i in 1:n) {
  for (j in 1:n) {
    if (i != j) {
      overlap_matrix[i, j] <- get_overlap_percent(granges_list[[i]], granges_list[[j]], minoverlap = 5)
    } else {
      overlap_matrix[i, j] <- 100 #when the pair of the motif combinations is the same, a value of 100 is given
    }
  }
}

overlap_df <- as.data.frame(overlap_matrix)
overlap_df <- cbind(Motif = rownames(overlap_df), overlap_df)

motif_class <- c(
  "MA1148.1" = "Full-PPARA::RXRA","MA1148.2" = "Full-PPARA::RXRA", "MA0065.1" = "Full-PPARG::RXRA",
  "PPARA.H13CORE.0.P.B" = "Half-PPARA", "PPARA.H13CORE.1.P.B" = "Half-PPARA", "PPARD.H13CORE.0.PSM.A" = "Half-PPARD", 
  "MA1550.1" = "Half-PPARD","MA1550.2" = "Half-PPARD", "PPARG.H13CORE.0.P.B" = "Half-PPARG", 
  "PPARG.H13CORE.1.P.B" = "Half-PPARG", "MA0066.1" = "Half-PPARG", "MA0066.2" = "Half-PPARG",
  "RXRA.H13CORE.0.PS.A" = "Half-RXRA", "RXRA.H13CORE.1.S.C" = "Half-RXRA", "RXRA.H13CORE.2.SM.B" = "Half-RXRA", 
  "RXRA.H13CORE.3.P.B" = "Half-RXRA", "RXRB.H13CORE.0.P.C" = "Half-RXRB", "RXRB.H13CORE.1.SM.B" = "Half-RXRB",
  "MA0855.1" = "Half-RXRB", "MA1555.1" = "Half-RXRB", "MA0856.1" = "Half-RXRG", "MA1556.1" = "Half-RXRG") #adding a annotation col of motif-type 

motif_groups <- data.frame(Motif = rownames(overlap_df))
motif_groups$Motif_type <- motif_class[motif_groups$Motif] #create a annotation list
rownames(motif_groups) <- motif_groups$Motif
motif_groups %<>%
  dplyr::mutate(Motif_db = case_when(
    grepl("H13CORE", Motif, ignore.case = TRUE) ~ "HOCOMOCO",
    grepl("^MA", Motif, ignore.case = TRUE) ~ "JASPAR",
    TRUE ~ "other")) #adds an annotation row, with motif db, jaspar and hocomoco

rownames(overlap_df) <- overlap_df$Motif
mat <- as.matrix(overlap_df[, -1])
mat[is.na(mat)] <- 100 #gives the overlapping pairs an overlap of 100%

pheatmap(round(overlap_matrix, 1), cluster_rows = TRUE, cluster_cols = TRUE)
```

```{r motif- heatmap saveblock}
pdf("data/PPAR_RXR_motifs_Hocomoco_Jaspar/Objective1_results/correlation_heatmap.pdf", height =11.7 , width = 12.3)

pheatmap(round(mat, 1), cluster_rows = TRUE, cluster_cols = TRUE, 
         display_numbers = FALSE, fontsize = 10, annotation_row = motif_groups["Motif_type"], annotation_col =  motif_groups["Motif_db"])
dev.off()

write_xlsx(overlap_df, path = "data/PPAR_RXR_motifs_Hocomoco_Jaspar/Objective1_results/pairwise_comparison_fimo.xlsx")
```

# Summary Table of FIMO and ChipSeekr results

```{r summary-table}
motif_summary_df <- do.call(rbind, motif_summary_list) 
colnames(motif_summary_df) <- c("N_amount", "Min_pvalue", "Mean_width", "Motif_name", "Sequence", "N_amount_genes", "Mean_dist_tss")
rownames(motif_summary_df) <- motif_summary_df$`Motif name`

#Combined table with both chipseeker and FIMO results summary
FIMO_chipseeker_results <- motif_summary_df %>%
  left_join(motif_groups, by = c("Motif_name" = "Motif")) %>%
  mutate(Motif_gene = str_remove(Motif_type, "^(Half-|Full-)")) %>% 
  dplyr::select(Motif_gene, Motif_name, Motif_name, N_amount, N_amount_genes, Mean_dist_tss, Mean_width) %>%
  gt(rowname_col = "Motif_name") %>%
  cols_label(
    Motif_gene = "Motif gene",
    N_amount = "Peaks",
    Mean_width = "Mean Width",
    N_amount_genes = "Amount of genes",
    Mean_dist_tss = "Average to TSS"
  ) %>%
  grand_summary_rows(
    columns = c(N_amount, Mean_width,N_amount_genes, Mean_dist_tss  ),
    fns = list(
      "Mean (SD)" = ~ sprintf("%.1f (%.1f)", mean(., na.rm = TRUE), sd(., na.rm = TRUE))
      )
  ) %>%
  fmt_number(columns = c("N_amount", "Mean_width", "N_amount_genes", "Mean_dist_tss"), decimals = 0)
FIMO_chipseeker_results 

gtsave(FIMO_chipseeker_results, "data/PPAR_RXR_motifs_Hocomoco_Jaspar/Objective1_results/motif_results_table.pdf", expand = 2)

```

```{r write table to XLSX}
display_df_num <-
  motif_summary_df %>%
  left_join(motif_groups, by = c("Motif_name" = "Motif")) %>%
  mutate(Motif_gene = str_remove(Motif_type, "^(Half-|Full-)")) %>%
  transmute(
    `Motif name`      = Motif_name,
    `Motif gene`      = Motif_gene,
    `Peaks`           = as.numeric(N_amount),
    `Amount of genes` = as.numeric(N_amount_genes),
    `Average to TSS`  = as.numeric(Mean_dist_tss),
    `Mean Width`      = as.numeric(Mean_width)
  )

# Compute mean(SD) on numeric columns
ms <- function(x) sprintf("%.1f (%.1f)", mean(x, na.rm = TRUE), sd(x, na.rm = TRUE))
sum_row <- tibble(
  `Motif name`      = "Mean (SD)",
  `Motif gene`      = "\u2014",
  `Peaks`           = ms(display_df_num$`Peaks`),
  `Amount of genes` = ms(display_df_num$`Amount of genes`),
  `Average to TSS`  = ms(display_df_num$`Average to TSS`),
  `Mean Width`      = ms(display_df_num$`Mean Width`)
)

fmt_int <- function(x) formatC(x, format = "d", big.mark = ",")
fmt_1   <- function(x) formatC(x, format = "f", digits = 1)

display_df_chr <- display_df_num %>%
  mutate(
    `Peaks`           = fmt_int(`Peaks`),
    `Amount of genes` = fmt_int(`Amount of genes`),
    `Average to TSS`  = paste0(ifelse(`Average to TSS` < 0, "\u2212", ""), fmt_1(abs(`Average to TSS`))),
    `Mean Width`      = fmt_1(`Mean Width`)
  )


display_df_sum <- bind_rows(display_df_chr, sum_row)

write_xlsx(list("PPRE table" = display_df_sum),
           path = "data/PPAR_RXR_motifs_Hocomoco_Jaspar/Objective1_results/motif_results_table.xlsx")

```

# Expression Overlap with GTEx Cardiomyocyte Genes

```{r gene-overlap}
#Create total total genelist
chip_file_directory = "data/PPAR_RXR_motifs_Hocomoco_Jaspar/ChIPseeker_output"
chip_file_list <- list.files(chip_file_directory)
chip_file_list_txt <- chip_file_list[str_ends(chip_file_list, ".txt")]

#for all motifs
chip_data_list <- list()

for (i in 1:length(chip_file_list_txt)){
  chiptable <- read.delim(paste0(chip_file_directory, "/", chip_file_list_txt[i]), sep = " ", header = TRUE)
  chip_data_list[[i]] <- chiptable
}

chip_all <- do.call(rbind, chip_data_list) #this is a dataframe with all the genes in there

overlapping_genes <- chip_all %>%
  dplyr::select(c(seqnames, motif_id, annotation, 
                  geneId, transcriptId, ENSEMBL, 
                  SYMBOL,  pvalue , qvalue, distanceToTSS )) %>%
  distinct(transcriptId, motif_id, .keep_all = TRUE) %>%
  add_count(transcriptId, name = "n_motifs") %>%
  distinct(ENSEMBL, .keep_all = TRUE) %>%
  dplyr::arrange(desc(n_motifs))
```

```{r add-GTEX-nozero}
##add snRNA seq data to overlap the expressed cardiomyocyte genes
GTEX_snRNA_cardiomyocyte_nonzero <- read.csv("data/GTEX_single_nuclear_data/GTEx_non_zeros_ensembl.csv") %>%
  left_join(overlapping_genes, by = c( "ensembl_gene_id" ="ENSEMBL"), keep = TRUE) %>%
  drop_na(c(ensembl_gene_id,ENSEMBL)) -> motif_hits_cardiomyocytes_nonzero 


write.csv(motif_hits_cardiomyocytes_nonzero, "data/GTEX_single_nuclear_data/Motifs_GTEX_cm_expressed.csv")
```

```{r add-GTEX-top20%}
##add snRNA seq data to overlap the expressed cardiomyocyte genes
read.csv("data/GTEX_single_nuclear_data/GTEx_top20%_ensembl.csv") %>%
  left_join(overlapping_genes, by = c( "ensembl_gene_id" ="ENSEMBL"), keep = TRUE) %>%
  drop_na(c(ensembl_gene_id,ENSEMBL)) -> motif_hits_cardiomyocytes_top20


write.csv(motif_hits_cardiomyocytes_top20, "data/GTEX_single_nuclear_data/Motifs_GTEX_cm_expressed_top20.csv")
```

# Venn Diagram of Gene Sets

```{r venn-diagram}
#venn_diagram
expression_cardiomyocytes_nozero <- read.csv( "data/GTEX_single_nuclear_data/GTEx_non_zeros_ensembl.csv" )
GTEX_genes <- read.csv("data/GTEX_single_nuclear_data/GTEx_8_genes.csv")
venn_genelist <- list( 'GTEX genes' = unique(GTEX_genes$gene_ids), 'Expressed cardiomyocyte genes' = sort(unique(na.omit(expression_cardiomyocytes_nozero$ensembl_gene_id))), 'All Motif PPRE genes' = sort((unique(chip_all$ENSEMBL))))

venn_diagram <- ggvenn(
  venn_genelist,
  c("GTEX genes" ,"Expressed cardiomyocyte genes", "All Motif PPRE genes" ),
  show_percentage = FALSE,
  fill_color = c("white", "white", "white")
) 
venn_diagram

pdf("data/PPAR_RXR_motifs_Hocomoco_Jaspar/Objective1_results/venn_diagram.pdf", width = 12.5)
venn_diagram
dev.off()
```

# Get the overlap between HCM and PLN regions and FIMO peaks

```{r differentially-overlap-regions}
# Load differential region files and make GRanges object from it

HCM_differentially_regions <- read_xlsx(
  "data/PPAR_RXR_motifs_Hocomoco_Jaspar/HCM_differentially_regions.xlsx",
  sheet = "Supplementary Table 2A"
)

grange_HCM_differentially_regions <- GRanges(
  seqnames = HCM_differentially_regions$`Chromosome number`,
  ranges = IRanges(
    start = HCM_differentially_regions$`Region start`,
    end   = HCM_differentially_regions$`Region end`
  ),
  pvalue = HCM_differentially_regions$`P-adj`,
  motif_id = "HCM_hearts"
)

PLN_differentially_regions <- read_xlsx(
  "data/PPAR_RXR_motifs_Hocomoco_Jaspar/PLN_differentially_regions.xlsx",
  skip = 1
) %>%
  filter(!is.na(`Region start†`))

grange_PLN_differentially_regions <- GRanges(
  seqnames = PLN_differentially_regions$Chromosome,
  ranges = IRanges(
    start = PLN_differentially_regions$`Region start†`,
    end   = PLN_differentially_regions$`Region end†`
  ),
  pvalue = PLN_differentially_regions$`Adjusted p-value§`,
  motif_id = "PLN_hearts"
)

# LiftOver from h19 to hg38


chain <- import.chain(
  "data/PPAR_RXR_motifs_Hocomoco_Jaspar/hg19ToHg38.over.chain"
)

grange_HCM_differentially_regions_hg38 <- liftOver(grange_HCM_differentially_regions, chain) %>%
  unlist()

grange_PLN_differentially_regions_hg38 <- liftOver(grange_PLN_differentially_regions, chain) %>%
  unlist()

# Count overlap per motif

overlap_df_regions <- data.frame()

for (i in seq_along(granges_list)) {
  
  overlap_hcm <- findOverlaps(
    grange_HCM_differentially_regions_hg38,
    granges_list[[i]],
    ignore.strand = TRUE,
    minoverlap = 1
  )
  
  overlap_pln <- findOverlaps(
    grange_PLN_differentially_regions_hg38,
    granges_list[[i]],
    ignore.strand = TRUE,
    minoverlap = 1
  )
  
  overlap_df_regions <- rbind(
    overlap_df_regions,
    data.frame(
      motif_id = names(granges_list)[i],
      overlap_hcm = length(unique(queryHits(overlap_hcm))),
      overlap_pln = length(unique(queryHits(overlap_pln)))
    )
  )
}

# Calculate percentages

total_hcm <- length(grange_HCM_differentially_regions_hg38)
total_pln <- length(grange_PLN_differentially_regions_hg38)

overlap_percent_df <- overlap_df_regions %>%
  mutate(
    percent_overlap_hcm_regions = (overlap_hcm / total_hcm) * 100,
    percent_overlap_pln_regions = (overlap_pln / total_pln) * 100
  )

# Add motif database annotation

overlap_percent_df <- overlap_percent_df %>%
  mutate(
    database = case_when(
      grepl("^MA", motif_id) ~ "JASPAR",
      grepl("H13CORE", motif_id) ~ "HOCOMOCO",
      TRUE ~ "Unknown"
    )
  )

# Convert to long format

overlap_percent_long <- overlap_percent_df %>%
  select(motif_id, database, percent_overlap_hcm_regions, percent_overlap_pln_regions) %>%
  pivot_longer(
    cols = c(percent_overlap_hcm_regions, percent_overlap_pln_regions),
    names_to = "condition",
    values_to = "percent_overlap"
  ) %>%
  mutate(
    condition = recode(
      condition,
      "percent_overlap_hcm_regions" = "HCM regions",
      "percent_overlap_pln_regions" = "PLN regions"
    )
  )


# Order motifs within database from high overlap to lower

motif_order <- overlap_percent_long %>%
  group_by(database, motif_id) %>%
  summarise(mean_overlap = mean(percent_overlap, na.rm = TRUE), .groups = "drop") %>%
  arrange(database, mean_overlap) %>%
  pull(motif_id)

overlap_percent_long <- overlap_percent_long %>%
  mutate(
    motif_id = factor(motif_id, levels = unique(motif_order)),
    condition = factor(condition, levels = c("HCM regions", "PLN regions")),
    database = factor(database, levels = c("JASPAR", "HOCOMOCO"))
  )

# Plot

DatabaseSplitPlot <- ggplot(
  overlap_percent_long,
  aes(x = motif_id, y = percent_overlap, fill = condition)
) +
  geom_col(
    position = position_dodge(width = 0.60),
    width = 0.7
  ) +
  coord_flip() +
  facet_wrap(~ database, ncol = 1, scales = "free_y") +
  scale_fill_manual(
    values = c("HCM regions" = "#3C8DBC", "PLN regions" = "#E69F00")
  ) +
  scale_y_continuous(
  limits = c(0, 25),
  expand = c(0, 0)
) +
facet_wrap(
    ~ database,
    ncol = 1,
    scales = "free_y",
    axes = "all_x",
    axis.labels = "all_x"
  ) +
  theme_bw() +
  theme(
    panel.border = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    axis.line = element_line(colour = "black"),
    strip.background = element_blank(),
    strip.text = element_text(face = "plain", size = 12),
    axis.text.y = element_text(size = 10),
    axis.text.x = element_text(size = 10),
    axis.title = element_text(size = 12),
    plot.title = element_text(size = 12, face = "plain"),
    legend.title = element_blank(),
    legend.position = "top",
    panel.spacing = unit(0.6, "lines")
  ) +
  labs(
    x = "Motif ID",
    y = "Differential regions overlapped (%)"
  )

DatabaseSplitPlot

#Calculate overlap percentages for article
total_hcm <- length(grange_HCM_differentially_regions_hg38)
total_pln <- length(grange_PLN_differentially_regions_hg38)

overlap_percent_df <- overlap_df_regions %>%
  pivot_wider(names_from = name, values_from = value, values_fn = list) %>%
  unnest(cols = c(total_hits, overlap_pln, overlap_hcm)) %>%
  mutate(
    percent_overlap_pln = (overlap_pln / total_pln) * 100,
    percent_overlap_hcm = (overlap_hcm / total_hcm) * 100)

pln_mean <- mean(overlap_percent_df$percent_overlap_pln, na.rm = TRUE)
pln_sd   <- sd(overlap_percent_df$percent_overlap_pln, na.rm = TRUE)

hcm_mean <- mean(overlap_percent_df$percent_overlap_hcm, na.rm = TRUE)
hcm_sd   <- sd(overlap_percent_df$percent_overlap_hcm, na.rm = TRUE)

cat(sprintf("PLN-specific H3K4me3 regions overlapping with the FIMO regions: %.2f%% ± %.2f SD\n", pln_mean, pln_sd)) 
cat(sprintf("HCM-specific H3K27ac regions overlapping with the FIMO regions: %.2f%% ± %.2f SD\n", hcm_mean, hcm_sd))
```
```{r}
library(GenomicRanges)
library(dplyr)
library(tibble)
library(purrr)
diff_regions_list <- list(
  HCM = grange_HCM_differentially_regions_hg38,
  PLN = grange_PLN_differentially_regions_hg38
)

calculate_overlap_metrics <- function(diff_regions, motif_gr, minoverlap = 5) {
  
  diff_regions <- GenomicRanges::reduce(diff_regions)
  motif_gr <- GenomicRanges::reduce(motif_gr)
  
  hits <- GenomicRanges::findOverlaps(
    diff_regions,
    motif_gr,
    minoverlap = minoverlap
  )
  
  n_total_regions <- length(diff_regions)
  n_overlap_regions <- length(unique(queryHits(hits)))
  percent_regions_overlap <- (n_overlap_regions / n_total_regions) * 100
  
  if (length(hits) > 0) {
    overlap_gr <- GenomicRanges::pintersect(
      diff_regions[queryHits(hits)],
      motif_gr[subjectHits(hits)]
    )
    
    overlap_bp <- sum(width(GenomicRanges::reduce(overlap_gr)))
  } else {
    overlap_bp <- 0
  }
  
  total_region_bp <- sum(width(diff_regions))
  percent_bp_overlap <- (overlap_bp / total_region_bp) * 100
  
  tibble::tibble(
    n_total_regions = n_total_regions,
    n_overlap_regions = n_overlap_regions,
    percent_regions_overlap = percent_regions_overlap,
    total_region_bp = total_region_bp,
    overlap_bp = overlap_bp,
    percent_bp_overlap = percent_bp_overlap,
    median_region_width = median(width(diff_regions)),
    mean_region_width = mean(width(diff_regions))
  )
}

overlap_results <- map_dfr(names(diff_regions_list), function(dataset_name) {
  
  diff_regions <- diff_regions_list[[dataset_name]]
  
  map_dfr(names(granges_list), function(motif_name) {
    
    motif_gr <- granges_list[[motif_name]]
    
    calculate_overlap_metrics(
      diff_regions = diff_regions,
      motif_gr = motif_gr,
      minoverlap = 5
    ) %>%
      mutate(
        dataset = dataset_name,
        motif_id = motif_name
      )
  })
})

overlap_results

summary_overlap <- overlap_results %>%
  dplyr::group_by(dataset) %>%
  dplyr::summarise(
    n_total_regions = dplyr::first(n_total_regions),
    median_region_width = dplyr::first(median_region_width),
    mean_region_width = dplyr::first(mean_region_width),
    mean_percent_regions_overlap = mean(percent_regions_overlap),
    sd_percent_regions_overlap = sd(percent_regions_overlap),
    mean_percent_bp_overlap = mean(percent_bp_overlap),
    sd_percent_bp_overlap = sd(percent_bp_overlap),
    .groups = "drop"
  )

summary_overlap
```
```{r}
BiocManager::install("regioneR")
library(regioneR)
library(GenomicRanges)

# Example for one dataset and one motif
observed_overlap <- calculate_overlap_metrics(
  diff_regions = grange_HCM_differentially_regions_hg38,
  motif_gr = granges_list[["MA0856.1"]],
  minoverlap = 5
)$percent_regions_overlap

background <- GenomicRanges::reduce(c(grange_HCM_differentially_regions_hg38, grange_PLN_differentially_regions_hg38))

perm <- permTest(
  A = grange_HCM_differentially_regions_hg38,
  B = granges_list[["MA0856.1"]],
  ntimes = 1000,
  randomize.function = randomizeRegions,
  evaluate.function = numOverlaps,
  genome = "hg38",
  background = background, 
  count.once = TRUE
)

perm
plot(perm)
```

#Statistics boxplots Which gives an idea about the distribution of the peaks and widths for every different motif.

```{r make-scatter_df}
measuring_method <- tibble::tribble(
  ~Motif_name,                         ~Data_type_source,                                 ~Data_type_source_simple,
  "MA1148.1",                          "SMILE-seq",                                       "SMILE-seq",
  "MA1148.2",                          "SMILE-seq",                                       "SMILE-seq",
  "MA0065.1",                          "SELEX",                                           "SELEX",
  "PPARA.H13CORE.0.P.B",              "ChIP-Seq",                                        "ChIP-Seq",
  "PPARA.H13CORE.1.P.B",              "ChIP-Seq",                                        "ChIP-Seq",
  "PPARD.H13CORE.0.PSM.A",            "ChIP-Seq + HT-SELEX + Methyl-HT-SELEX",           "SELEX",
  "MA1550.1",                          "HT-SELEX",                                        "SELEX",
  "MA1550.2",                          "HT-SELEX",                                        "SELEX",
  "PPARG.H13CORE.0.P.B",              "ChIP-Seq",                                        "ChIP-Seq",
  "PPARG.H13CORE.1.P.B",              "ChIP-Seq",                                        "ChIP-Seq",
  "MA0066.1",                          "SELEX",                                           "SELEX",
  "MA0066.2",                          "SELEX",                                           "SELEX",
  "RXRA.H13CORE.0.PS.A",              "ChIP-Seq + HT-SELEX",                             "SELEX",
  "RXRA.H13CORE.1.S.C",               "HT-SELEX",                                        "SELEX",
  "RXRA.H13CORE.2.SM.B",              "HT-SELEX + Methyl-HT-SELEX",                      "SELEX",
  "RXRA.H13CORE.3.P.B",               "ChIP-Seq",                                        "ChIP-Seq",
  "RXRB.H13CORE.0.P.C",               "ChIP-Seq",                                        "ChIP-Seq",
  "RXRB.H13CORE.1.SM.B",              "HT-SELEX + Methyl-HT-SELEX",                      "SELEX",
  "MA0855.1",                          "HT-SELEX",                                        "SELEX",
  "MA1555.1",                          "HT-SELEX",                                        "SELEX",
  "MA0856.1",                          "HT-SELEX",                                        "SELEX",
  "MA1556.1",                          "HT-SELEX",                                        "SELEX"
)
#Make scatterplot mean width of motif vs number of peaks
motif_summary_df %>% 
  dplyr::select(c(Motif_name, Mean_width, N_amount)) %>%
  left_join(motif_groups, by = c("Motif_name" = "Motif")) %>%
  dplyr::mutate(Motif_type_simple = case_when(
    grepl("^Full", Motif_type, ignore.case = TRUE) ~ "Full-PPAR:RXR",
    grepl("RXR", Motif_type, ignore.case = TRUE) ~ "Half-RXR",
    grepl("PPAR", Motif_type, ignore.case = TRUE) ~ "Half-PPAR",
    TRUE ~ "other")) %>%
  dplyr::mutate(Motif_db = case_when(
    grepl("H13CORE", Motif_name, ignore.case = TRUE) ~ "HOCOMOCO",
    grepl("^MA", Motif_name, ignore.case = TRUE) ~ "JASPAR",
    TRUE ~ "other")) %>%
  left_join(measuring_method, by = "Motif_name") %>%
  mutate(
    Data_type_source = factor(Data_type_source, levels = c(
      "ChIP-Seq + HT-SELEX",
      "ChIP-Seq",
      "ChIP-Seq + HT-SELEX + Methyl-HT-SELEX",
      "HT-SELEX",
      "SELEX",
      "HT-SELEX + Methyl-HT-SELEX",
      "SMILE-seq")), #Reorder the names
    Data_type_source = fct_recode(Data_type_source, 
                                  "ChIP-Seq \n+ HT-SELEX" = "ChIP-Seq + HT-SELEX",
                                  "HT-SELEX \n+ Methyl-HT-\nSELEX" = "HT-SELEX + Methyl-HT-SELEX",
                                  "ChIP-Seq \n+ HT-SELEX \n+ Methyl-HT-SELEX" = "ChIP-Seq + HT-SELEX + Methyl-HT-SELEX"), #Reorder the names --> cause them to be split on newline
    Motif_type = factor(Motif_type, levels = c("Half-PPARA", 
                                               "Half-PPARD", 
                                               "Half-PPARG", 
                                               "Half-RXRA", 
                                               "Half-RXRB", 
                                               "Half-RXRG", 
                                               "Full-PPARA::RXRA", 
                                               "Full-PPARG::RXRA")),  #Reordering
    Motif_type = gsub("^(Half|Full)-", "\\1-\n", Motif_type) #Adding a newline
  ) %>%
   mutate(Synthetic_or_Not = case_when(
    grepl("SMILE-seq", Data_type_source, ignore.case = TRUE) ~ "Synthetic",
    grepl("SELEX", Data_type_source, ignore.case = TRUE) & !grepl("ChIP", Data_type_source, ignore.case = TRUE) ~ "Synthetic",
    grepl("ChIP-Seq", Data_type_source, ignore.case = TRUE) & grepl("SELEX", Data_type_source, ignore.case = TRUE) ~ "Mixed",
    grepl("ChIP-Seq", Data_type_source, ignore.case = TRUE) ~ "Non-synthetic",
    TRUE ~ "Unknown"
  )) -> scatter_df
```

Asses nomality Since sample size is 21, we need to check nomality of the continous variables. We do this by the shapiro-wilks test and qq-plot

```{r statistic-nomality}
print("The shapiro-wilk test result of width is")
shapiro.test(scatter_df$Mean_width)
print("So the data is normally distributed")

print("The shapiro-wilk test result of n_amount of peaks is")
shapiro.test(scatter_df$N_amount)
print("So the data is normally distributed")

check_normality <- function(df, group_var) {
  df %>%
    group_by({{ group_var }}) %>%
    summarise(
      n = n(),
      shapiro_p_width  = if (n() >= 3) shapiro.test(Mean_width)$p.value else NA_real_,
      shapiro_p_amount = if (n() >= 3) shapiro.test(N_amount)$p.value else NA_real_,
      .groups = "drop"
    )
}

# Run for different grouping variables
check_normality(scatter_df, Motif_type_simple)
check_normality(scatter_df, Motif_db)
check_normality(scatter_df, Data_type_source)
check_normality(scatter_df, Motif_type)

ggplot(scatter_df, aes(sample = Mean_width)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  labs(title = "Q-Q plots of Mean_width")

# Q-Q plots for N_amount per group
ggplot(scatter_df, aes(sample = N_amount)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  labs(title = "Q-Q plots of N_amount")
```

# Make statitstic boxplots

```{r statistic-boxplots}
ggplot(scatter_df, aes(x = Mean_width, y = N_amount)) +
  geom_point(size = 3)  + theme_bw() + 
  geom_text_repel(aes(label = Motif_type), size = 3) +
  geom_smooth(method = "lm", se = FALSE) +  # regression line
  stat_cor( method = "pearson",
            label.x.npc = "left", label.y.npc = "top",
            label.sep = ", ", label.y = max(scatter_df$N_amount, na.rm = TRUE))+
  theme( panel.border = element_blank(), panel.grid.major = element_blank(),
         panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"),legend.title = element_blank()
  ) +
  labs(x = "Motif width (bp)", y = "Number of FIMO peaks (genome wide)") -> scatter_df_wdth_amount
scatter_df_wdth_amount

ggboxplot(scatter_df, 
          x = "Motif_db", 
          y = "N_amount", 
          add = "jitter",
          width = 0.5) +
  stat_compare_means(method = "t.test", label.y = max(scatter_df$N_amount, na.rm = TRUE)) +
  labs(x = "Motif database", y = "Number of FIMO peaks") +
  theme( panel.border = element_blank(), panel.grid.major = element_blank(),
         panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"),legend.title = element_blank()
  ) +
  theme(legend.position = "none") -> boxplt_db_number
boxplt_db_number

ggboxplot(scatter_df, 
          x = "Motif_db", 
          y = "Mean_width", 
          add = "jitter",
          width = 0.5) +
  stat_compare_means(method = "t.test", label.y = max(scatter_df$Mean_width, na.rm = TRUE)) +
  labs(x = "Motif database", y = "Motif width (bp)") +
  theme( panel.border = element_blank(), panel.grid.major = element_blank(),
         panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"),legend.title = element_blank()
  ) +
  theme(legend.position = "none") -> boxplt_db_wdth
boxplt_db_wdth

ggboxplot(scatter_df, 
          x = "Synthetic_or_Not", 
          y = "N_amount", 
          add = "jitter",
          width = 0.5) +
  stat_compare_means(method = "kruskal.test", label.y = max(scatter_df$N_amount, na.rm = TRUE)) +
  labs(x = "Synthetic", y = "Number of FIMO peaks") +
  theme( panel.border = element_blank(), panel.grid.major = element_blank(),
         panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"),legend.title = element_blank()
  ) +
  theme(legend.position = "none") -> boxplt_synthetic_number
boxplt_synthetic_number

ggboxplot(scatter_df, 
          x = "Synthetic_or_Not", 
          y = "Mean_width", 
          add = "jitter",
          width = 0.5) +
  stat_compare_means(method = "kruskal.test", label.y = max(scatter_df$Mean_width, na.rm = TRUE)) +
  labs(x = "Synthetic", y = "Motif width (bp)") +
  theme( panel.border = element_blank(), panel.grid.major = element_blank(),
         panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"),legend.title = element_blank()
  ) +
  theme(legend.position = "none") -> boxplt_synthetic_wdth
boxplt_synthetic_wdth

ggboxplot(scatter_df, 
          x = "Motif_type_simple", 
          y = "N_amount", 
          add = "jitter",
          width = 0.5) +
  stat_compare_means(method = "kruskal.test", label.y = max(scatter_df$N_amount, na.rm = TRUE)) +
  labs(x = "Motif type simplified ", y = "Number of FIMO peaks") +
  theme(legend.position = "none") +
  theme( panel.border = element_blank(), panel.grid.major = element_blank(),
         panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"),legend.title = element_blank()
  ) +
  theme(legend.position = "none") -> boxplt_mottyp_simple_number
boxplt_mottyp_simple_number

ggboxplot(scatter_df, 
          x = "Motif_type_simple", 
          y = "Mean_width", 
          add = "jitter", 
          width = 0.5) +
  stat_compare_means(method = "kruskal.test", label.y = max(scatter_df$Mean_width, na.rm = TRUE)) +
  labs(x = "Motif type simplified", y = "Motif width (bp)") +
  theme(legend.position = "none") +
  theme( panel.border = element_blank(), panel.grid.major = element_blank(),
         panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"),legend.title = element_blank()
  ) + 
  theme(legend.position = "none") -> boxplt_mottyp_simple_wdth
boxplt_mottyp_simple_wdth

ggboxplot(scatter_df, 
          x = "Data_type_source", 
          y = "N_amount", 
          add = "jitter", 
          width = 0.5) +
  stat_compare_means(method = "kruskal.test", label.y = max(scatter_df$N_amount, na.rm = TRUE)) +
  labs(x = "Data type", y = "Number of FIMO peaks") +
  theme(legend.position = "none") +
  theme( panel.border = element_blank(), panel.grid.major = element_blank(),
         panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"),legend.title = element_blank()
  ) + 
  theme(legend.position = "none") -> boxplt_dattyp_number
boxplt_dattyp_number

ggboxplot(scatter_df, 
          x = "Data_type_source", 
          y = "Mean_width", 
          add = "jitter", 
          width = 0.5) +
  stat_compare_means(method = "kruskal.test", label.y = max(scatter_df$Mean_width, na.rm = TRUE)) +
  labs(x = "Data type", y = "Motif width (bp)") +
  theme(legend.position = "none") +
  theme( panel.border = element_blank(), panel.grid.major = element_blank(),
         panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"),legend.title = element_blank()
  ) + 
  theme(legend.position = "none") -> boxplt_dattyp_wdth
boxplt_dattyp_wdth

ggboxplot(scatter_df, 
          x = "Motif_type", 
          y = "N_amount", 
          add = "jitter", 
          width = 0.5) +
  stat_compare_means(method = "kruskal.test", label.y = max(scatter_df$N_amount, na.rm = TRUE)) +
  labs(x = "Motif type", y = "Number of FIMO peaks") +
  theme( panel.border = element_blank(), panel.grid.major = element_blank(),
         panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"),legend.title = element_blank()
  ) +
  theme(legend.position = "none") -> boxplt_mottyp_number
boxplt_mottyp_number

ggboxplot(scatter_df, 
          x = "Motif_type", 
          y = "Mean_width", 
          add = "jitter", 
          width = 0.5/length(unique(scatter_df$Motif_type))*length(unique(scatter_df$Motif_type))) +
  stat_compare_means(method = "kruskal.test", label.y = max(scatter_df$Mean_width, na.rm = TRUE)) +
  labs(x = "Motif type", y = "Motif width (bp)") +
  theme( panel.border = element_blank(), panel.grid.major = element_blank(),
         panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"),legend.title = element_blank()
  ) + 
  theme(legend.position = "none") -> boxplt_mottyp_wdth
boxplt_mottyp_wdth
```

# Saveblock boxplots

```{r boxplot save, include=FALSE}
combined_plot_r1 <- 
  boxplt_db_number + boxplt_db_wdth + 
  boxplt_mottyp_simple_number + boxplt_mottyp_simple_wdth + 
  plot_layout(ncol = 4)

ggsave("data/PPAR_RXR_motifs_Hocomoco_Jaspar/Objective1_results/Statistics_plots/even_boxplot_panel.pdf", combined_plot_r1, width = 16, height = 4)


combined_plot_r2 <- 
  boxplt_dattyp_number + boxplt_dattyp_wdth +
  plot_layout(ncol = 2)

ggsave("data/PPAR_RXR_motifs_Hocomoco_Jaspar/Objective1_results/Statistics_plots/even_boxplot_panel_2.pdf", combined_plot_r2, width = 16, height = 4)


combined_plot_r3 <- 
  boxplt_mottyp_number + boxplt_mottyp_wdth +
  plot_layout(ncol = 2)

ggsave("data/PPAR_RXR_motifs_Hocomoco_Jaspar/Objective1_results/Statistics_plots/even_boxplot_panel_3.pdf", combined_plot_r3, width = 16, height = 4)



pdf("data/PPAR_RXR_motifs_Hocomoco_Jaspar/Objective1_results/Statistics_plots/FIMO_scat_plts.pdf", width = 8, height = 3, useDingbats = FALSE)  
print(scatter_df_wdth_amount)
dev.off()


pdf("data/PPAR_RXR_motifs_Hocomoco_Jaspar/Objective1_results/Statistics_plots/FIMO_db_plts.pdf", width = 1.5 * length(unique(scatter_df$Motif_db)), height = 3)
print(boxplt_db_number)
print(boxplt_db_wdth)
dev.off()

pdf("data/PPAR_RXR_motifs_Hocomoco_Jaspar/Objective1_results/Statistics_plots/FIMO_mottyp_smp_plts.pdf", width = 1.5* length(unique(scatter_df$Motif_type_simple)), height = 3)
print(boxplt_mottyp_simple_number)
print(boxplt_mottyp_simple_wdth)
dev.off()

pdf("data/PPAR_RXR_motifs_Hocomoco_Jaspar/Objective1_results/Statistics_plots/FIMO_dattyp_plts.pdf", width = 1 * length(unique(scatter_df$Data_type_source)), height = 3)
print(boxplt_dattyp_number)
print(boxplt_dattyp_wdth)
dev.off()

pdf("data/PPAR_RXR_motifs_Hocomoco_Jaspar/Objective1_results/Statistics_plots/FIMO_mottyp_plts.pdf", width = 1 * length(unique(scatter_df$Motif_type)), height = 3)
print(boxplt_mottyp_number)
print(boxplt_mottyp_wdth)
dev.off()
```

# A BEDFILE is created from all the ChipSeekr genes and the PPRE score is added

```{r final genelist nonzero}
chip_scored <- chip_all %>%
  dplyr::mutate(ppar_score = score * exp(-(abs(distanceToTSS)) / 3000)) #thousand is a constant, which is set to 2000 since PPAR is promoter dominant. Also used in cistrome go


GTEX_snRNA_cardiomyocyte_nonzero <- read.csv("data/GTEX_single_nuclear_data/GTEx_non_zeros_ensembl.csv")

GTEX_snRNA_cardiomyocyte_nonzero %>%
  left_join(chip_scored, by = c( "ensembl_gene_id" ="ENSEMBL"), keep = TRUE) %>%
  drop_na(c(ensembl_gene_id,ENSEMBL)) -> motif_hits_cardiomyocytes_nonzero_scored

motif_presence <- motif_hits_cardiomyocytes_nonzero_scored %>%
  distinct(ENSEMBL, motif_id) %>%          # 1 rij per (gene, motif) die voorkomt
  mutate(hit = 1L) %>%                     # als het voorkomt is het een hit
  pivot_wider(
    names_from  = motif_id,
    values_from = hit,
    values_fill = list(hit = 0L),
    names_prefix = "motif_"
  )

ranked_scores_nonzero <- motif_hits_cardiomyocytes_nonzero_scored %>%
  group_by(ENSEMBL) %>%
  dplyr::summarise(PPRE_score_sum = sum(ppar_score, na.rm = TRUE),
            PPRE_score_max = max(ppar_score, na.rm = TRUE),
            PPRE_score_median = median(ppar_score, na.rm = TRUE),
            PPRE_score_average = mean(ppar_score, na.rm = TRUE),
            chromStart = dplyr::first(as.integer(start)),
            chromEnd   = dplyr::first(end),
            Chr       = dplyr::first(seqnames),
            SYMBOL    = dplyr::first(SYMBOL),
            ensembl   = dplyr::first(ENSEMBL),
            score     = dplyr::first(PPRE_score_median),
            strand    = dplyr::first(strand),
            geneStart = dplyr::first(geneStart),
            geneEnd   = dplyr::first(geneEnd), 
            name      = dplyr::first(motif_id),
            distanceToTss = dplyr::first(distanceToTSS),
            expression= dplyr::first(expression_value)) %>%
  dplyr::mutate(across(.cols = starts_with("PPRE_score_"),
    .fns = ~ rank(-., ties.method = "first"),  # minus sign for descending ranking
    .names = "rank_{.col}")) #the name for the col is rank_ + colname 

ranked_scores_nonzero_with_motifs <- ranked_scores_nonzero %>%
  left_join(motif_presence, by = "ENSEMBL")

top2000_genes_nonzero <- ranked_scores_nonzero %>%
  arrange(desc(PPRE_score_average)) %>%
  slice_head(n = 2000)
```

```{r}


chip_scored_single_gens <- chip_scored %>%
  group_by(ENSEMBL) %>%
  dplyr::summarise(PPRE_score_sum = sum(ppar_score, na.rm = TRUE),
            PPRE_score_max = max(ppar_score, na.rm = TRUE),
            PPRE_score_median = median(ppar_score, na.rm = TRUE),
            PPRE_score_average = mean(ppar_score, na.rm = TRUE),
            chromStart = dplyr::first(as.integer(start)),
            chromEnd   = dplyr::first(as.integer(end)),
            Chr       = dplyr::first(seqnames),
            SYMBOL    = dplyr::first(SYMBOL),
            ensembl   = dplyr::first(ENSEMBL),
            score     = dplyr::first(PPRE_score_median),
            strand    = dplyr::first(strand),
            geneStart = dplyr::first(geneStart),
            geneEnd   = dplyr::first(geneEnd), 
            name      = dplyr::first(motif_id),
            distanceToTSS = dplyr::first(distanceToTSS)) %>%
  dplyr::mutate(across(.cols = starts_with("PPRE_score_"),
    .fns = ~ rank(-., ties.method = "first"),  # minus sign for descending ranking
    .names = "rank_{.col}")) #the name for the col is rank_ + colname 


chip_scored_single_gens_bed <- data.frame(
  chrom      = chip_scored_single_gens$Chr,
  chromStart = chip_scored_single_gens$chromStart - 1,   # BED is 0-based
  chromEnd   = chip_scored_single_gens$chromEnd,
  motifName       = if ("name"  %in% names(chip_scored_single_gens)) chip_scored_single_gens$name  else ".",
  score      = if ("score" %in% names(chip_scored_single_gens)) chip_scored_single_gens$PPRE_score_median else NA_character_,
  strand     = as.character(chip_scored_single_gens$strand),
  ensembl    = if ("ensembl" %in% names(chip_scored_single_gens))
                 chip_scored_single_gens$ensembl else NA_character_,
  symbol     = if ("SYMBOL" %in% names(chip_scored_single_gens)) {
                 chip_scored_single_gens$SYMBOL
               } else NA_character_,
  geneStart  = if ("geneStart" %in% names(chip_scored_single_gens))
                 chip_scored_single_gens$geneStart else NA_integer_,
  geneEnd    = if ("geneEnd" %in% names(chip_scored_single_gens))
                 chip_scored_single_gens$geneEnd else NA_integer_,
  distanceToTSS = chip_scored_single_gens$distanceToTSS,
  PPRE_score_sum = chip_scored_single_gens$PPRE_score_sum,
  PPRE_score_max = chip_scored_single_gens$PPRE_score_max,
  PPRE_score_average = chip_scored_single_gens$PPRE_score_average,
  ensembl_genes = chip_scored_single_gens$ensembl,
  check.names = FALSE
)


write.table(chip_scored_single_gens_bed, "data/PPAR_RXR_motifs_Hocomoco_Jaspar/Objective1_results/Bedfiles_output/PPRE_CHIP_all_genes_test.bed",
            sep = "\t", quote = FALSE,
            row.names = FALSE, col.names = TRUE)
```
```{r}
BiocManager::install("Gviz")
library(Gviz)
library(GenomicRanges)
fam <- motif_hits_cardiomyocytes_nonzero_scored %>%
  filter(SYMBOL == "FAM53A") %>%
  filter(grepl("^Promoter", annotation)) %>%   # of laat dit weg als je alles wil
  transmute(
    chrom = as.character(seqnames),
    chromStart = as.integer(start) - 1L,       # BED 0-based
    chromEnd   = as.integer(end),
    name  = motif_id,                          # label in IGV
    score = as.integer(round(score)),          # IGV gebruikt dit soms voor shading
    strand = as.character(strand)
  )

write.table(
  fam,
  file = "data/PPAR_RXR_motifs_Hocomoco_Jaspar/Objective1_results/Bedfiles_output/FAM53A_promoter_motif_hits.bed",
  sep = "\t",
  quote = FALSE,
  row.names = FALSE,
  col.names = FALSE
)
```

```{r}
nanog <- motif_hits_cardiomyocytes_nonzero_scored %>%
  filter(SYMBOL == "NANOG") %>%  # of laat dit weg als je alles wil
  transmute(
    chrom = as.character(seqnames),
    chromStart = as.integer(start) - 1L,       # BED 0-based
    chromEnd   = as.integer(end),
    geneStart = as.integer(geneStart),
    geneEnd = as.integer(geneEnd),
    distanceToTSS = as.integer(distanceToTSS),
    name  = motif_id,                          # label in IGV
    score = as.integer(round(score)),          # IGV gebruikt dit soms voor shading
    strand = as.character(strand)
  )

write.table(
  nanog,
  file = "data/PPAR_RXR_motifs_Hocomoco_Jaspar/Objective1_results/Bedfiles_output/nanog_promoter_motif_hits.bed",
  sep = "\t",
  quote = FALSE,
  row.names = FALSE,
  col.names = FALSE
)
```

```{r create-BEDfile-PPRE_cardiomyocyte_expressed}
ranked_scores_nonzero <- ranked_scores_nonzero_with_motifs
ranked_scores_nonzero_bed <- data.frame(
  chrom      = ranked_scores_nonzero$Chr,
  chromStart = ranked_scores_nonzero$chromStart - 1,   # BED is 0-based
  chromEnd   = ranked_scores_nonzero$chromEnd,
  motifName       = if ("name"  %in% names(ranked_scores_nonzero)) ranked_scores_nonzero$name  else ".",
  score      = if ("score" %in% names(ranked_scores_nonzero)) ranked_scores_nonzero$PPRE_score_median else 0,
  strand     = as.character(ranked_scores_nonzero$strand),
  ensembl    = if ("ensembl" %in% names(ranked_scores_nonzero))
                 ranked_scores_nonzero$ensembl else NA_character_,
  symbol     = if ("SYMBOL" %in% names(ranked_scores_nonzero)) {
                 ranked_scores_nonzero$SYMBOL
               } else NA_character_,
  geneStart  = if ("geneStart" %in% names(ranked_scores_nonzero))
                 ranked_scores_nonzero$geneStart else NA_integer_,
  geneEnd    = if ("geneEnd" %in% names(ranked_scores_nonzero))
                 ranked_scores_nonzero$geneEnd else NA_integer_,
  distancetoTSS = ranked_scores_nonzero$distanceToTss,#
  PPRE_score_sum = ranked_scores_nonzero$PPRE_score_sum,
  PPRE_score_max = ranked_scores_nonzero$PPRE_score_max,
  PPRE_score_average = ranked_scores_nonzero$PPRE_score_average,
  GTEx_expression = ranked_scores_nonzero$expression,
  ensembl_genes = ranked_scores_nonzero$ensembl,
  check.names = FALSE
)

motif_cols <- grep("^motif_", names(ranked_scores_nonzero_with_motifs), value = TRUE)

ranked_scores_nonzero_bed <- cbind(
  ranked_scores_nonzero_bed,
  ranked_scores_nonzero[, motif_cols, drop = FALSE]
)

write.table(ranked_scores_nonzero_bed, "data/PPAR_RXR_motifs_Hocomoco_Jaspar/Objective1_results/Bedfiles_output/PPRE_gtex_ranked_scores_abs_3k_test.bed",
            sep = "\t", quote = FALSE,
            row.names = FALSE, col.names = TRUE)
```

```{r final genelist top20}
GTEX_snRNA_cardiomyocyte_top20 <- read.csv("data/GTEX_single_nuclear_data/GTEx_top20%_ensembl.csv" )

GTEX_snRNA_cardiomyocyte_top20 %>%
  left_join(chip_scored, by = c( "ensembl_gene_id" ="ENSEMBL"), keep = TRUE) %>%
  drop_na(c(ensembl_gene_id,ENSEMBL)) -> motif_hits_cardiomyocytes_top20_scored

ranked_scores_top20 <- motif_hits_cardiomyocytes_top20_scored %>%
  group_by(ENSEMBL) %>%
  dplyr::summarise(PPRE_score_sum = sum(ppar_score, na.rm = TRUE),
            PPRE_score_max = max(ppar_score, na.rm = TRUE),
            PPRE_score_median = median(ppar_score, na.rm = TRUE),
            PPRE_score_average = mean(ppar_score, na.rm = TRUE), .groups = "drop") %>%
  dplyr::mutate(across(.cols = starts_with("PPRE_score_"),
    .fns = ~ rank(-., ties.method = "first"),  # minus sign for descending ranking
    .names = "rank_{.col}")) %>% #the name for the col is rank_ + colname
  left_join(GTEX_snRNA_cardiomyocyte_nonzero, by = c("ENSEMBL" = "ensembl_gene_id")) %>%
  dplyr::select(!X)

top2000_genes_top20 <- ranked_scores_top20 %>%
  arrange(desc(PPRE_score_average)) %>%
  slice_head(n = 2000)
```



# Distributions of topX genes 

with different aggregation methods per gene (Sum, Max, Median, Average)

```{r Distribution-of-counts-SUM_MAX_MEDIAN_AVERAGE}
scoring_list <- list(
  nonzero = ranked_scores_nonzero
)

genes_to_annotate <- c("CPT1A","HADHA","HADHB","ACADVL","PPARA", "PPARB", "PPARG", "RXRA", "RXRB", "RXRG" )
plot_list <- list()

for (i in seq_along(scoring_list)) {
  df <- scoring_list[[i]] 
  ds_name <- names(scoring_list)[i]

  gene_candidates <- c("SYMBOL", "HGNC_gene_name", "hgnc_symbol", "gene_symbol")
  gene_col <- intersect(gene_candidates, names(df))[1]
  if (is.na(gene_col))  
    next
  df <- dplyr::rename(df, SYMBOL = dplyr::all_of(gene_col))
  for (col_name in names(df)) {
    
    if (!startsWith(tolower(col_name), "ppre"))
      next

    rank_col <- paste0("rank_", col_name)
    if (!rank_col %in% names(df))
      next

    if (!is.numeric(df[[col_name]])) 
      next

    # values for annotated genes
    gene_values <- df %>%
      dplyr::filter(SYMBOL %in% genes_to_annotate) %>%
      dplyr::select(SYMBOL, value = dplyr::all_of(col_name))

    p <- ggplot2::ggplot(df, ggplot2::aes(x = .data[[col_name]])) +
      ggplot2::geom_histogram(ggplot2::aes(fill = .data[[rank_col]]), bins = 300) +
      ggplot2::scale_fill_manual(values = c("TRUE" = "black", "FALSE" = "red")) +
      ggplot2::labs(title = gsub("_", " ", col_name),
                    x = gsub("_", " ", col_name),
                    y = "Frequency") +
      ggplot2::scale_x_continuous(limits = c(0, NA), expand = c(0, 0)) +
      ggplot2::scale_y_continuous(limits = c(0, NA), expand = c(0, 0)) +
      ggplot2::coord_cartesian(clip = "off") +
      ggplot2::guides(fill = "none") +
      ggplot2::theme_bw() +
      ggplot2::theme(
        panel.border = ggplot2::element_blank(),
        panel.grid.major = ggplot2::element_blank(),
        panel.grid.minor = ggplot2::element_blank(),
        axis.line.x = ggplot2::element_line(color = "black"),
        axis.ticks.x = ggplot2::element_line(color = "black"),
        axis.line.y = ggplot2::element_line(color = "black"),
        legend.title = ggplot2::element_blank()
      )

    # add vertical lines & labels for annotated genes
    if (nrow(gene_values) > 0) {
      p <- p +
        ggplot2::geom_vline(data = gene_values,
                            ggplot2::aes(xintercept = value),
                            color = "blue", linetype = "dashed") +
        ggplot2::geom_text(data = gene_values,
                           ggplot2::aes(x = value, y = Inf, label = SYMBOL),
                           inherit.aes = FALSE, angle = 90, vjust = 1.2, size = 3, color = "blue")
    }

    # store plot with dataset + column name
    plot_list[[paste(ds_name, col_name, sep = "_")]] <- p
  }
}

print(plot_list[["nonzero_PPRE_score_sum"]])
print(plot_list[["nonzero_PPRE_score_max"]])
print(plot_list[["nonzero_PPRE_score_median"]])
print(plot_list[["nonzero_PPRE_score_average"]])
```

```{r}
pdf("data/PPAR_RXR_motifs_Hocomoco_Jaspar/Objective1_results/Scoring_FIMO/PPRE_score_histograms.pdf", width = 11, height = 6)  # adjust size if you want
for (nm in names(plot_list)) {
  print(plot_list[[nm]])
}
dev.off()
```
# Over Representation Analysis

Dotplots per different aggregation method

```{r run-ORA-dtplt}
run_ora_dotplots <- function(df, pdf_path, witdh, height){
  # open a multi-page PDF device
  dir.create(dirname(pdf_path), showWarnings = FALSE, recursive = TRUE)
  grDevices::pdf(pdf_path, onefile = TRUE, width = witdh, height = height, useDingbats = FALSE) #8.9 for article
  on.exit(try(grDevices::dev.off(), silent = TRUE), add = TRUE)

  cols <- names(df)[startsWith(tolower(names(df)), "ppre")]
  cat("Processing", length(cols), "columns:", paste(cols, collapse = ", "), "\n")
  printed <- 0

  for (col_name in cols) {
    cat("\n--", col_name, "--\n")
    if (!("SYMBOL" %in% names(df))) { cat("No SYMBOL column.\n"); next }

    top_syms <- df %>%
      arrange(desc(.data[[col_name]])) %>%
      slice_head(n = 5000) %>%
      pull(SYMBOL) %>%
      unique() %>% na.omit()

    cat("Top symbols:", length(top_syms), "\n")
    if (length(top_syms) < 10) { cat("Too few symbols.\n"); next }

    gene_map <- bitr(top_syms, fromType="SYMBOL", toType="ENTREZID", OrgDb=org.Hs.eg.db)
    cat("Mapped to ENTREZ:", nrow(gene_map), "\n")
    if (nrow(gene_map) < 10) { cat("Too few mapped; skipping.\n"); next }

    ego <- suppressMessages(enrichGO(
      gene=gene_map$ENTREZID, OrgDb=org.Hs.eg.db, ont="BP",
      pAdjustMethod="BH", pvalueCutoff=0.05, qvalueCutoff=0.05,
      minGSSize=10, maxGSSize=5000, readable=TRUE))

    n_terms <- nrow(as.data.frame(ego))
    cat("Significant terms:", n_terms, "\n")
    if (n_terms == 0) { cat("No significant enrichment under thresholds.\n"); next }

    p <- dotplot(ego, showCategory=20, x="GeneRatio", orderBy="x",
                 title=gsub("_", " ", col_name))
    print(p)  # each print adds a new page to the PDF
    printed <- printed + 1
  }

  if (printed == 0) {
    grid::grid.newpage()
    grid::grid.text("No ORA plots produced.\nCheck mapping/thresholds/top_n.", x=.5, y=.6)
  }
  cat("\nDone. Printed", printed, "plots to", pdf_path, "\n")
}

outdir <- "data/PPAR_RXR_motifs_Hocomoco_Jaspar/Objective1_results"
run_ora_dotplots(ranked_scores_top20,  file.path(outdir, "ORA_top20.pdf"), 8, 9)
run_ora_dotplots(ranked_scores_nonzero,file.path(outdir, "ORA_nonzero.pdf"), 8,9)
run_ora_dotplots(ranked_scores_nonzero,file.path(outdir, "ORA_nonzero_graphical abstract.pdf"), 5,5)
```
```{r}
library(plotly)
funnel_df <- data.frame(
  x = c(0, 1, 2, 3),
  y = c(nrow(PPRE_FIMO_bed), nrow(overlapping_genes), nrow(GTEX_genes), nrow(ranked_scores_nonzero)),
  name = as.factor(c("Amount of FIMO peaks", "All PPRE genes", "All GTEx genes", "GTEx cardiomyocyte expressed PPREs")))

funnel_df$name <- factor(funnel_df$name, levels = as.character(funnel_df$name))

fig <- plot_ly(
  data = funnel_df,
  type = "funnel",
  y = ~name,     # step labels
  x = ~y,        # values
  text = ~y,
  textinfo = "value",
  textposition = "inside"
)

fig <- fig %>%
  layout(
    xaxis = list(title = "Count"),
    yaxis = list(title = "")
  )

fig
```

